### 2015-02-16 manuscripts/biophysical-journal-cover-letter/response-letter.tex
Initial import from http://simtk.org/home/bhmm Adding everything but experimental datasets.

\documentclass[ucb,qb3,10pt,fullfrom]{ucletter}

\usepackage{palatino}

\name{John D. Chodera}
\telephone{415.867.7384}
\email{jchodera@berkeley.edu}
%\department{Department of Chemistry}
%\fromaddress{QB3 Institute}
%260J Stanley Hall #3220\\
%University of California\\
%Berkeley, CA 94720-3220}
%\mailcode{5447}
\date{\today}

%\documentclass[10pt]{letter}
%\address{
%John D. Chodera\\
%QB3 Institute\\
%260J Stanley Hall \#3220\\
%University of California\\
%Berkeley, CA 94720-3220, USA \\
%phone: 415.867.7384\\
%email: \href{mailto:jchodera@berkeley.edu}{jchodera@berkeley.edu}\\
%\name{John D. Chodera}
%}

\usepackage{fancyhdr}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}

\def\bs{$\backslash$}

\pagestyle{fancy}
\fancyhf{}
\rfoot{\small\textsc{John D. Chodera - Cover Letter - \thepage}}

\signature{
\resizebox{2.5in}{!}{\includegraphics{JDC-signature}} \\
John D.~Chodera,\\corresponding author
}

\usepackage{times}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}

% We will be using PDF figures and generating PDF output.
\pdfoutput=1
\pdfcompresslevel=9
\usepackage[pdftex]{graphicx}
\DeclareGraphicsExtensions{.pdf}

%        \addtolength{\oddsidemargin}{-.875in}
%        \addtolength{\evensidemargin}{-.875in}
%        \addtolength{\textwidth}{1.75in}
%
%        \addtolength{\topmargin}{-.875in}
%        \addtolength{\textheight}{1.75in}
%
%\begin{document}
%
%\begin{letter}{To:\\MFPL Faculty Search Committee}
%
%\opening{Dear Search Committee,}

\begin{document}
%\small
\begin{letter}{}
\opening{Dear \emph{Biophysical Journal} Editors,}

We are submitting a revised manuscript for consideration for publication as an {\bf Article} in \emph{Biophysical Journal}:

{\bf Bayesian hidden Markov model analysis of single-molecule force spectroscopy: Characterizing kinetics under measurement uncertainty}, by John D.~Chodera, Phillip J.~Elms, Frank No\'{e}, Bettina Keller, Christian M.~Kaiser, Aaron Ewall-Wice,  Susan Marqusee, Carlos Bustamante, and Nina Singhal Hinrichs.

We apologize for the long period of time that elapsed between the receipt of referee comments and the submission of this revision.
The first author, John D.~Chodera, was engaged in a protracted faculty search that did not permit proper attention to be devoted to thoroughly addressing the referee comments until now.
We request that the referee critiques (included in this letter) and our responses be considered in the evaluation of this manuscript revision, especially considering the time devoted by the referees to providing thoughtful and useful commentary, for which we are highly appreciative.
We were also happy to see the enthusiasm the referees share for this work.

Below, we address the referee comments (set in red) with detailed responses (in black), explaining where and how we have adapted the manuscript to meet their concerns.
We believe the revised manuscript is much stronger and more broadly accessible, and hope the Editors will find this version suitable for publication in \emph{Biophysical Journal}.

\color{red}
{\bf Peer Reviewer \#1}

This paper applies hidden Markov models to extract thermodynamic and kinetic information from single molecule hopping traces. The authors explain different methods to extract transition rates between an arbitrary number states with Gaussian distributed observables from hopping data.

The problem is certainly very interesting but the reading of the manuscript is difficult and the presentation of the methodology obscure to non-experts. I think this paper should be primarily written in a way that is understandable to an audience of experimentalists working in single molecule biophysics and readers of Biophys. J in general. The core of this manuscript is too specialised (pages 11-18 describe all the steps of the algorithms and are difficult to follow) failing to deliver the main message.
\color{black}

We have reorganized the manuscript to streamline delivery of the main message and make it more accessible to a broad audience, including experimentalists.
While we feel strongly that it is essential to present a \emph{complete} description of the algorithms so that other researchers can independently implement them and reproduce our work as closely as possible, we have tried to minimize the impact detailed descriptions have on the main flow for the general reader, so we have relocated such verbose descriptions in the Appendix.

\color{red}
Moreover the experimental part is weak with just one trace of one molecular species extracted from a pool of RNAs that supposedly should only contain the wild type P5ab hairpin. Although I can understand this, it remains unclear whether results are reproducible and the conclusions drawn are tenous: the sequence and the structure of the molecule are unknown and there is no evidence whether the distinct three-state chemical species will appear in another synthesis of P5ab. This paper should start first describing the experimental results and then the theory behind the results in a away that is accessible to experimentalists. Otherwise this paper is suitable for publication in a more specialised journal on optimization algorithms and statistical inference methods.
\color{black}

The p5ab RNA used in this study comes from the same sample used in previous reports~\cite{liphardt:science:2001:p5ab,wen-manosas:biophys-j:2007:rna-optical-tweezers,wen-manosas:biophys-j:2007:rna-optical-tweezers-2,elms:biophys-j:2012:force-feedback}.
While we cannot rule out the presence of a diversity of molecules resulting from synthesis or storage, we note that the resolution of the instrument used in the current study is superior to that used in previous reports on this RNA system, allowing for the possibility that previous measurements missed the three-state-like behavior we observe in some traces. 
{\color{blue}[JDC: We can show the analysis of two such fibers in the table to demonstrate the results agree within error bars.]}

We also note that the main purpose of our inclusion of experimental data was not to draw new conclusions about the behavior of this well-studied RNA hairpin or its potential degradation products or synthesis byproducts, but to illustrate the performance of our Bayesian extension of HMMs on real data containing all of the features of real measurements.
In this revision, we therefore include analysis of traces of another system, apomyoglobin at pH 5, a system studied previously~\cite{elms:pnas:2012:apo-myoglobin} on the same instrument. 
Rate estimation on this system system is especially challenging due to the highly overlapping nature of the force histograms.
{[\color{blue} TODO: include results on apomyoglobin.]}

As suggested, we move up discussion of the experimental measurements earlier in the manuscript to make it more accessible to experimentalists, and add a paragraph summarizing our primary results at the end of the Introduction.

\color{red}
Other comments

Major concerns:

1) What is the potential influence of drift in the method?. In the presence of slight drift the output signal corresponding to different states gets mixed and this may compromise the successfulness of the method. Ideally one could always look for drift-less traces but this may be very difficult in some cases or some experimental setups. The authors should discuss this problem.
\color{black}

Compensating for significant drift solely in analysis is difficult, as changes in the relative optical trap position will change the external potential felt by the molecule, and hence alter the kinetics and thermodynamics.
Without imposing a model of how the rate constants and equilibrium state distributions change with the applied trap potential, as well as a physical or empirical model for the dynamics of instrumental drift, it is impossible to compensate for these effects during analysis.

Fortunately, obtaining many drift-free measurement traces of the duration presented here with the generation of miniaturized instrument we employed (where one bead is suctioned onto a pipette and the other bead held in a dual counter-propagating laser trap within a miniaturized, suspended frame) is not particularly challenging---see comments later on this response for further information on how we analyzed drift.
Drift is even less problematic for newer generations of instrument that eliminate the pipette altogether, replacing it with another optical trap.

We have added additional discussion to the paper to this effect. 
{\color{blue} [TODO: add discussion about drift to manuscript.]}
{\color{blue} [TODO: Make sure we cite papers on instrument discussing drift (Jeff Moffitt's PNAS paper on dual trap with analysis of S:N ratio, patent, Block papers on dual traps).]}

\color{red}
2) How do the finite decorrelation time of the bead in the trap influences the value of the kinetic rates? In the presence of a finite decorrelation time the assumption of uncorrelated (in time) output signal is not anymore valid. Measurements shown in Fig. 3 were subsampled at 1kHz because apparently 1ms is much smaller than the relaxation time of the bead in the trap (bead under tension subjected to the trap and to the tether). The relaxation time (page 2, supplementary information and figure 3 of the Supp.~Info) is about 0.5ms (2kHz). Being 1kHz and 2kHz comparable values I do not see how the motion of the bead can be taken as uncorrelated and its effect in the dynamics of the output signal neglected.
\color{black}

The intrinsic timescale or correlation time $\tau$ generally indicates that the autocorrelation function of a process has decayed exponentially to $1/e \approx 37\%$ of the original correlation, such that waiting $2\tau$ indicates this process has decayed to $1/e^2 \approx 14\%$ of the original correlation; by $3 \tau$, this is $1/e^3 \approx 5\%$.
Because this correlation decays exponentially, it is not the case that 1 kHz and 2 kHz are comparable values.
Examination of the power spectrum reveals that the corner frequency of the bead (indicating the timescale for correlation loss by diffusive motion of the bead) is $\sim$2.5 kHz, so $\sim 2.5 \tau = 1~\mathrm{ms} = (1~\mathrm{kHz})^{-1}$ results in a decay of this correlation from $100\%$ to $1/e^{2.5} \approx 8\%$, making the bead position within the state effectively uncorrelated at 1 kHz.

See also our response to \#3 below.

\color{red}
3) I am concerned about the time decay of the correlation function shown in figure 3 (supplementary). As said in page 2 (supplementary): "An initial non-exponential or multi-exponential (with unclear number of components) decay is seen in the first 0.5ms to approximately one tenth of the initial value, with slow decay following." Why the initial decay of the bead in the trap should show anarmonicities (leading to multi-exponential behavior)? What is the source of the slow decay observed in the bottom of Fig.3(supplementary)? Surely this cannot be attributed to drift (as this is very small, see my comment 6 below). What is the cause of the apparently "non-ergodic" slow decay? This experimental features of the measurements should be explained.
\color{black}

We apologize for the error in preparing Supplementary Figure 3.
We had chosen a trace we had presumed to show single-state behavior from inspection of the histogram (due to the trap position), but further inspection of the temporal trace reveals that this trace in fact shows multi-state kinetic behavior.
The long tail of the correlation function is in fact indicative of the infrequent transitions between the kinetically metastable states in this trace, with the initial non- or multi-exponential behavior (which would not be treated by our HMM/BHMM) essentially complete by 0.3 ms, meaning the corresponding intrinsic relaxation times are shorter than this.
To clarify the point this figure was intended to communicate, rather than display the correlation function from a different trace that shows single-state behavior, we have added a panel to this figure illustrating the multistate behavior, and amended the caption to point out the short-time decay (not treated by the HMM/BHMM) and long-time decay (explicitly treated by the HMM/BHMM). 
{[\color{blue} TODO: Add additional panel to supplementary figure 3 showing hopping in kinetic trace; plot exponential relaxation fits; amend caption]}

\color{red}
4) In page 17 the authors comment on the problem of reshuffling the time data series and the invariance of the full BHMM posterior Eq.~8. How they resolve this question is a bit unclear to me and appears to be a strong limitation of maximum likelihood methods. This problem should be commented much before after Eq.~8 when these methods are first described.
\color{black}

The BHMM posterior is invariant to exchanging labels of the states, not ``reshuffling the timeseries''; we have amended the manuscript text to make this issue clearer, as well as moved up discussion of this subtle and somewhat complex issue earlier in the manuscript.
This issue does not appear to be a limitation of maximum likelihood methods, but rather simply an issue complicating analysis of MCMC sampling from the BHMM posterior.
{\color{blue} [JDC: Clarify the state exchange issue and move up discussion.]}

\color{red}
5) Page 20, "The P5ab RNA hairpin from Tetrahymena termophilia..... was provided " I would not call the molecule studied P5ab because this leads to confusion. The molecule studied in this work corresponds to a species chemically distinct from p5ab. They should use another name for that species. What is the primary sequence of this molecule? Does it have a contour length compatible with P5ab? Is this new chemical species reproducible in synthesis protocols? What are the secondary structure of this molecule? I have doubts any experimental group whatsoever---maybe neither the authors themselves---can repeat the experiments and test the same molecule again after a new synthesis (??). Why not designing and synthesising an RNA molecule with a reproducible sequence exhibiting a third state with fast kinetic rates? That should not be difficult. Please, remove the term p5ab from the text and figure captions and call it something else.
\color{black}

As the main point of our manuscript is to illustrate the performance of our analysis approach on real experimental data, we have taken two steps to address this issue:
\begin{enumerate}
\item We have limited our description of the RNA molecule to an ``RNA hairpin'' and only describe its origin in the Methods section of the text, so as not to mislead readers that the chemical composition is known to correspond to the p5ab sequence. 
\item We have added a section describing our analysis of two-state hopping behavior for another molecule, apomyoglobin at pH 5, whose study by force spectroscopy measurements was recently reported in detail elsewhere \cite{elms:pnas:2012:apo-myoglobin}. 
\end{enumerate}

\color{red}
6) page 20, "Drift in the instrument was less than 1nm/minute..." Drift is one of the main drawbacks of hopping measurements. 1nm/min looks to me really quite remarkable. One minute is a very long time for a single-trap (rather than dual trap) setup that contains one pipette glued to a glass chamber (that strongly suffers from thermal expansion effects) and a laser beams subject to piezo actuators with creep effects. How many digital-to-analog conversion units do correspond to 1nm in their instrument? I think the authors should show the experimental trace shown in figure 3 filtered at several bandwidth values (10,1,0.1Hz) to show how really big is the drift.
\color{black}

A 32-bit ADC was used where, after calibration, the measured force corresponded to 0.0016697 pN/ADC unit, which is $\sim$600 ADC units/pN.
This is now noted in the manuscript.

The average drift in the trap position was determined by analyzing 60-second traces at extreme trap positions where the system was either predominantly folded or unfolded, making few short excursions to the other state.  
The average force over one second was determined at both the beginning and the end of the trace in a region where there were no obvious transitions.  
The difference in the force was then divided by the effective spring constant of the system providing the change and therefore the drift in the trap position.  
Over four such traces, the average drift was 1.1 nm, and the standard deviation 0.7 nm, which correspond to 0.088 pN and 0.056 pN, respectively, given the trap force constant of 0.08 pN.
This information is now noted in the manuscript.

We have also added a figure showing filtering (by averaging) at the suggested frequencies to the Supplementary Material. 
{\color{blue} [TODO: Add supplementary figure.]}

\color{red}
7) page 21. The large values for k12 and k23 as compared to k13 indicate that a sequential mechanism for hopping is preferred. However the authors suggest that this may not be the only folding pathway. On what basis they can conclude this if they do not know what the molecular structures of the native nor the intermediate are? How can they exclude that indeed the smallness of k13 is just due to numerical uncertainty and that there is just one folding pathway?
\color{black}

The referee is correct that we cannot say for certain what the molecular structures associated with these force states are, so we restrict our language to use the terms ``high force,'' ``intermediate force,'' and ``low force.''
The value of $k_{13}$ is statistically bounded by the 95\% confidence interval $[7.2,11.3]$ s$^{-1}$, so it is actually quite well determined.
However, $k_{13}$ is at least an order of magnitude smaller than $k_{12}$ and $k_{23}$.
If one chooses their criteria for the existence of only one pathway based on a single pathway carrying an order of magnitude more flux than all other pathways, then there would indeed be only one pathway---but this is a matter of how one chooses the criteria for saying so.
For whichever criteria are desired, the model provides a statistical way to evaluate this criteria and express its confidence.

\color{red}
Minor concerns:

1) page 5, second line. The acronym MLHMM is not defined.

2) page 5, "...for assessing whether the cost of collecting additional data is outweighed by their benefits." What does this sentence mean? Getting more data is never a problem. Getting good data it is.

3) Eq.1 The variable $s_t$ should be defined.

4) After eq.2 "..that collected data o" "o" should be upper case (rather than lower case)

5) In eq.16 and also below in several places , what does the distance $||..||_2$ mean?

6) a closing parenthesis is missing before eq.22

7) Page 12, "The HMM model.."the word "model" is already contained in the acronym

8) page 15, in subsection "Updating the transition probabilities...." starts with a long, mysterious and convoluted sentence that I could not understand. I suggest to rewrite it in more simple terms.

9) page 17, "We use the (improper) Jeffreys prior..." Why is it improper? Is it because it is non-normalizable? Please, specify.

10) page 19, second paragraph: "As an example synthetic model...high-force folded state". It seems to me this is only true in the passive mode hopping condition (where the force changes in the hopping mode). Please, explain in which mode the synthetic simulation was carried out.

11) page 21, "..the predominant contribution to the observed force....or other sources of measurement noise". I do not quite agree with this statement. Have the authors estimated the expected change in stiffness upon releasing of the folded ssRNA? Did they really expect a marked decrease in the stiffness of the tether upon unfolding?

12) Ref.57 is incomplete.

13) Fig.1: color code for biotin (blue) and digoxigenin (green) should be exchanged
\color{black}

We thank the referee for their careful attention to detail and willingness to provide these helpful corrections and suggestions where the text could be further clarified.
We have modified the manuscript accordingly.

\color{red}
{\bf Peer Reviewer \#2}

This manuscript addresses the highly relevant issue of how to extract accurate kinetics from single force spectroscopy experiments, as well as on how to obtain reliable error bounds, using a Bayesian approach. The method is developed in detail, and then applied to synthetic data as well as to optical tweezer data on an RNA hairpin.

The paper is very well written, and the mathematical treatment is very clear and convincing. I'm convinced the proposed approach is a significant advance over the established methods used to interpret force probe data, and the authors are to be applauded for this. Personally, I'd very much like to see this method published in BJ, as in my view this and similar types of methodological advancement forms an essential part of Biophysics. However, current Editorial policy unfortunately requires each single paper to immediately present significant new biological insight, which is very marginal in the present manuscript. Due to this unfortunate restriction, BJ will probably not be able to enjoy the well above average number of citations this paper will undoubtedly get in the long run.
\color{black}

We thank the reviewer for their high praise, and hope the Editor is persuaded our paper is sufficiently exceptional to warrant publication in \emph{Biophysical Journal}.

\color{red}
{\bf Peer Reviewer \#3} %(Please explain):

This paper is worthy of publication after some flaws are corrected. These flaws are not all minor (see full review), but I do not expect them to pose any insuperable problems (hence my belief that the paper can already be judged worthy of publication).

%Peer Reviewer \#3 (Supplemental):

%Supplemental Material (with text and 3 figures)

%Peer Reviewer \#3 (Review Text):

Chodera et al present a new approach to analyzing single-molecule force spectroscopy data, using a Bayesian extension to the classic hidden Markov models that have been used successfully by others. What's new here is the use of Bayesian inference, which provides good estimates of the uncertainties and allows the usefulness of extra data collection to be assessed. The authors also point to a number of useful extensions of the scheme at the end of the paper, such as ways to determine the number of states automatically. The authors describe the conceptual and mathematical bases for the model, demonstrate it on simulated data, and finally test it with experimental data.

The manuscript is generally well-written (though see points below). It presents a worthwhile advance in analytical methods which is likely to find application to a range of single-molecule experiments of interest to the biophysics community. It is therefore a good fit for the Biophysical Journal. 
\color{black}

We thank the referee for their support of the publication of this work in \emph{Biophysical Journal}.

\color{red}
However, there are a number of points which should be addressed by the authors before the manuscript is ready for publication.

Major issues:

1. The authors motivate the work in the context specifically of force spectroscopy measurements, where the trajectory of the force is plotted as a function of time for a system under equilibrium. The method seems to be much more general than this, however, and presumably would be applicable to FRET traces, ion channel recordings, and other such single-molecule measurements. Is there some reason that the authors have limited the discussion of the applicability of the technique to force spectroscopy? If so, the reasons should be discussed in the manuscript. Otherwise, the appeal and utility of the manuscript would be greatly enhanced by a discussion (even if brief!) of how the method can be applied to other types of measurements (and any pitfalls that might need to be overcome with other types of data).
\color{black}

The referee is correct in that our Bayesian HMM extension can, in principle, be applied to other single-molecule experiments such as FRET traces.
There is, however, a key distinguishing feature between force spectroscopy and these other measurement methods: In force spectroscopy, the observables (force or extension) are continuous numbers on the real number line, while in the case of FRET traces, instantaneous photon arrival times and colors (often along with other properties) are recorded.
This necessitates binning the data in order to produce a continuous number, a procedure that does not retain all of the information encoded in the photon arrival times.
Indeed, this binning procedure has been shown to reduce the amount of information that can be extracted from the data {\color{blue}[JDC: insert reference from Frank No\'{e} here]}, which is why we feel we should not recommend this binning procedure.
Instead, we focus on force spectroscopy, where the measurements are inherently performed on a continuous observable.
We now note this in the manuscript.
{\color{blue}[JDC: Note this. Problem also occurs for ion channel recording.] }

\color{red}
2. Whenever a new method is demonstrated, it is best to show that it "works" by testing it on a dataset for which the results are known from previously-established methods, and then to demonstrate it for data that can't be analyzed well by previous methods. The authors have effectively done the first (with the simulated data, though this isn't as stringent a test as using real experimental data, see point 4), but they haven't quite done the second--they still need to show that their method works on datasets where others don't. They could show this easily using the dataset in Figure 3, for example comparing to the results obtained when the states are separated into different regions of force. 
\color{black}

We have added a much more difficult test of the BHMM method for a two-state system with strongly overlapping states, and show how the standard analysis technique of dividing the states by a threshold and estimating the rate constant from the number of threshold crossings fails to recover the true rate, while the BHMM method recovers the correct rate estimate and statistical uncertainty in this estimate.
To ensure that we can rigorously test the accuracy of the method, we again use synthetic data for this test. {\color{blue} TODO: new synthetic data test and comparison with thresholding}

\color{red}
3. More specifically, since this paper is about an extension of classic hidden Markov modeling, the authors should provide at some point in the manuscript or supporting information a comparison of the results from the new approach to the results from a standard HMM model (both the values and uncertainties). Some discussion about comparative computational cost would also be helpful.
\color{black}

We have added a discussion of and comparison to the only other practical approach to estimating uncertainties from standard hidden Markov models, which is bootstrapping based on the generation of synthetic data from the MLHMM. {\color{blue} [TODO: comparison with HMM bootstrapping.]}

\color{red}
4. In the simulated data, the authors seem to have included Brownian noise but not drift. Since drift is often a problem with experimental data (not all of which is as nice as the data shown in Fig 3!), the authors should show how robust the method is when dealing with drift that is not insignificant (say, half the spacing between the two higher-force peaks in Fig. 2).
\color{black}

The method we describe will obviously fail in the presence of drift as large as suggested, but it is fortunately straightforward to collect data with sufficiently small drift on current-generation instruments that this is not a significant limitation~\cite{bustamante:csh-protocols:2009:optical-tweezers}.
As an aid to the readers, we have added a section in the Supplementary Material illustrating the behavior on a simple two-state system as a function of drift, so that the reader may more easily understand the pathologies that develop when large amounts of drift are present. {\color{blue} TODO: drift discussion in supplemental info}

{\color{blue}[JDC: Point out that modern systems do not have a problem with drift.]}

\color{red}
5. Another issue with the simulated data and the comparison to the threshold method (Supplementary Material) is that the lifetimes in the simulated data appear to be long enough compared to the filtering window that was used (the shortest lifetime, ~ 10 ms, is many times longer than the 1 ms interval time) that the simulated data could very easily have been filtered enough to reduce the overlap between the states to the point that the threshold method works equally well. Hence this is not a particularly impressive test of the method, nor is it a demonstration that the threshold method doesn't work. The authors should be demonstrating the method for a case in which it would simply not be possible to apply the threshold method.
\color{black}

We have chosen the additional ``demanding'' test described above (in response to point \#2) to address this request as well.

\color{red}
6. The authors mention that the experimental data they use to demonstrate the method, in Figure 3, come from RNA molecules that do not display the "usual" two-state behavior previously reported, but instead a different three-state behavior. This raises a major concern because it seems likely that the folding behavior is no longer the intrinsic behavior of the RNA, but may instead reflect damage to the RNA (for example, during storage as mentioned by the authors, or possibly due to oxidative damage from the laser beam) or some other effect. Effectively, it is not clear what is the true nature of the molecule being measured here. The BHMM method really should be demonstrated on data where it is known what is being measured and the results are repeatable, rather than something where there is a strong concern that the data are in some way artifactual. In principle, it should not matter what the data comes from for testing the method, but in practice the validation is weakened if
the sample is effectively "unknown".
Given that the authors are world experts in force measurements of RNA and protein folding, one would expect that they should have access to lots of other data that would be more appropriate in this regard, and does not suffer from such uncertainty.
\color{black}

As discussed in the response to comments from the other referees, we have made it clear that the main point of this paper is to illustrate the performance and utility of the proposed procedure on real data, and so the choice of systems studied is secondary.
We have clarified our description of this system as well as added another system (apomyoglobin at pH 5) recently described in another paper~\cite{elms:pnas:2012:apo-myoglobin}.

\color{red}
7. Again regarding the experimental data in Fig 3, the authors should show that a 3-state model works better than a 2-state or 4-state model. The authors state (line 602) ``there is a clearly-resolved intermediate state'', but do not show that a single intermediate explains the data better than more than 1 intermediate.
\color{black}

The issue of determining how many states are best supported by the data is a complex one, and outside the scope of the current manuscript, which demonstrates a procedure for inferring the rates, state descriptions, and uncertainties given that the number of states has been decided upon by the experimenter.  
We discuss extensions of this approach to decide the appropriate number of states, but such extensions are sufficiently complex that it would double the manuscript length, as well as the amount of time required to develop the methodology and code. 
The issue of how best to decide the appropriate number of states supported by the data is a matter of ongoing research in the field, and we have done our best to describe this in our Discussion.

\color{red}
8. The authors should be more careful in their discussion in the introduction of the difficulties that can occur when analyzing transitions (p1, last paragraph). In a great many cases, simple analysis schemes, for example using a threshold to divide the observed force or extension range into regions (as discussed here), do not lead to "a high degree of state mis-assignment" as claimed by the authors. Such problems occur in specific situations, such as for the data in Fig 2 and 3, where there is substantial overlap between the states. When the signals from different states are well-resolved, there may be no need for applying any more complex analysis scheme. Since this issue relates directly to the conditions under which the approach proposed by the authors would be better than other approaches, the authors should correct their discussion and make it more nuanced.
\color{black}

We have amended the discussion in the suggested manner. {\color{blue} [TODO]}

\color{red}
9. Overall, the authors' description of the model would be more accessible to a general audience of biophysicists if the various steps were motivated physically, not just mathematically. While each instance of this problem is minor in itself (see some examples listed in the "minor issues" below), together they make the paper more difficult to understand.
\color{black}

We hope the reorganized manuscript (where much algorithmic detail has been moved to the Appendix and illustrative examples have been moved forward) makes the revised manuscript much more accessible to a general biophysics audience.

\color{red}
Minor issues:

10. Several variables in Equation 1 (P, $\phi$) are not explicitly defined in the text when the equation is first presented.

11. Authors must define acronyms when they are first used (MLHMM, BHMM, $\ldots$).

12. The use of italics for emphasis in the text is irritating and unnecessary, it should be removed.

13. The "first approximation" cited on line 305 (p 4) should be motivated physically. This seems to be the equivalent of setting thresholds; if so it should be pointed out (and provides a point of comparison for performance of the methods), if not it should be explained.
\color{black}

The first approximation previously described was indeed motivated by the threshold approach, and as suggested by the referee, can be used to provide a point of comparison between the initial (threshold) model, the maximum likelihood HMM, and the Bayesian HMM.  {\color{blue}[JDC: Mention if we decide to do this for synthetic data in the SI.]}

To increase the robustness of our approach in handling pathological situations noted after initial submission, we now use a new initialization approach in which the likelihood maximization approach is initialized multiple times from randomized initial models, and the model with the highest log likelihood selected for the maximum likelihood HMM. 
This new procedure is now described in the manuscript.

{\color{blue}[JDC: Make changes to manuscript and code.]}

\color{red}
14. Line 312: the authors should explain what is improved over the initial approximation, and why.

15. Line 326: the authors should explain what "Algorithm 1" does, and why it is used.
\color{black}

We have taken these suggestions and made appropriate corrections.

\color{red}
16. Figures 2 and 3 are very difficult to read. The authors should show a zoom into a small region of the graphs, so that the readers may see the transitions more clearly. The figures would also be easier to parse if the individual data points were connected by lines, so that readers could see the data as a trajectory that evolves in time rather than scattered points. This would also make the transitions between states clearer.
\color{black}

We strongly disagree regarding the connection with lines, as this can deceive the eye into perceiving features that do not exist in the data, which is more properly represented by ordered points.

We do, however, now provide zoomed regions of these plots as Supplementary figures. {\color{blue} [TODO: supplementary figure]}

\color{red}
17. Show Figure 3 in the Supplementary Material on a log scale (the standard way to do it for correlation functions, since the exponential decays can be seen more readily).
\color{black}

We now show Supplementary Figure 3 on both normal and logarithmic scales. {\color{blue} [TODO: supplementary figure]}

\bibliographystyle{cell}
\bibliography{../bayesian-hmm-method/bhmm}

\closing{Kind regards,}
\end{letter}
\end{document}

### 2015-02-16 manuscripts/bayesian-hmm-method/bhmm.tex
Initial import from http://simtk.org/home/bhmm Adding everything but experimental datasets.

% PRL look and style (easy on the eyes)

\documentclass[aps,pre,twocolumn,superscriptaddress,nofootinbib,longbibliography]{revtex4-1}
%\documentclass[prl,twocolumn,superscriptaddress,nofootinbib,linenumbers,longbibliography]{revtex4-1}
%\documentclass[prl,twocolumn,superscriptaddress,nofootinbib,linenumbers]{revtex4-1}
%\documentclass[preprint,endfloats,prl,superscriptaddress,nofootinbib,linenumbers]{revtex4-1}
% Two-column style (for submission/review/editing)
%\documentclass[aps,preprint,prl,superscriptaddress,showpacs]{revtex4}

\usepackage{palatino}
\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{boxedminipage}
\usepackage{verbatim}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}

\graphicspath{{figures/}}

%%% FROM NINA

%\usepackage{fullpage}
%\usepackage{amsmath}

\newcommand{\mus}{\mu_{1 \dots n}}
\newcommand{\sigmas}{\sigma_{1 \dots n}}
\newcommand{\obsv}[1]{A_{1 \dots #1}}
\newcommand{\states}[1]{s_{1 \dots #1}}

%%%%%%%%%%% OLD STUFF (UNUSED?) %%

% italicized boldface for math (e.g. vectors)
\newcommand{\bfv}[1]{{\mbox{\boldmath{$#1$}}}}
% non-italicized boldface for math (e.g. matrices)
\newcommand{\bfm}[1]{{\bf #1}}

%\newcommand{\bfm}[1]{{\mbox{\boldmath{$#1$}}}}
%\newcommand{\bfm}[1]{{\bf #1}}
\newcommand{\expect}[1]{\left \langle #1 \right \rangle}                % <.> for denoting expectations over realizations of an experiment or thermal averages

% vectors
%\newcommand{\x}{\bfv{x}}
%\newcommand{\y}{\bfv{y}}
%\newcommand{\f}{\bfv{f}}
%\newcommand{\z}{\bfv{z}}
%\newcommand{\p}{\bfv{p}}
%\newcommand{\q}{\bfv{q}}
%\newcommand{\T}{\mathrm{T}}

\newcommand{\op}[1]{\mathcal{#1}}
%\newcommand{\vektor}[1]{{\bf #1}}
%\renewcommand{\vec}[1]{\bfm{#1}} % italicized boldface
%\renewcommand{\vec}[1]{{\bf #1}}
\renewcommand{\vec}[1]{{\mbox{\boldmath{$#1$}}}}
\newcommand{\q}{\vec{q}} % coordinates
\newcommand{\p}{\vec{p}} % momenta
\newcommand{\z}{\vec{z}} % phase space point
\newcommand{\x}{\vec{x}} % general vector x
\newcommand{\M}{\bfm{M}} % diagonal mass matrix
\newcommand{\grad}{\nabla}
\newcommand{\timeavg}[1]{\overline{#1}}                    % time average over a trajectory

\newcommand{\bfc}{\bfm{c}}
\newcommand{\hatf}{\hat{f}}

\newcommand{\bTheta}{\bfm{\Theta}}
\newcommand{\btheta}{\bfm{\theta}}
\newcommand{\bhatf}{\bfm{\hat{f}}}
\newcommand{\Cov}[1] {\mathrm{cov}\left( #1 \right)}
\newcommand{\Ept}[1] {{\mathrm E}\left[ #1 \right]}
\newcommand{\Eptk}[2] {{\mathrm E}\left[ #2 \,|\, #1\right]}
\newcommand{\T}{\mathrm{T}}                                % T used in matrix transpose
\newcommand{\qed}{\hfill \mbox{\raggedright \rule{.07in}{.1in}}}  % QED mark
\newcommand{\var}[1]{\mathrm{var} \, #1}
\newcommand{\cov}[2]{\mathrm{cov}\left( {#1}, {#2} \right)}

%% DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%% TITLE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Bayesian hidden Markov model analysis of single-molecule force spectroscopy: Characterizing kinetics under measurement uncertainty}
%\title{Characterizing the uncertainty in equilibrium kinetics from single-molecule force spectroscopy measurements using Bayesian hidden Markov models}
% JDC: Possible subtitle: Characterizing metastable states and transition rates under measurement uncertainty

\author{John D.~Chodera}
 \thanks{Corresponding author}
% \email{jchodera@berkeley.edu}
% \affiliation{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720, USA}
 \email{john.chodera@choderalab.org}
 \affiliation{Computational Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, NY 10065, USA}

\author{Bettina Keller}
\email{bettina.keller@fu-berlin.de}
 \affiliation{DFG Research Center Matheon, FU Berlin, Arnimallee 6, 14195 Berlin, Germany}

\author{Phillip J.~Elms}
 \email{elms@berkeley.edu}
 \affiliation{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720, USA}
 \affiliation{Biophysics Graduate Group, University of California, Berkeley, CA 94720, USA}
 \affiliation{Jason L.~Choy Laboratory of Single Molecule Biophysics, Institute for Quantitative Biosciences, University of California, Berkeley, CA 94720, USA}

\author{Frank No\'{e}}
 \email{frank.noe@fu-berlin.de}
 \affiliation{DFG Research Center Matheon, FU Berlin, Arnimallee 6, 14195 Berlin, Germany}

\author{Christian M.~Kaiser}
 \email{cmkaiser@berkeley.edu}
 \affiliation{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720, USA}
 \affiliation{Department of Physics, University of California, Berkeley, CA 94720, USA}

\author{Aaron Ewall-Wice}
% \email{olivercromwell@uchicago.edu}
 \affiliation{University of Chicago, IL 60637, USA}

\author{Susan Marqusee}
% \email{marqusee@berkeley.edu}
 \affiliation{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720, USA}
 \affiliation{Department of Molecular \& Cell Biology, University of California, Berkeley, CA 94720, USA}
 \affiliation{Jason L.~Choy Laboratory of Single Molecule Biophysics, Institute for Quantitative Biosciences, University of California, Berkeley, CA 94720, USA}

\author{Carlos Bustamante}
% \email{carlos@alice.berkeley.edu}
 \affiliation{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720, USA}
 \affiliation{Department of Molecular \& Cell Biology, University of California, Berkeley, CA 94720, USA}
 \affiliation{Jason L.~Choy Laboratory of Single Molecule Biophysics, Institute for Quantitative Biosciences, University of California, Berkeley, CA 94720, USA}
 \affiliation{Department of Physics, University of California, Berkeley, CA 94720, USA}
 \affiliation{Department of Chemistry, University of California, Berkeley, CA 94720, USA}
 \affiliation{Howard Hughes Medical Institute, University of California, Berkeley, CA 94720, USA}

\author{Nina Singhal Hinrichs}
 \email{nshinrichs@uchicago.edu}
 \affiliation{Departments of Statistics and Computer Science, University of Chicago, IL 60637, USA}

\date{\today}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT/pacs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}

Single-molecule force spectroscopy has proven to be a powerful tool for studying the kinetic behavior of biomolecules.
Through application of an external force, conformational states with small or transient populations can be stabilized, allowing them to be characterized and the statistics of individual trajectories studied to provide insight into biomolecular folding and function.
Because the observed quantity (force or extension) is not necessarily an ideal reaction coordinate, individual observations cannot be uniquely associated with kinetically distinct conformations.
While maximum-likelihood schemes such as hidden Markov models have solved this problem for other classes of single-molecule experiments by using temporal information to aid in the inference of a sequence of distinct conformational states, these methods do not give a clear picture of how precisely the model parameters are determined by the data due to instrument noise and finite-sample statistics, both significant problems in force spectroscopy.
We solve this problem through a Bayesian extension that allows the experimental uncertainties to be directly quantified, and build in detailed balance to further reduce uncertainty through physical constraints.
We illustrate the utility of this approach in characterizing the {\color{red}observed multistate kinetics of two molecules in a stationary optical trap: an RNA hairpin that exhibits three-state behavior and a protein system (apomyglobin at pH 5) that exhibits two-state behavior with highly overlapping force distributions characterizing each state.}

%\emph{Keywords: force spectroscopy; optical tweezers; single-molecule experiments; hidden Markov model (HMM); Bayesian analysis; statistical error; statistical uncertainty}

\end{abstract}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MAKE FONT SMALLER TO MATCH JOURNAL STYLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\small

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% INTRODUCTION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\label{section:introduction}

Recent advances in biophysical measurement have led to an unprecedented ability to monitor the dynamics of single biological macromolecules, such as proteins and nucleic acids~\cite{ritort:j-phys:2006:single-molecule-review}.
As a new approach to probing the behavior of biological macromolecules, these experiments promise to change the way we study folding, dynamics, catalysis, association, transcription, translation, and motility, providing otherwise-inaccessible information about microscopic kinetics, energetics, mechanism, and the stochastic heterogeneity inherent in these processes.
Advances in instrumentation for optical force spectroscopy in particular have produced instruments of extraordinary stability, precision, and temporal resolution~\cite{block:rev-sci-instrum:2004:optical-trapping,bustamante:annu-rev-biochem:2008:optical-tweezers} that have already demonstrated great utility in the study of biomolecules in the presence of externally perturbative forces~\cite{bustamante:nature:2003:dna-mechanics,block:curr-opin-chem-biol:2008:rna,bustamante:cell:2011:central-dogma}. 
Under external force, it becomes possible to stabilize and characterize short-lived conformational states, such as protein folding and unfolding intermediates~\cite{clarke:jmb:2002:titin-afm,cecconi-shank-bustamante-marqusee:science:2005:rnase-h,rief:pnas:2010:pmf}.

In a typical single-molecule optical trapping experiment, a protein or nucleic acid is tethered to two polystyrene beads by dsDNA handles that {\color{red}serve as spacers to} prevent the molecule under study from interacting with the beads {({\color{red} as in Figure~\ref{figure:trap-configuration})}.
The handle-biomolecule-handle assembly---referred to as a fiber---is associated with the beads through tight noncovalent interactions, with one bead held in an optical trap and the other either suctioned to a micropipette or held in a second optical trap.
During an experiment, the position of the bead within the laser trap is monitored, and either the relative displacement from the trap center or the total force on the bead is recorded, resulting in a timeseries such as the one depicted in Fig.~\ref{figure:p5ab-observed-trajectory}.
{\color{red}The instrument can generally be operated in several modes: a force ramp mode, in which the trap is translated rapidly enough to potentially carry the system {\color{red}transiently} out of equilibrium; a passive mode, in which the trap is held fixed and the equilibrium fluctuations of the system are observed; and a constant force-feedback mode, in which the trap is continually repositioned to maintain a specified average constant force on the fiber.}
Here, we concern ourselves with the latter two classes of experiment, though nonequilibrium experiments remain an exciting topic of active research~\cite{ritort:adv-chem-phys:2008:nonequilibrium-review}.

Often, the dynamics observed in these experiments appears to be dominated by stochastic transitions between two or more strongly metastable conformational states~\cite{schuette-huisinga:2002:biomolecular-conformations-as-metastable-states,chodera-singhal:jcp:2007:automatic-state-decomposition}---regions of conformation space in which the system remains for long times before making a transition to another conformational state.
{\color{red}The observed force trace in Fig.~\ref{figure:p5ab-observed-trajectory}, for example, demonstrates apparent hopping behavior between two or more such states.}
{\color{red}Transitions among strongly metastable states such as these} are generally well-described by first-order kinetics~\cite{noe:pnas:2009:ww-domain}.
While visual inspection of the dynamics may suggest the clear presence of multiple metastable states, quantitative characterization of these states is often difficult.
First, the observed force or extension is unlikely to correspond to a true reaction coordinate easily able to separate all metastable states~\cite{best-paci-hummer-dudko:jpcb:2008:pulling-coordinate,rief-paci:pre:2010:multidimensional-landscape,thirumalai:2011:prl:pfold,chodera-pande:arxiv:2011:single-molecule-pfold}, and second, measurement noise may further broaden the force or extension signatures of individual states, increasing their overlap.
{\color{red} As a result, a histogram of the observed temporal trace can have strongly overlapping state distributions, as in Fig.~\ref{figure:p5ab-observed-trajectory}, right.}
Attempting to separate these states by simply dividing the observed force or extension range into regions, following current practice~\cite{woodside:science:2006:dna-hairpin-optical-trap,ritort:biophys-j:2011:short-handles}, can often lead to a high degree of state mis-assignment that results in the estimated rate constants and state distributions containing a significant amount of error~\cite{chodera:rate-theory} (see \emph{Supplementary Material: Comparison with threshold model}).

Hidden Markov models (HMMs)~\cite{rabiner:ieee-proceedings:1989:hmm-tutorial}, which use temporal information in addition to the instantaneous value of the observable (force or extension) to determine which conformational states the system has visited during the experiment, have provided an effective solution to the hidden state problem in many other classes of single-molecule experiments, such as ion channel currents~\cite{becker:pfluegers-arch:1994:ion-channel-hmm,sachs:biophys-j:2000:ion-channel-hmm,degunst-kuensch-schouten:j-am-stat-assoc:2001:ion-channel-hmm,qin:biophys-j:2004:ion-channel-hmm}, single-molecule FRET~\cite{andrec-levy-talaga:jpca:2003:photon-arrival-hmm,mckinney-joo-ha:biophys-j:2006:hmm-fret,lee:jpcb:2009:hmm-fret,scherer-dinner:jpcb:2009:rna-fret-hmm,gopich-szabo:jpcb:2009:decoding-fret}, and the stepping of motor proteins~\cite{smith:biophys-j:2001:hmm-actomyosin,sachs:biophys-j:2006:motor-hmm,sigworth:biophys-j:2010:hmm-motors}.
In applying hidden Markov modeling to the analysis of single-molecule force spectroscopy data, the observed force or extension trace is assumed to come from a realization of an underlying Markov chain, where the system makes history-independent transitions among a set of discrete conformational states with probabilities governed by a transition or rate matrix.
Data, in the form of force or bead-to-bead extension measurements, is sampled at an interval that ensures that sequential observations satisfy the Markov property of history-independence, though the appropriate interval depends on the properties of the experimental configuration.
Under a given set of external force conditions, each state has a distribution of forces or extensions associated with it.
Given observed timeseries data for forces or extensions, the maximum likelihood estimate (MLE) of the model parameters (transition rates and state force or extension distributions) and sequence of hidden states corresponding to the observed data {\color{red} (used to color the points in Fig.~\ref{figure:p5ab-observed-trajectory})} can be determined by standard methods~\cite{baum:1970:ann-math-statist:baum-welch,viterbi:1967:IEEE-trans-info-theory:viterbi-algorithm}, as demonstrated in recent work~\cite{kruithof-vannoort:biophys-j:2009:hmm-force}.

\begin{figure}[tbp]
\noindent
\resizebox{\columnwidth}{!}{\includegraphics{optical-trap-configuration-threestate-trimmed}}
\caption{{\bf Single-molecule optical trapping configuration.}
The biomolecule of interest---here, %the p5ab 
{\color{red} an} RNA hairpin---is tethered to two polystyrene beads by dsDNA handles.
The fluctuating force on one bead held in an optical trap is monitored, while the other bead is held suctioned to a micropipette tip.
Conformational transitions of the hairpin---such as transitions among the three kinetically metastable states illustrated here---are observed indirectly through motion of the bead in the trap.
\label{figure:trap-configuration}}
\end{figure}

Unfortunately, this approach has a number of significant drawbacks.
Due to technical limitations, experiments often suffer from limited statistics---the events of interest (transitions between states or visits to rare states) may occur only a few times during the course of the measurement, and data for additional fibers is time-consuming to collect.
As a result, while the MLE yields the most likely set of model parameters, there may be enormous uncertainty in some of these parameters, and the uncertainty in multiple parameters may be correlated in complex nonlinear ways.
While methods exist for estimating the standard error or confidence intervals from 
%\mbox{MLHMMs}
{\color{red} hidden Markov models}
~\cite{aittokallio-uusipaikka:technical-report:2000:hmm-standard-error}, these schemes can be prohibitively costly for long traces, and may still significantly underestimate the statistical error for short traces due to the normally-distributed error approximation inherent in the approach.
{\color{blue}[JDC: Should we say something about bootstrap here since we will compare to it later?]}
The high cost (both in terms of instrument and experimenter time) of collecting additional data also means that it is not a simple task to judge how much data need be collected to test a particular hypothesis in a statistically meaningful way.
Worse yet, the standard algorithms employed to find the MLE may not even find the true maximum likelihood solution, instead converging to a local maximum in likelihood that is far from optimal~\cite{merialdo:ieee-trans-speech-audio-proc:1993:baum-welch-localit}.

Here, we resolve this issue through the use of a Bayesian extension of hidden Markov models~\cite{robert:stat-prob-lett:1993:bayesian-hmm,chib:j-econometrics:1996:bayesian-hmm,scott:j-am-stat-assoc:2002:bayesian-hmm,ryden:bayesian-analysis:2008:bayesian-hmm} applicable to single molecule force experiments.
By sampling over the posterior distribution of model parameters and hidden state assignments instead of simply finding the most likely values, the experimenter is able to accurately characterize the correlated uncertainties in both the model parameters (transition rates and state force or extension distributions) and hidden state sequences corresponding to observed data.
Additionally, prior information (either from additional independent measurements or physical constraints) can be easily incorporated.
We also include a reversibility constraint on the transition matrix---in which microscopic detailed balance is imposed on the kinetics, as dictated by the physics of equilibrium systems~\cite{van-kampen}---which has been shown to significantly reduce statistical uncertainties in data-poor conditions~\cite{noe:jcp:2008:transition-matrix-sampling,metzner-noe-schuette:pre:2009:transition-matrix-sampling}.
The framework we present is based on Gibbs sampling~\cite{geman-geman:ieee-trans-pattern-analysis:1984:gibbs-sampling,jun-s-liu:mcmc}, allowing simple swap-in replacement of models for observable distributions, extension to multiple observables, and alternative models for state transitions.
Additionally, the Bayesian method provides a straightforward way to model the statistical outcome and assess the utility of additional experiments given some preliminary data, providing the experimenter with a powerful tool for assessing whether time spent collecting additional data will yield sufficient reductions in uncertainty to allow competing hypotheses to be differentiated with statistical significance.

{\color{red}To demonstrate the utility of the Bayesian HMM in the analysis of real force spectroscopy data, we illustrate its application to the analysis of an RNA hairpin exhibiting apparent three-state behavior (Figure~\ref{figure:p5ab-observed-trajectory}) and a protein system that exhibits two-state behavior with highly overlapping force signatures (Figure~\ref{figure:apomyoglobin-observed-trajectory}) collected in a passive (equilibrium) optical trap.
In addition to maximum likelihood estimates of equilibrium state probabilities, state-to-state transition rates, and the force distributions characterizing each state, the Bayesian extension provides estimates of confidence intervals for individual parameters (Tables~\ref{table:p5ab-bhmm-confidence-intervals} and \ref{table:apomyoglobin-bhmm-confidence-intervals}) as well as the full joint distribution of parameters.

In the following sections, we briefly describe the theory behind maximum-likelihood hidden Markov models of force spectroscopy data (\emph{Theory}), the validation of our methodology on synthetic data for which the computed confidence intervals can be checked for correctness (\emph{Validation}), and our specific applications to real force spectroscopy data (\emph{Applications}).
We follow with a discussion of future extensions of this approach, and include a complete description of the computational implementation in \emph{Algorithms}.
A Matlab implementation of the approach described here is also made freely available online [\url{http://github.com/choderalab/bhmm}].
}

\begin{figure*}[tb]
\noindent
\resizebox{\textwidth}{!}{\includegraphics{algorithm-workflow/Fig_ProbabilitySpace_trimmed.pdf}}
\caption{{\bf Illustration of the Bayesian hidden Markov model (BHMM) sampling algorithm.}
During each BHMM update, a modified forward-backward algorithm is used to sample a new hidden state trace $\mathbf{S}$, which is then used to generate a new sample of the reversible transition matrix $\mathbf{T}$ using MCMC sampling, followed by sapling of the state emission probabilities $\mathbf{E}$.
The cycle then repeats for each subsequent iteration, which together sample from the Bayesian posterior of the HMM and fully characterize the uncertainty in the kinetic model given the data.
\label{figure:bhmm-workflow}}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEORY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Hidden Markov models for force spectroscopy}
\label{section:theory}

Suppose the temporal behavior of some observable $O(x)$ that is a function of molecular configuration $x$---here, generally force or molecular extension---is observed at temporal intervals $\Delta t$ to produce a timeseries $o_t$, where $t = 0, 1, \ldots, L$.
An instantaneous observation $o_t$ does not necessarily contain enough information to unambiguously identify the current conformational state the molecule occupies; to infer the hidden state, we must also make use of the temporal information in the observed trace.
We restrict ourselves to consideration of scalar functions $O(x)$, but the generalization to multidimensional probes (or multiple probes, such as combined force and fluorescence measurements~\cite{chemla:nature-methods:2011:fluorescence-force}) and multiple observed temporal traces is straightforward.

We presume the system under study has $M$ kinetically distinct states, in the sense that the system generally remains in a given state for several observation intervals $\Delta t$, but these states may not necessarily represent highly populated states of the system at equilibrium.
We treat these conformational states as the hidden states of the model, because we cannot directly observe the identity of the metastable state in which the system resides.
The hidden Markov model presumes the observed data $\bfm{O} \equiv \{o^{}_t\}$ was generated according to the following model dependent on {\color{red} the hidden states $\bfm{S} \equiv \{s_t\}$,} parameters $\bfm{\Theta} \equiv \{\bfm{T}, \bfm{E}\}$, where $\bfm{T}$ is an $M \times M$ row-stochastic transition matrix and $\bfm{E}$ a set of emission parameters governing the observable (force or extension) distributions for each of the $M$ hidden states, prior information about the initial state distribution $\bfv{\rho}^{}$, {\color{red} and emission probability distribution $\varphi(o | \bfm{e})$},
\begin{eqnarray}
%s^{}_0 &\sim& \bfv{\rho} \nonumber \\
%s^{}_t \mid s^{}_{t-1}, \bfm{T} &\sim& T_{s^{}_{t-1} \cdot} \nonumber \\
%o^{}_t \mid s^{}_t, \bfm{e}_{s^{}_{t}} &\sim& \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-process}
%s^{}_0 &\sim& \bfv{\rho} \nonumber \\
%s^{}_t &\sim& T_{s^{}_{t-1} \cdot} \nonumber \\
%o^{}_t &\sim& \varphi(\cdot \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-process}
\mathbb{P}(s^{}_0) &=& \rho^{}_{s^{}_0} \nonumber \\
\mathbb{P}(s^{}_t \mid s^{}_{t-1}, \bfm{T}) &=& T_{s^{}_{t-1} s^{}_{t}} , \:\: t \ge 1 \nonumber \\
\mathbb{P}(o^{}_t \mid s^{}_t, \bfm{e}_{s^{}_{t}}) &=& \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-process} .
\end{eqnarray}
In diagrammatic form, the observed state data $\{o_t\}$ and corresponding hidden state history $\{s_t\}$ can be represented 
\begin{eqnarray}
\begin{array}{llllllllll}
 \stackrel{\bfv{\rho}}{\longrightarrow} & s^{}_{0} & \stackrel{\bfm{T}}{\longrightarrow} & s^{}_{1} & \stackrel{\bfm{T}}{\longrightarrow} & s^{}_{2} & \stackrel{\bfm{T}}{\longrightarrow} & \cdots & \stackrel{\bfm{T}}{\longrightarrow} & s^{}_{L^{}} \\
 & \downarrow \varphi &  & \downarrow \varphi & & \downarrow \varphi & & & & \downarrow \varphi \\
 & o^{}_{0} & & o^{}_{1} & & o^{}_{2} &  & & & o^{}_{L^{}} \\
\end{array}
\end{eqnarray} 
The initial state distribution $\bfv{\rho}^{}$  reflects our knowledge of the initial conditions of the experiment that collected data $\bfm{{\color{red} O}}^{}$.
In the case that the experiment was prepared in equilibrium, $\bfv{\rho}$ corresponds to the equilibrium distribution $\bfv{\pi}$ of the model transition matrix $\bfm{T}$; if the experiment was prepared out of equilibrium, $\bfv{\rho}$ may be chosen to reflect some other prior distribution (e.g.~the uniform prior).

State transitions $(s^{}_{t-1} \rightarrow s^{}_{t})$ are governed by the discrete transition probability $T_{s^{}_{t-1} s^{}_{t}}$.
The Markov property of HMMs prescribes that the probability that a system originally in state $i$ at time $t$ is later found in state $j$ at time $t+1$ is dependent only on knowledge of the state $i$, and given by the corresponding matrix element $T_{ij}$ of the (row-stochastic) transition matrix $\bfm{T}$.
Alternatively, one could instead use the rate matrix $\bfm{K}$, related to the transition matrix $\bfm{T}$ through the equation $\bfm{T} = e^{\bfm{K} \Delta t}$.  
If the processes described by $\bfm{T}$ or $\bfm{K}$ are slow compared to the observation interval $\Delta t$, then we can easily estimate the rate matrix from the associated transition matrix in a way that avoids the matrix logarithm, through the expansion $\bfm{K} \approx (\bfm{T} - \bfm{I}) / \Delta t$, where $\bfm{I}$ denotes the $M \times M$ identity matrix.

The probabilistic ``emission'' of observables from each state $(s^{}_{t} \rightarrow o^{}_{t})$ is governed by the continuous emission probability $\varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}})$, parametrized by observable emission parameters $\bfm{e}$.
For example, in the force spectroscopy applications described here, $\varphi(o \mid \bfm{e}_s)$ is taken to be a univariate normal (Gaussian) distribution parameterized by a mean $\mu$ and variance $\sigma^2$ that characterize each state, such that $\bfm{e}_i \equiv \{\mu_i, \sigma^2_i\}$.
Other choices of observable distribution can easily be substituted in a modular way without affecting the structure of the algorithms presented here.

Given the HMM process specified in Eq.~\ref{equation:hmm-process}, the probability of observing data $\bfm{O}$ given the model parameters $\bfm{\Theta}$ is then,
\begin{eqnarray}
P(\bfm{O} \mid \bfm{\Theta}) &=& \sum_{\bfm{S}}  \rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{t-1} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-likelihood} ,
\end{eqnarray}
where the sum over hidden state histories $\bfm{S}$ is shorthand for
\begin{eqnarray}
\sum_{\bfm{S}} \equiv \sum_{s^{}_{0}=1}^M \sum_{s^{}_{1}=1}^M \cdots \sum_{s^{}_{L^{}}=1}^M .
\end{eqnarray}
If multiple independent traces $\{o_t\}$ are available, the probability $P(\bfm{O} \mid \bfm{\Theta})$ is simply the product of Eq.~\ref{equation:hmm-likelihood} for the independent traces.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Maximum likelihood hidden Markov model (MLHMM)}

The standard approach to construct an HMM from observed data is to compute the maximum likelihood estimator (MLE) for the model parameters $\bfm{\Theta} \equiv \{\bfm{T}, \bfm{E}\}$, which maximize the probability of the observed data $\bfm{O}$ given the model,
\begin{eqnarray}
\hat{\bfm{\Theta}} &=& \arg\max_\bfm{\Theta} P(\bfm{O} \mid \bfm{\Theta}) ,
\end{eqnarray}
yielding MLE estimates of transition matrix $\hat{\bfm{T}}$ and state emission parameters $\hat{\bfm{E}}$.
Typically, determination of the model parameters $\bfm{\Theta}$ is carried out using the Baum-Welch algorithm~\cite{baum:1970:ann-math-statist:baum-welch}.

Once the MLE parameters $\hat{\bfm{\Theta}}$ are determined, the most likely hidden state history that produced the observations $\bfm{O}$ can be determined using these parameters:
\begin{eqnarray}
\hat{\bfm{S}} &=& \arg\max_\bfm{S} P(\bfm{S} \mid \bfm{O}, \hat{\bfm{\Theta}}) .
\end{eqnarray}
This is typically carried out using the Viterbi algorithm~\cite{viterbi:1967:IEEE-trans-info-theory:viterbi-algorithm}, a classic example of dynamic programming.

%Both the Baum-Welch and Viterbi schemes are described in detail in \emph{Algorithms}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Bayesian hidden Markov model (BHMM)}

Instead of simply determining the model that maximizes the likelihood of observing the data $\bfm{O}$ given the model parameters $\bfm{\Theta}$, we can make use of Bayes' theorem to compute the posterior distribution of model parameters given the observed data:
\begin{eqnarray}
P(\bfm{\Theta} \mid \bfm{O}) &\propto& P(\bfm{O} \mid \bfm{\Theta}) P(\bfm{\Theta}) \label{equation:bayes-theorem}.
\end{eqnarray}
Here, $P(\bfm{\Theta})$ denotes a prior distribution that encodes any \emph{a priori} information we may have about the model parameters $\bfm{\Theta}$.
This prior information might include, for example, physical constraints (such as ensuring the transition matrix satisfies detailed balance) or prior rounds of inference from other independent experiments.

Making use of the likelihood (Eq.~\ref{equation:hmm-likelihood}), the model posterior is then given by,
\begin{eqnarray}
P(\bfm{\Theta} \mid \bfm{O}) &\propto& P(\bfm{\Theta}) \sum_{\bfm{S}}\rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{t-1} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}})  \label{equation:full-posterior} .
\end{eqnarray}
Drawing samples of $\bfm{\Theta}$ from this distribution will, in principle, allow the confidence with which individual parameters and combinations thereof are known, given the data (and subject to the validity of the model of Eq.~\ref{equation:hmm-process} in correctly representing the process by which the observed data is generated).
While the posterior $P(\bfm{\Theta}|\bfm{O})$ is complex, we could in principle use a Markov chain Monte Carlo (MCMC) approach~\cite{jun-s-liu:mcmc} to sample it.
In its current form, however, this would be extremely expensive due to the sum over all hidden state histories $\bfm{S}$ appearing in ratios involving Eq.~\ref{equation:full-posterior}.
Instead, we introduce the hidden state histories $\bfm{S}$ as an auxiliary variable, sampling from the augmented posterior,
\begin{eqnarray}
P(\bfm{\Theta}, \bfm{S} \mid \bfm{O}) &\propto& \left[ \rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{t-1} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \right] P(\bfm{\Theta}) . \nonumber \\\label{equation:augmented-posterior} 
\end{eqnarray}
which makes it much less costly to compute the ratios required for MCMC on the augmented $(\bfm{\Theta},\bfm{S})$ parameter space.

If we presume the prior is separable, such that $P(\bfm{\Theta}) \equiv P(\bfm{T}) P(\bfm{E})$, we can sample from the augmented posterior (Eq.~\ref{equation:augmented-posterior}) using the framework of Gibbs sampling~\cite{jun-s-liu:mcmc}, in which the augmented model parameters are updated by cycles of sampling from conditional distributions,
\begin{eqnarray}
\bfm{S}' &\sim& P(\bfm{S} \mid \bfm{T}, \bfm{E}, \bfm{O}) \propto \rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{t-1} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \nonumber \\
\bfm{T}' &\sim& P(\bfm{T} \mid \bfm{E}, \bfm{S}', \bfm{O}) = P(\bfm{T} \mid \bfm{S}') \propto P(\bfm{T}) \prod_{t=1}^{L^{}} T_{{s'}^{}_{t-1} {s'}^{}_{t}} \label{equation:conditional-gibbs} \\
\bfm{E}' &\sim& P(\bfm{E} \mid \bfm{S}', \bfm{T}', \bfm{O}) = P(\bfm{E} \mid \bfm{S}', \bfm{O}) \propto P(\bfm{E}) \prod_{t=0}^{L^{}} \varphi(o^{}_{t} \mid \bfm{e}_{{s'}^{}_{t}}) \nonumber .
\end{eqnarray}
The equalities on the second and third lines reflect the conditional independence of the hidden Markov model defined by Eq.~\ref{equation:hmm-process}.
When only the model parameters $\bfm{\Theta} \equiv \{ \bfm{T}, \bfm{E} \}$ or the hidden state histories $\bfm{S}$ are of interest, we can simply marginalize out the uninteresting variables by sampling from the augmented joint posterior for $\{\bfm{T}, \bfm{E}, \bfm{S}\}$ and examine only the variables of interest.
In addition, the structure of the Gibbs sampling scheme above allows individual components (such as the observable distribution model $\varphi(o \mid \bfm{e})$ or transition probability matrix $\bfm{T}$) to be modified without affecting the structure of the remainder of the calculation.

In the applications presented here, we employ a Gaussian observable distribution model for $\varphi(o \mid \bfm{e})$, which is expected to be a good model of the distribution of observed force or positions $o$ for force spectroscopy experiments,
\begin{eqnarray}
\varphi(o \mid \bfm{e}) &=& \varphi(o \mid \mu, \sigma^2) = \frac{1}{\sqrt{2 \pi} \sigma} \exp\left[-\frac{1}{2}\frac{(o - \mu)^2}{\sigma^2} \right] \label{equation:gaussian-observable} ,
\end{eqnarray}
where $\mu$ is the mean force or extension characterizing a particular state, and $\sigma$ is the standard deviation or width of forces or extensions corresponding to that state.
Other choices for the observable distribution model $\phi(o | \mathbf{e})$ can easily be substituted should they be more appropriate for analyzing particular classes of experiments, and only require substituting an appropriate parameter update step.
We note that marginal posterior distributions of each mean $P(\mu_i | \bfm{O})$ reflect the statistical uncertainty in how well the mean force or position is determined, and need not correspond to the standard deviation $\sigma_i$, which may be much broader (or narrower, depending on the situation).

{\color{red} Detailed descriptions of the algorithms used to fit the maximum likelihood hidden Markov model (MLHMM) and sample from the posterior distribution of the Bayesian hidden Markov model (BHMM) are given in the Appendix.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% VALIDATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}[tbp]
\noindent
\resizebox{\textwidth}{!}{\includegraphics{model-mlhmm-stateassignments}}
\caption{{\bf Synthetic force trajectory and inferred state assignments in MLHMM.}
Observed samples are colored by their hidden state assignments.
Dark horizontal lines terminating in triangles to the right denote state means, while lightly colored bands indicate one standard deviation on either side of the state mean.
The gray histogram on the right side shows the total observed probability of samples, while the colored peaks show the weighted Gaussian output contribution from each state, and the black outline the weighted sum of the Gaussian output contributions from the HMM states.
\label{figure:model-stateassignments}}
\end{figure*}

\begin{table*}
\caption{{\bf Estimated mean model parameters and confidence intervals for synthetic timeseries data}}
\label{table:synthetic-confidence-intervals}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lccccc}
\hline
&  &  & \multicolumn{3}{c}{\bf Estimated Model Parameters}  \\ \cline{4-6}
\multicolumn{2}{l}{\bf Property} & \bf True Value & \bf 1 000 observations & \bf 10 000 observations & \bf 100 000 observations\\ \hline
stationary probability & $\pi_{1}$ & $0.308$ & $0.228_{\:0.074}^{\:0.480}$ & $0.318_{\:0.244}^{\:0.407}$ & $0.324_{\:0.292}^{\:0.355}$ \\
& $\pi_{2}$ & $0.113$ & $0.093_{\:0.042}^{\:0.172}$ & $0.124_{\:0.098}^{\:0.155}$ & $0.112_{\:0.104}^{\:0.121}$ \\
& $\pi_{3}$ & $0.579$ & $0.679_{\:0.415}^{\:0.870}$ & $0.558_{\:0.455}^{\:0.648}$ & $0.564_{\:0.531}^{\:0.599}$ \\
\hline
transition probability & $T_{11}$ & $0.980$ & $0.970_{\:0.945}^{\:0.987}$ & $0.972_{\:0.966}^{\:0.978}$ & $0.979_{\:0.978}^{\:0.981}$ \\
& $T_{12}$ & $0.019$ & $0.023_{\:0.009}^{\:0.045}$ & $0.026_{\:0.021}^{\:0.032}$ & $0.020_{\:0.018}^{\:0.021}$ \\
& $T_{13}$ & $0.001$ & $0.007_{\:0.001}^{\:0.018}$ & $0.002_{\:0.001}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
& $T_{21}$ & $0.053$ & $0.054_{\:0.018}^{\:0.106}$ & $0.067_{\:0.053}^{\:0.082}$ & $0.057_{\:0.052}^{\:0.061}$ \\
& $T_{22}$ & $0.900$ & $0.868_{\:0.790}^{\:0.931}$ & $0.890_{\:0.870}^{\:0.907}$ & $0.897_{\:0.892}^{\:0.903}$ \\
& $T_{23}$ & $0.050$ & $0.078_{\:0.035}^{\:0.136}$ & $0.043_{\:0.033}^{\:0.056}$ & $0.046_{\:0.042}^{\:0.050}$ \\
& $T_{31}$ & $0.001$ & $0.002_{\:0.000}^{\:0.006}$ & $0.001_{\:0.000}^{\:0.002}$ & $0.001_{\:0.000}^{\:0.001}$ \\
& $T_{32}$ & $0.009$ & $0.010_{\:0.004}^{\:0.019}$ & $0.010_{\:0.007}^{\:0.012}$ & $0.009_{\:0.008}^{\:0.010}$ \\
& $T_{33}$ & $0.990$ & $0.988_{\:0.978}^{\:0.995}$ & $0.990_{\:0.987}^{\:0.992}$ & $0.990_{\:0.989}^{\:0.991}$ \\
\hline
state mean force (pN) & $\mu_{1}$ & $3.000$ & $2.947_{\:2.812}^{\:3.082}$ & $2.998_{\:2.963}^{\:3.033}$ & $3.001_{\:2.990}^{\:3.013}$ \\
& $\mu_{2}$ & $4.700$ & $4.666_{\:4.612}^{\:4.721}$ & $4.699_{\:4.683}^{\:4.716}$ & $4.702_{\:4.696}^{\:4.707}$ \\
& $\mu_{3}$ & $5.600$ & $5.597_{\:5.583}^{\:5.614}$ & $5.602_{\:5.596}^{\:5.607}$ & $5.602_{\:5.600}^{\:5.603}$ \\
\hline
state std dev force (pN) & $\sigma_{1}$ & $1.000$ & $1.037_{\:0.951}^{\:1.134}$ & $0.992_{\:0.967}^{\:1.018}$ & $0.999_{\:0.991}^{\:1.007}$ \\
& $\sigma_{2}$ & $0.300$ & $0.254_{\:0.217}^{\:0.300}$ & $0.287_{\:0.275}^{\:0.300}$ & $0.301_{\:0.296}^{\:0.305}$ \\
& $\sigma_{3}$ & $0.200$ & $0.200_{\:0.190}^{\:0.211}$ & $0.203_{\:0.199}^{\:0.207}$ & $0.201_{\:0.200}^{\:0.203}$ \\
\hline
\end{tabular*}
\end{table*}

%\begin{figure}[tbp]
%\noindent
%\resizebox{\columnwidth}{!}{\includegraphics{confidence-intervals-random}}
%\caption{{\bf Validation of confidence intervals using randomly-generated synthetic test data.}
%Observed confidence intervals (points) are plotted as a function of the desired confidence intervals for equilibrium probabilities ($\pi_i$), transition probabilities ($T_{ij}$), state means ($\mu_i$), and state standard deviations ($\sigma_i$).
%The black diagonal line indicates perfect agreement between expected and observed confidence intervals, while observed confidence intervals above the diagonal indicate overestimates of the uncertainty, and below the diagonal indicate underestimates.
%Because only 50 random models were evaluated, error bars denote a 95\% confidence interval in the estimated observed confidence intervals.
%\label{figure:validation-confidence-intervals}}
%\end{figure}

\section{Validation using synthetic data}
\label{section:model-system}

To verify that our BHMM posterior sampling scheme reflects the true uncertainty in the model parameters, we can test that the scheme produces the expected estimates for synthetic data generated from a model with known parameters $\bfm{\Theta}^*$. 
Given observed data $\bfm{O}$ generated from $P(\bfm{O} \mid \bfm{\Theta}^*)$, sampling from the posterior $P(\bfm{\Theta} \mid \bfm{O})$ using the scheme described in \color{blue} \emph{Sampling from the posterior of the BHMM} \color{black} will provide us with confidence intervals $[\theta_\mathrm{low}, \theta_\mathrm{high}]$ for a specified confidence interval $\alpha \in [0,1]$.
If these computed confidence intervals are accurate, we should find that the true model parameter $\theta^*$ lies in the computed confidence interval $[\theta^{(\alpha)}_\mathrm{low}, \theta^{(\alpha)}_\mathrm{high}]$ with probability $\alpha$.
A 95\% confidence interval ($\alpha = 0.95$), for example, should contain the BHMM estimate in 95\% of such trials.
This can be tested by generating synthetic observed data $\bfm{O}$ from $P(\bfm{O} \mid \bfm{\Theta}^*)$ and verifying that we find $\theta^* \in [\theta^{(\alpha)}_\mathrm{low}, \theta^{(\alpha)}_\mathrm{high}]$ in a fraction $\alpha$ of these synthetic experiments.

\color{red}
As an illustration of the computation of confidence intervals, we simulated a three-state system intended to mimic a protein with (1) a highly-compliance, low-force unfolded state, (2) a moderately compliant low-population intermediate at intermediate force, and (3) a low-compliance, high-force folded state.
Here, the term ``compliance'' refers to the width of the force or extension distribution characterizing the state.
Parameters of the model are given in Table~\ref{table:synthetic-confidence-intervals}, and the observation interval was taken to be $\tau = 1$ ms.
An example realization of a model trajectory mimicking a passive-mode experiment---in which the trap was stationary---is shown in Figure~\ref{figure:model-stateassignments} along with the MLHMM state assignment.
\color{black}
We generated a trajectory of 100 000 observations, and characterized the BHMM mean parameter estimate and 95\% confidence intervals for a subset of this trajectory of varying lengths.
The results, shown in Table~\ref{table:synthetic-confidence-intervals}, show that the confidence intervals contract as trajectory length increases, as expected, and the BHMM-computed 95\% confidence intervals contain the true model parameters with the expected statistics.
\color{red}
In contrast, a model created from simply segmenting the observed forces into disjoint region and assigning state membership based on the force value alone estimates model parameters with significant bias even for 1 000 000 observations (see \emph{Supporting Information}).
\color{black}

As a more rigorous test, we sampled 50 random models from the prior $P(\bfm{\Theta})$ with two to six states, generated a 10 000 observation synthetic trajectory for each, and accumulated statistics on the observed fraction of time the true model parameters were within the BHMM confidence intervals for various values of the confidence interval width $\alpha$.
The results of this test are depicted in \emph{Supplementary Figure 1}. %~\ref{figure:validation-confidence-intervals}.
We expect that the plot traces the diagonal if the observed and expected confidence intervals are identical; an overestimate of the confidence interval will be above the diagonal, and an underestimate will fall below it.
Because only a finite number of independent replicates of the experiment are conducted, there is some associated uncertainty with the observed confidence intervals.
The results show that the observed confidence intervals line up with the expected confidence intervals to within statistical error, suggesting the BHMM confidence intervals neither underestimate nor overestimate the actual uncertainty in model parameters.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% p5ab HAIRPIN
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}[tbp]
\noindent
\resizebox{\textwidth}{!}{\includegraphics{fiber3-trace11-mlhmm-stateassignments}}
\caption{{\bf Experimental force trajectory of %the p5ab 
{\color{red} an RNA} hairpin and MLHMM state assignments.}
Observed samples are colored by their hidden state assignments.
Dark horizontal lines terminating in triangles to the right denote state means, while lightly colored bands indicate one standard deviation on either side of the state mean.
The gray histogram on the right side shows the total observed probability of samples, while the colored peaks show the weighted Gaussian output contribution from each state, and the black outline the weighted sum of the Gaussian output contributions from the HMM states.
\label{figure:p5ab-observed-trajectory}}
\end{figure*}

\begin{table}
\caption{{\bf BHMM model estimates for %p5ab 
{\color{red} an RNA} hairpin data.}
}
\label{table:p5ab-bhmm-confidence-intervals}
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}lcc}
\hline
\multicolumn{2}{l}{\bf Property} & \bf Value\\ \hline
Equilibrium probability & $\pi_{1}$ & $0.215_{\:0.193}^{\:0.236}$ \\
& $\pi_{2}$ & $0.046_{\:0.041}^{\:0.050}$ \\
& $\pi_{3}$ & $0.740_{\:0.717}^{\:0.762}$ \\
\hline
Transition probability ($\Delta t = 1$ ms) & $T_{11}$ & $0.954_{\:0.950}^{\:0.959}$ \\
& $T_{12}$ & $0.033_{\:0.029}^{\:0.037}$ \\
& $T_{13}$ & $0.013_{\:0.011}^{\:0.015}$ \\
& $T_{21}$ & $0.154_{\:0.139}^{\:0.169}$ \\
& $T_{22}$ & $0.650_{\:0.627}^{\:0.673}$ \\
& $T_{23}$ & $0.196_{\:0.180}^{\:0.216}$ \\
& $T_{31}$ & $0.004_{\:0.003}^{\:0.004}$ \\
& $T_{32}$ & $0.012_{\:0.011}^{\:0.013}$ \\
& $T_{33}$ & $0.984_{\:0.983}^{\:0.985}$ \\
\hline
State force mean (pN) & $\mu_{1}$ & $12.549_{\:12.544}^{\:12.552}$ \\
& $\mu_{2}$ & $13.016_{\:13.006}^{\:13.027}$ \\
& $\mu_{3}$ & $13.849_{\:13.848}^{\:13.852}$ \\
\hline
State force std dev (pN) & $\sigma_{1}$ & $0.210_{\:0.207}^{\:0.213}$ \\
& $\sigma_{2}$ & $0.201_{\:0.193}^{\:0.208}$ \\
& $\sigma_{3}$ & $0.213_{\:0.211}^{\:0.214}$ \\
\hline \hline
Transition rate (s$^{-1}$) & $k_{12}$ & $41.4_{\:36.3}^{\:46.6}$ \\
& $k_{13}$ & $9.1_{\:7.2}^{\:11.3}$ \\
& $k_{21}$ & $194.7_{\:173.1}^{\:216.7}$ \\
& $k_{23}$ & $243.7_{\:219.0}^{\:271.5}$ \\
& $k_{31}$ & $2.6_{\:2.1}^{\:3.2}$ \\
& $k_{32}$ & $15.0_{\:13.4}^{\:16.6}$ \\
\hline
State mean lifetime (ms) & $\tau_{1}$ & $21.9_{\:20.0}^{\:24.1}$ \\
& $\tau_{2}$ & $2.9_{\:2.7}^{\:3.1}$ \\
& $\tau_{3}$ & $63.1_{\:58.4}^{\:68.5}$ \\
\hline
\end{tabular*}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPLICATIONS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Applications}
\label{section:application}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% apomyoglobin
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Apomyoglobin at pH 5 in a passive optical trap}
\label{section:application-apomyoglobin}

{\color{red}{\bf [JDC: This section remains to be completed.]}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% RNA hairpin
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{RNA hairpin in a passive optical trap}
\label{section:application-rna-hairpin}

We illustrate the BHMM approach applied to real force spectroscopy data by characterizing the average forces and transition rates among kinetically distinct states of {\color{red} an} RNA hairpin in an optical trap under passive (equilibrium) conditions.

\color{red}
A sample of p5ab RNA hairpin from \emph{Tetrahymena thermophilia} utilized in a previous study~\cite{wen-manosas:biophys-j:2007:rna-optical-tweezers} was provided by Jin-Der Wen, and prepared as previously described~\cite{wen-manosas:biophys-j:2007:rna-optical-tweezers}.  
Within the population of RNA hairpin molecules in the examined sample, there were two chemically distinct species, exhibiting either apparent two-state (as reported previously~\cite{wen-manosas:biophys-j:2007:rna-optical-tweezers}) or three-state behavior (studied here).  
For the purposes of testing this method, we examined a fiber that appeared to consistently exhibit three-state behavior upon visual inspection of the force timeseries data.

The instrument used in this experiment was a dual-beam counter-propagating optical trap with a spring constant of 0.08 pN/nm.  
A piezoactuator controlled the position of the trap and allowed position resolution to within 0.5 nm~\cite{bustamante-smith:2006:minitweezers-patent}.  
The instrument was used in passive (equilibrium) mode, in which the trap remained stationary relative to the pipette into which one bead of the tether was suctioned (corresponding to the geometry depicted in Fig.~\ref{figure:trap-configuration}).
The voltage on the position-sensitive detectors, reporting on the force on the bead held in the optical trap, was recorded at 50 kHz using a 32-bit ADC operating with $\sim$600 ADC units/pN sensitivity.
Each recorded experimental trace was 60 s in duration.
From averaging forces over 1 s intervals at the initial and final parts of each trace, average force drift in the instrument was estimated to be 0.0015(5) pN/s.  
The appropriate observation interval (1 ms) for BHMM analysis was determined by examination of force autocorrelation functions, which suggested that intrastate motion had completely decayed by 1 ms (see \emph{Supplementary Material: Choice of observation interval}).
The observed 50 kHz force traces were therefore subsampled  to 1 kHz for analysis by retaining every 50th sample.
\color{black}

A single observed force trajectory at a fixed trap position adequate to cause hopping among multiple states is shown in Figure~\ref{figure:p5ab-observed-trajectory}.
The most likely state trajectory from the MLHMM fit with three states is shown by coloring the observations most likely to be associated with each state, with bands of color indicating the mean and standard deviation about the mean force characterizing each state.

Table~\ref{table:p5ab-bhmm-confidence-intervals} lists the BHMM posterior means and confidence intervals characterizing the three-state model extracted from this single 60 s observed force trace.
Several things are notable about the estimated model parameters.
Surprisingly, while there is a clearly-resolved intermediate-force state (\emph{state 2}) through which most of the flux from the high- and low-force states passes (as seen from large $K_{12}$ and $K_{23}$), there are nontrivial rate constants connecting the high and low force states directly ($K_{13}$), indicating that while a sequential mechanism involving passing through the intermediate state is preferred, it may not be an obligatory step in hairpin formation under these conditions.
While the state mean forces are clearly distinct, the state standard deviations---which reflect the width of the observed force distribution characterizing each state, rather than the uncertainty in state means---possess overlapping confidence intervals.
These standard deviations reflect not only contributions from both the distribution of extensions sampled by the hairpin in each conformational state, but also from fluctuations in the handles and beads, and other sources of mechanical and electrical noise in the measurement.
The overlapping confidence intervals in these standard deviations suggest there is no significant change in the overall stiffness of the tether upon unfolding.
%As we would expect the unfolded hairpin to be more compliant (i.e.~possess a wider distribution of forces) than the folded hairpin, the inability to distinguish the standard deviations among states is suggestive that, for this experimental configuration and observation time, the predominant contribution to the observed force distributions for each state may be in the form of handle or bead fluctuations or other sources of measurement noise.

Finally, the lifetime of the intermediate-force state is significantly shorter than for the low- and high-force states by nearly an order of magnitude, and only a few times longer than the observation interval of 1 ms---despite this, the lifetime appears to be well-determined, as indicated by the narrow confidence intervals.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DISCUSSION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{section:discussion}

We have described an approach to determining the first-order kinetic parameters and observable (force or extension) distributions characterizing conformational states in single-molecule force spectroscopy.
By use of a Bayesian extension of hidden Markov models, we are able to characterize the experimental uncertainty in these parameters due to instrument noise and finite-size datasets.
The use of a detailed balance constraint additionally helps reduce the experimental uncertainty over standard hidden Markov models, as both transitions into and out of conformational states provide valuable information about state kinetics and populations in data-poor conditions~\cite{noe:jcp:2008:transition-matrix-sampling,metzner-noe-schuette:pre:2009:transition-matrix-sampling}.
Additionally, the Gibbs sampling framework used to sample from the Bayesian posterior can be easily extended to incorporate additional nuisance parameters, such as stochastic models of {\color{red}measurement noise} or laser power fluctuations.

\color{red}
Drift has historically been a challenging problem for force spectroscopy measurements, with experiments using previous-generation instruments resorting to short force-ramp experiments to minimize the cumulative effects of drift in recorded traces~\cite{collin:nature:2005:crooks-verification}.
If the drift only affects the measured signal (e.g.~force or bead position), it is possible to correct for this drift during the analysis procedure using various approaches, such as the introduction of nuisance parameters characterizing linear drift (along with associated priors), or heuristic corrections applied directly to the observed data~\cite{moghtaden:comp-stat-data-anal:2012:empirical-mode-decomposition}.
However, compensating for significant drift in the physical trap-pipette or trap-trap distance in analysis is difficult, as changes in the relative optical trap position will change the external potential felt by the molecule, and hence alter the kinetics and thermodynamics.
Without imposing a model of how those change with this distance, as well as a physical or empirical model for the dynamics of drift, it is impossible to include these effects during analysis.

Fortunately, obtaining many drift-free measurement traces of the duration presented here with current-generation miniaturized optical traps that employ a pipette to suction one of the beads in the tether (such as the one employed here~\cite{bustamante-smith:2006:minitweezers-patent}), the collection of data sufficiently free of drift is regularly achievable~\cite{bustamante:csh-protocols:2009:optical-tweezers}.
Moreover, drift is even less problematic for newer generations of instrument that utilize two optical traps instead of a pipette~\cite{block:nature:2003:dual-trap,bustamante:annu-rev-biochem:2008:optical-tweezers}.
\color{black}

We have opted to make use of a reversible transition matrix to describe the statistical kinetic behavior between the observation intervals $\Delta t$, but we note that it is possible to use a reversible rate matrix instead by substituting a rate matrix sampling scheme~\cite{hummer:njp:2005:rate-matrix-sampling} in the appropriate stage of the Gibbs sampling updates.
\color{red}
Because the use of a rate matrix instead of a transition matrix involves a parametric change of variables, an appropriate prior must be selected, or a Jacobian required to obtain the same posterior distribution as we obtain sampling the reversible transition matrix.
\color{black}

While the experimenter must currently choose the number of conformational states by hand, a number of extensions of Bayesian hidden Markov models can be used to automatically determine the number of states best supported by the data, including reversible-jump schemes~\cite{robert:j-r-statist-soc-b:2000:reversible-jump-mcmc,degunst-schouten:bernoulli:2003:reversible-jump-hmm} and variational Bayes methods~\cite{beal:variational-bayes,bronson:biophys-j:2009:variational-bayesian-hmm}.

We note that the experimenter in principle has access to the full posterior distribution of models given the observed data, so that instead of looking at the confidence of single parameters, confidence intervals in more complex functions of parameters---such as the rates or lifetimes in Table~\ref{table:p5ab-bhmm-confidence-intervals}---can be computed, or joint posterior distributions of multiple parameters examined.
It is also possible to generate synthetic data from the current model, or family of models, to examine how the collection of additional data will further reduce uncertainties or allow discrimination among particular hypotheses.
The field of Bayesian experimental design~\cite{chaloner:stat-sci:1995:bayesian-experimental-design-review} holds numerous possibilities for selecting how future experiments can maximize information gain, and whether the information gain from the collection of additional data will be of sufficient utility to justify the expense.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ACKNOWLEDGMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Acknowledgments}
The authors thank Sergio Bacallado (Stanford University) {\color{red}and anonymous referees} for helpful feedback on an earlier version of this manuscript, and Steve Presse (UCSF) for engaging discussions on this topic.
\color{red}
JDC acknowledges support from a QB3-Berkeley Distinguished Postdoctoral Fellowship and the Memorial Sloan Kettering Cancer Center.
BGK gratefully acknowledges the support by the German Research Foundation (DFG) within the collaborative research center SFB 765 (associated project).
\color{black}
FN acknowledges DFG Grant 825/2.
This work was supported in part by a grant from the NSF (SM).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% BIBLIOGRAPHY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliography{bhmm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ALGORITHMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Appendix: Algorithms}
\label{section:algorithms}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Generating an initial model}

To initialize either computation of the MLHMM or sampling from the posterior for the BHMM, an initial model that respects any constraints imposed in the model prior $P(\bfm{\Theta})$ must be selected.
Here, we employ a Gaussian observable distribution model for $\varphi(o \mid \bfm{e})$ (Eq.~\ref{equation:gaussian-observable}) and enforce that the transition matrix $\bfm{T}$ satisfy detailed balance.

{\color{blue}[JDC: We need to revise this section to discuss the new randomized multi-start initialization.]}

%\footnote{Physical systems that are not driven in time by an external force or fed by an energy reservoir should satisfy detailed balance~\cite{van-kampen}, and its use has been shown to provide a large reduction in transition matrix uncertainty in data-poor conditions~\cite{noe:jcp:2008:transition-matrix-sampling,metzner-noe-schuette:pre:2009:transition-matrix-sampling}. Detailed balance specifies that $\pi_i T_{ij} = \pi_j T_{ji}$ for all $i,j$, where $\bfv{\pi}$ is the equilibrium distribution of the row-stochastic transition matrix $\bfm{T}$.}.

\subsubsection{Observable parameter estimation}

We first initialize the observed distributions of each state by fitting a Gaussian mixture model with $M$ states to the pooled observed data $\bfm{O}$, ignoring temporal information:
\begin{eqnarray}
P(\bfm{O} \mid \bfv{\pi}, \bfm{E}) &=& \prod_{t=0}^{L^{}} \sum_{m=1}^M \pi_m \varphi(o^{}_{t} \mid \mu_m, \sigma^2_m) ,
\label{equation:gaussian-mixture-likelihood}
\end{eqnarray}
where the state observable emission probability vector $\bfm{E} \equiv \{\bfm{e}_1,\ldots,\bfm{e}_M\}$ and $\bfm{e}_m \equiv \{\mu_m, \sigma_m^2\}$ with $\mu_m$ denoting the observable mean and $\sigma_m^2$ the variance for state $m$ for the Gaussian mixture model.
The vector $\bfv{\pi}$ is composed of equilibrium state populations $\{ \pi_1, \ldots , \pi_M\}$ with $\pi_m \ge 0$ and $\sum_{m=1}^M \pi_m = 1$.

A first approximation to $\bfv{\pi}$ and $\bfm{E}$ is computed by pooling and sorting the observed $o^{}_{t}$, and defining $M$ indicator functions $h_m(o)$ that separate the data into $M$ contiguous regions of the observed range of $o$ of roughly equal population.
Let $N_m \equiv \sum_{t=0}^{L^{}} h_m(o^{}_{t})$ denote the total number of observations falling in region $m$, and $N_\mathrm{tot} = \sum_{m=1}^M N_m$.
The initial parameters are then computed as,
\begin{eqnarray}
\pi_m &=& N_m / N_\mathrm{tot} \nonumber \\
\mu_m &=& N_m^{-1} \sum_{t=0}^{L^{}} o^{}_{t} \, h_m(o^{}_{t}) \\
\sigma^2_m &=& N_m^{-1} \sum_{t=0}^{L^{}} (o^{}_{t} - \mu_m)^2 \, h_m(o^{}_{t}) .
\end{eqnarray}

{\color{red} To try to maximize the likelihood in Eq.~\ref{equation:gaussian-mixture-likelihood}, this} approximation is then improved upon by iterating the expectation-maximization procedure described by Bilmes~\cite{bilmes:1998:expectation-maximization},
\begin{eqnarray}
\pi'_m &=& N_\mathrm{tot}^{-1} \sum_{t=0}^{L^{}} \chi_m(o^{}_t, \bfm{E}, \bfv{\pi}) \nonumber \\
\mu'_m &=& (\pi'_m N_\mathrm{tot})^{-1} \sum\limits_{t=0}^{L^{}} o^{}_{t} \, \chi_m(o^{}_{t}, \bfm{E}, \bfv{\pi}) \nonumber \\
{\sigma'}^2_m &=& (\pi'_m N_\mathrm{tot})^{-1} \sum\limits_{t=0}^{L^{}} (o^{}_{t} - \mu'_m)^2 \, \chi_m(o^{}_{t}, \bfm{E}, \bfv{\pi}) 
\end{eqnarray}
where the function $\chi_m(o, \bfm{E}, \bfv{\pi})$ is given by the fuzzy membership function,
\begin{eqnarray}
\chi_m(o, \bfm{E}, \bfv{\pi}) &=& \frac{\pi_m \, \varphi(o \mid \bfm{e}_m)}{\sum\limits_{l=1}^M \pi_l \, \varphi(o \mid \bfm{e}_l)} .
\end{eqnarray}
The iterative procedure is terminated at iteration $j$ when the change in the parameters $\{\bfv{\pi}, \bfv{\mu}, \bfv{\sigma}^2\}$ falls below a certain relative threshold, such as $\|\bfv{\pi}^{[j]} - \bfv{\pi}^{[j-1]}\|_2 / \|\bfv{\pi}^{[j]}\|_2 < 10^{-4}$ {\color{red}, where $\| \ldots \|_2$ denotes the euclidean L2 norm}.

\subsubsection{Transition matrix estimation}

Once initial state observable emission parameters $\bfm{E}$ are determined, an initial transition matrix is estimated using an iterative likelihood maximization approach that enforces detailed balance~\cite{noe:jcp:2011:msm-review}.
First, a matrix of fractional transition counts $\bfm{C} \equiv (c_{ij})$ is estimated using the membership function:
\begin{eqnarray}
c_{ij} &=& \sum_{t=1}^{L^{}} \chi_i(o^{}_{t-1}, \bfm{E}, \bfv{\pi}) \, \chi_j(o^{}_{t}, \bfm{E}, \bfv{\pi})
\end{eqnarray}
A symmetric $M \times M$ matrix $\bfm{X} \equiv (x_{ij})$ is initialized by
\begin{eqnarray}
x_{ij} = x_{ji} &=& c_{ij} + c_{ji} .
\end{eqnarray}
\color{red}
The iterative procedure described in Algorithm 1 of \cite{noe:jcp:2011:msm-review} is then applied to update the transition matrix in a manner that preserves the detailed balance constraint while maximizing the likelihood function $P(\bfm{T} | \bfm{S})$.
This algorithm proceeds as follows.
\color{black}

For each update iteration, we first update the diagonal elements of $\bfm{X}$:
\begin{eqnarray}
x'_{ii} = \frac{c_{ii} (x_{i*} - x_{ii})}{c_{i*} - c_{ii}}  \:\:;\;\; c_{i*} = \sum_{j=1}^M c_{ij} \:\:;\:\: x_{i*} = \sum_{j=1}^M x_{ij} ,
\end{eqnarray}
followed by the off-diagonal elements:
\begin{eqnarray}
x'_{ij} = x'_{ji} &=& \frac{-b + \sqrt{b^2 - 4 a c}}{2 a}
\end{eqnarray}
where the quantities $a$, $b$, and $c$ are computed from $\bfm{X}$ and $\bfm{C}$,
\begin{eqnarray}
a &\equiv& c_{i*} - c_{ij} + c_{j*} - c_{ji} \nonumber \\
b &\equiv& c_{i*} (x_{j*} - x_{ji}) + c_{j*} (x_{i*} - x_{ij}) \nonumber \\
&&\mbox{}- (c_{ij} + c_{ji}) (x_{i*} - x_{ij} + x_{j*} - x_{ji}) \nonumber \\
c &\equiv& - (c_{ij} + c_{ji}) (x_{i*} - x_{ij}) (x_{j*} - x_{ji}) .
\end{eqnarray}
Once a sufficient number of iterations $j$ have been completed to compute a stable estimate of $\bfm{X}$ (such as the relative convergence criteria $\|\bfm{X}^{[j]} - \bfm{X}^{[j-1]}\|_2 / \|\bfm{X}^{[j]}\|_2 < 10^{-4}$), the maximum likelihood transition matrix estimate $\bfm{T}$ is computed as
\begin{eqnarray}
%T_{ij} &=& \frac{x_{ij}}{x_{i*}} . 
T_{ij} &=& x_{ij} / x_{i*} . 
\end{eqnarray}
Note that the equilibrium probability vector $\bfv{\pi}$ computed during the Gaussian mixture model fitting is not respected during this step.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fitting a maximum likelihood HMM}
\label{section:fitting-mlhmm}

The HMM parameters $\bfm{\Theta} \equiv \{ \bfm{T}, \bfm{E} \}$ are fit to the observed data $\bfm{O}$ through use of the expectation-maximization (EM) algorithm~\cite{dempster-laird-rubin:1977:j-royal-statist-soc-b:em-algorithm}.
This is an iterative procedure, where the model parameters are subsequently refined through successive iterations.
The initial HMM is usually quick to compute, and can give the experimenter a rough idea of the model parameters, as well as providing a useful starting point for sampling models from the Bayesian posterior.

During each iteration, the Baum-Welch algorithm~\cite{baum:1970:ann-math-statist:baum-welch} is used to compute $\bfm{\Xi}^{} \equiv (\xi^{}_{tij})$, which represents the probability that the system transitions from hidden state $i$ at time $t-1$ to hidden state $j$ at time $t$, and $\gamma^{}_{ti}$, the probability that the system occupied state $i$ at time $t$.
This is accomplished by first executing the forward algorithm,
\begin{eqnarray}
\alpha_{tj} &=& \begin{cases}
\rho_j \, \varphi(o_0 \mid \bfm{e}_j) & t = 0 \\
\varphi(o_t \mid \bfm{e}_j) \sum_{i=1}^M \alpha_{(t-1)i} T_{ij} & t = 1,\ldots,L
\end{cases}
\end{eqnarray}
followed by the backward algorithm,
\begin{eqnarray}
\beta_{ti} &=& \begin{cases}
1 & t = L \\
\sum_{j=1}^M T_{ij} \varphi(o_{t+1} \mid \bfm{e}_j) \beta_{(t+1)j} & t = (L-1),\ldots,0
\end{cases}
\end{eqnarray}
The $L \times M \times M$ matrix $\bfm{\Xi}$ is then computed for $t = 0,\ldots,(L-1)$ as,
\begin{eqnarray}
\xi_{tij} &=& \alpha_{ti} \varphi(o_{t+1} \mid \bfm{e}_i) T_{ij} \beta_{(t+1)j} / \sum_{i=1}^M \alpha_{Ti} \\
\gamma_{ti} &=& \sum_{j=1}^M \xi_{tij}
\end{eqnarray}
In practice, the logarithms of these quantities are computed instead to avoid numerical underflow.

The aggregate matrix of expected transition counts $\bfm{C}~\equiv~(c_{ij})$ is then computed from $\bfm{\Xi}^{}$ as,
\begin{eqnarray}
c_{ij} &=& \sum_{t=0}^{L^{}-1} \xi_{tij}^{} .
\end{eqnarray}
This count matrix is used to update the maximum-likelihood transition matrix $\bfm{T}$ using the method of Prinz et al.~\cite{noe:jcp:2011:msm-review} described in the previous section.

The state observable distribution parameters $\bfm{E}$ are then updated from the $\gamma_{ti}$.
For the univariate normal distribution applied to force spectroscopy data here, we update the mean $\mu_i$ and variance $\sigma^2_i$ for state $i$ using the scheme,
\begin{eqnarray}
\mu'_i = \frac{\sum\limits_{t=0}^{L} o^{}_{t} \gamma^{}_{ti}}{\sum\limits_{t=0}^{L^{}} \gamma^{}_{ti}} \:\:;\:\: {\sigma'}^2_i = \frac{\sum\limits_{t=0}^{L^{}} (o^{}_{t} - \mu'_i)^2 \gamma^{}_{ti}}{\sum\limits_{t=0}^{L^{}} \gamma^{}_{ti}} .
\end{eqnarray}

Once the model parameters have been fitted by iteration of the above update procedure to convergence (which may only converge to a local maximum of the likelihood), the most likely hidden state sequence can be determined given the observations $\bfm{O}$ and the MLE model $\hat{\bfm{\Theta}}$ using the Viterbi algorithm~\cite{viterbi:1967:IEEE-trans-info-theory:viterbi-algorithm}.
Like the forward-backward algorithm employed in the Baum-Welch procedure, the Viterbi algorithm also has a forward recursion component,
\begin{eqnarray}
\epsilon_{jt} &=& \begin{cases}
\rho_j \varphi(o_t \mid \bfm{e}_j) & t = 0 \\
\varphi(o_t \mid \bfm{e}_j) \max_{i} \epsilon_{i(t-1)} T_{ij}  & t = 1,\ldots,L
\end{cases} \label{equation:viterbi-forward} \\
\Phi_{jt} &=& \begin{cases}
1 & t = 0 \\
\arg\max_{i} \epsilon_{i(t-1)} T_{ij}  & t = 1,\ldots,L
\end{cases} \nonumber
\end{eqnarray}
as well as a reverse reconstruction component to compute the most likely state sequence $\hat{\bfm{S}}$,
\begin{eqnarray}
\hat{s}_t &=& \begin{cases}
\arg\max_i \epsilon_{it} & t = L\\
\Phi_{\hat{s}_{t+1}(t+1)} & t = (L-1),\ldots,0
\end{cases} 
\end{eqnarray}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sampling from the posterior of the BHMM}
\label{section:algorithms:bhmm-posterior-sampling}

Sampling from the posterior of the BHMM (Eq.~\ref{equation:full-posterior}) proceeds by rounds of Gibbs sampling, where each round consists of an update of the augmented model parameters $\{\bfm{T}, \bfm{E}, \bfm{S}\}$ by sampling
\begin{eqnarray}
\begin{array}{lcl}
\bfm{S}' \mid \bfm{T}, \bfm{E}, \bfm{O} &\sim& P(\bfm{S}' \mid \bfm{T}, \bfm{E}, \bfm{O}) \nonumber \\
\bfm{T}' \mid \bfm{S}' &\sim& P(\bfm{T}' \mid \bfm{S}') \nonumber \\
\bfm{E}' \mid \bfm{S}', \bfm{O} &\sim& P(\bfm{E}' \mid \bfm{S}', \bfm{O}) 
\end{array}
\end{eqnarray}
where the conditional probabilities are given by Eq.~\ref{equation:conditional-gibbs}.

%%%%%%%%%%%
\subsubsection{Updating the hidden state sequences}

We use a modified form of the Viterbi process to generate an independent sample of the hidden state history $\bfm{S}$ given the transition probabilities $\bfm{T}$, state observable distribution parameters $\bfm{E}$, and observed data $\bfm{O}$.
Like the Viterbi scheme, a forward recursion is applied to each observation trace $\bfm{o}^{}$, but instead of computing the most likely state history on the reverse pass, a new hidden state history $\bfm{S}$ is drawn from the distribution $P(\bfm{S} \mid \bfm{O}, \bfm{T}, \bfm{E})$.
The forward recursion uses the same forward algorithm as used in Baum-Welch~\cite{baum:1970:ann-math-statist:baum-welch}, 
\begin{eqnarray}
\alpha_{tj} &=& \begin{cases}
\rho_j \, \varphi(o_0 \mid \bfm{e}_j) & t = 0 \\
\varphi(o_t \mid \bfm{e}_j) \sum_{i=1}^M \alpha_{(t-1)i} T_{ij} & t = 1,\ldots,L
\end{cases}
\end{eqnarray}
In the reverse recursion, we now sample a state sequence by sampling each hidden state from the conditional distribution $s_t \sim P(s_t \mid s_{t+1}, \ldots, s_L)$ starting from $t = L$ and proceeding down to $t = 0$, where the conditional distribution is given by,
\begin{eqnarray}
\lefteqn{P(s_t = i \mid s_{t+1}, \ldots, s_L)} \\
&\propto& \begin{cases}
\alpha_{t i} / \sum_{j=1}^M \alpha_{t j} & t = L \\
\alpha_{t i} T_{i s_{t+1}} / \sum_{j=1}^M \alpha_{t j} T_{j s_{t+1}} & t = (L-1),\ldots,0
\end{cases} \nonumber
\end{eqnarray} 
It is straightforward to show the result of these sampling steps reconstitutes the probability distribution $P(\bfm{S} | \bfm{T}, \bfm{E}, \bfm{O})$ (see \emph{Supplementary Material: Proof of state history sampling scheme}).

%%%%%%%%%%%
\subsubsection{Updating the transition probabilities}

%If no detailed balance constraint is used and the prior $P(\bfm{T})$ is Dirichlet in each row of the transition matrix $\bfm{T}$, it is possible to generate an independent sample of the transition matrix from the conditional distribution $P(\bfm{T}' \mid \bfm{S}')$ by sampling each row of the transition matrix from the conjugate Dirichlet posterior using the transition counts from the sampled state sequence $\bfm{S}'$~\cite{noe:jcp:2008:transition-matrix-sampling}.
{\color{red} If we assume each row of the transition matrix $\bfm{T}$ is independent with Dirichlet prior $P(\bfm{T})$, then it is possible to sample from the transition matrix posterior Dirichlet distribution $P(\bfm{T}' \mid \bfm{S}')$, given the transition counts calculated from the sampled state sequence $\bfm{S}'$~\cite{noe:jcp:2008:transition-matrix-sampling}.}
However, because physical systems in the absence of energy input through an external driving force should satisfy detailed balance, we make use of this constraint in updating our transition probabilities, since this has been demonstrated to substantially reduce parameter uncertainty in the data-limited regime~\cite{noe:jcp:2008:transition-matrix-sampling}.

The transition matrix is updated using the reversible transition matrix sampling scheme of No\'{e}~\cite{noe:jcp:2008:transition-matrix-sampling,chodera-noe:jcp:2011:observable-sampling}.
Here, an adjusted count matrix $\bfm{C} \equiv (c_{ij})$ is computed using the updated hidden state sequence $\bfm{S}'$,
\begin{eqnarray}
c_{ij} &=& b_{ij} +\sum_{t=1}^{L} \delta_{is_{t-1}} \delta_{js_{t}} ,
\end{eqnarray}
where the Kronecker $\delta_{ij} = 1$ if $i = j$ and zero otherwise, and $\bfm{B} \equiv (b_{ij})$ is a matrix of prior pseudocounts, which we take to be zero following the work of No\'{e} et al.~\cite{noe:pnas:2009:ww-domain}.
Using the adjusted count matrix $\bfm{C}$, a Metropolis-Hastings Monte Carlo procedure~\cite{metropolis-hastings} is used to update the matrix and produce a new sample from $P(\bfm{T}' \mid \bfm{S}')$.
Two move types are attempted, selected with equal probability, and 1000 moves are attempted to generate a new sample $\bfm{T}'$ that is approximately uncorrelated from the previous $\bfm{T}$.
Prior to starting the Monte Carlo procedure, the vector of equilibrium probabilities for all states $\bfv{\pi}$ is computed according to 
\begin{eqnarray}
\bfm{T}^\T \bfv{\pi} = \bfv{\pi} .
\end{eqnarray}

The first move type is a reversible element shift.
A pair of states $(i,j)$, $i \ne j$, are selected with uniform probability, and a random number $\Delta$ is selected uniformly over the interval,
\begin{equation*}
\Delta \in[\max(-T_{ii},-\frac{\pi_{j}}{\pi_{i}}T_{jj}),T_{ij}] .
\end{equation*}
The changed elements in the proposed transition matrix $\bfm{T}'$ are then given by:
\begin{align*}
T'_{ij} & =T_{ij}-\Delta \:\: ; \:\: T'_{ji}  =T_{ji}-\frac{\pi_{i}}{\pi_{j}}\Delta \\
T'_{ii} & =T_{ii}+\Delta \:\: ; \:\: T'_{jj}  =T_{jj}+\frac{\pi_{i}}{\pi_{j}}\Delta  .
\end{align*}
This move is accepted with probability
\begin{eqnarray}
\lefteqn{P_\mathrm{accept}(\bfm{T}' | \bfm{T} ) = \mathrm{min}\left\{1, \sqrt{\frac{(T'_{ij})^{2}+(T'_{ji})^{2}}{(T_{ij})^{2}+(T_{ji})^{2}}} \right. }\\
&\times& \left. \left(\frac{T'_{ii}}{T_{ii}}\right)^{c_{ii}}\left(\frac{T'_{ij}}{T_{ij}}\right)^{c_{ij}} \left(\frac{T'_{jj}}{T_{jj}}\right)^{c_{jj}}\left(\frac{T'_{ji}}{T_{ji}}\right)^{c_{ji}} \right\} . \nonumber
\end{eqnarray}
This move will leave the vector of stationary probabilities $\bfv{\pi}$ unchanged.

The second move type is a row shift.
A row $i$ of $\bfm{T}$ is selected with uniform probability, and a random number $\eta$ chosen uniformly over the interval 
\begin{equation*}
\eta\in\left[0,\frac{1}{1-T_{ii}}\right]
\end{equation*}
and used to update row $i$ of $\bfm{T}$ according to
\begin{eqnarray}
T'_{ij} &=& \begin{cases}
\eta T_{ij} & j = 1, \ldots, M, \:\: j \ne i\\
\eta (T_{ii} - 1) + 1 & j = i
\end{cases}
\end{eqnarray}
This move is accepted with probability
\begin{eqnarray}
P_\mathrm{accept}(\bfm{T}' | \bfm{T}) &=& \mathrm{min} \left\{1, \eta^{(M-2)} \eta^{(c_{i*}-c_{ii})}\left(\frac{1-\eta(1-T_{ii})}{T_{ii}}\right)^{c_{ii}} \right\} . \nonumber\\
\end{eqnarray}
The row shift operation will change the stationary distribution of $\bfv{\pi}'$, but it may be efficiently updated with
\begin{align*}
\pi'_{i} & =\frac{\pi_{i}}{\pi_{i}+\eta(1-\pi_{i})} \:\: ; \:\: \pi'_{j} =\frac{\eta \, \pi_{j}}{\pi_{i}+\eta(1-\pi_{i})}.
\end{align*}
Since this update scheme is incremental, it will accumulate numerical errors over time that cause the updated $\bfv{\pi}$ to drift away from the stationary distribution of the current transition matrix. 
To avoid this, $\bfv{\pi}$ is recomputed from the current sample of the transition matrix in regular intervals (here, every 100 sampling steps).

%%%%%%%%%%%
\subsubsection{Updating the observable distribution parameters}

Following the update of the transition matrix $\bfm{T}$, the observable distribution parameters $\bfm{E}$ are updated by sampling $\bfm{E}$ from the conditional probability $P(\bfm{E}' \mid \bfm{S}', \bfm{O})$.
The conditional probability for the observable distribution parameters for state $m$, denoted $\bfm{e}_m$, is given in terms of the output model $\varphi(o \mid \bfm{e})$ by Bayes' theorem,
\begin{eqnarray}
P(\bfm{E} \mid \bfm{O}, \bfm{S}) &=& \left[\prod_{t=0}^{L^{}} \varphi(o_{t} \mid \bfm{e}_{s^{}_{t}}) \right] P(\bfm{E}) \label{equation:conditional-emission-probability} .
\end{eqnarray}

An important choice must be made with regards to the prior, $P(\bfm{E})$.
If the prior is chosen to be composed of independent priors for each state, as in
\begin{eqnarray}
P(\bfm{E}) &=& \prod_{m=1}^M P(\bfm{e}_m) \label{equation:separable-emission-prior} ,
\end{eqnarray}
then the full BHMM posterior (Eq.~\ref{equation:full-posterior}) will be invariant under any permutation of the states.
This behavior might be undesirable, as the states may switch labels during the posterior sampling procedure; this will require any analysis of the models sampled from the posterior to account for the possible permutation symmetry in the states.
On the other hand, breaking this symmetry (e.g., by enforcing an ordering on the state mean observables) can artificially restrict the confidence intervals of the states, which might additionally complicate data analysis.

Here, we make the choice that the prior be separable (Eq.~\ref{equation:separable-emission-prior}), which has the benefit of allowing the conditional probability for $\bfm{E}$ (Eq.~\ref{equation:conditional-emission-probability}) to be decomposed into a separate posterior for each state.
For each state $m$, collect all the observations $o^{}_{t}$ whose updated hidden state labels ${s^{}_{t}}' = m$ into a single dataset $\bfm{o} \equiv \{o_n\}_{n=1}^{N_m}$, where $N_m$ is the total number of times state $m$ is visited, for the purposes of this update procedure.
Then, the observable parameters $\bfm{e}$ for this state are given by
\begin{eqnarray}
P(\bfm{e} \mid \bfm{o}) &=& P(\bfm{o} \mid \bfm{e}) P(\bfm{e}) = \left[ \prod_{n=1}^{N_m} \varphi(o_n \mid \bfm{e}) \right] P(\bfm{e}) .
\end{eqnarray}

In the application presented here, we use a Gaussian output model (Eq.~\ref{equation:gaussian-observable}) 
for the state observable distributions $P(o \mid \bfm{e})$, where $\bfm{e} \equiv \{\mu, \sigma^2\}$, with $\mu$ the state mean observable and $\sigma^2$ the variance (which will include both the distribution of the observable characterizing the state and any broadening from measurement noise).
Other models (including multidimensional or multimodal observation models) are possible, and require replacing only the observation model $\varphi(o \mid \bfm{e})$ and corresponding prior $P(\bfm{e})$.

We use the (improper) Jeffreys prior~\cite{jeffreys:proc-royal-soc-a:1946:jeffreys-prior} which has the information-theoretic interpretation as the prior that maximizes the information content of the data~\cite{goyal:bayesian-inference-book:2005:jeffreys-prior}, (suppressing the state index subscript $m$),
\begin{eqnarray}
P(\bfm{e}) &\propto& \sigma^{-1} \label{equation:emission-parameter-prior} ,
\end{eqnarray}
which produces the posterior
\begin{eqnarray}
P(\bfm{e} \mid \bfm{o}) &\propto& \sigma^{-(N+1)} \exp\left[-\frac{1}{2 \sigma^2} \sum_{n=1}^N (o_n - \mu)^2 \right] ,
\end{eqnarray}
where we remind the reader that here and in the remainder of this section, the symbols $\bfm{e}$, $\bfm{o}$, $\sigma$, $\mu$, and $N$ refer to $\bfm{e}_m$, $\bfm{o}_m$, $\sigma_m$, $\mu_m$, and $N_m$, respectively.

Updating $\{\mu, \sigma^2\}$ also proceeds by a Gibbs sampling scheme, alternately updating $\mu$ and $\sigma$, as earlier described in Ref.~\citep{chodera-noe:jcp:2011:observable-sampling},
\begin{eqnarray}
\mu &\sim& P(\mu \mid \sigma^2, \bfm{o}) \nonumber \\
\sigma^2 &\sim& P(\sigma^2 \mid \mu, \bfm{o})
\end{eqnarray}

The conditional distribution of the mean $\mu$ is then given by
\begin{eqnarray}
%P(\mu \mid \sigma^2, \bfm{o}) &\equiv& \frac{P(\mu, \sigma^2 \mid \bfm{o})}{\int d\mu \, P(\mu, \sigma^2 \mid \bfm{o})} \nonumber \\
%&\propto& \exp\left[-\frac{1}{2 (\sigma^2/N)} (\mu - \hat{\mu})^2 \right] 
P(\mu \mid \sigma^2, \bfm{o}) &\propto& \exp\left[-\frac{1}{2 (\sigma^2/N)} (\mu - \hat{\mu})^2 \right] 
\end{eqnarray}
where $\hat{\mu}$ is the sample mean for $\bfm{o}$, the samples in state $m$,
\begin{eqnarray}
\hat{\mu} &\equiv& \frac{1}{N} \sum_{n=1}^N o_n
\end{eqnarray}
This allows us to update $\mu$ according to 
\begin{eqnarray}
\mu' &\sim& \mathcal{N}(\hat{\mu}, \sigma^2 / N)
\end{eqnarray}

The conditional distribution of the variance $\sigma^2$ is given by
\begin{eqnarray}
%P(\sigma^2 \mid \mu, \bfm{o}) &=& \frac{p(\mu, \sigma^2 \mid \bfm{o})}{\int d\sigma^2 \, p(\mu, \sigma^2 \mid \bfm{o})} \nonumber \\
%&\propto& \sigma^{-(N+1)} \exp\left[-\frac{1}{2 \sigma^2} \sum_{n=1}^N (o_n - \mu)^2 \right] \nonumber \\
%&\propto& \sigma^{-(N+1)} \exp\left[-\frac{N \hat{\sigma}^2}{2 \sigma^2} \right] 
P(\sigma^2 \mid \mu, \bfm{o}) &\propto& \sigma^{-(N+1)} \exp\left[-\frac{N \hat{\sigma}^2}{2 \sigma^2} \right] 
\end{eqnarray}
where the quantity $\hat{\sigma}^2$, which is not in general identical to the sample variance, is given by
\begin{eqnarray}
\hat{\sigma}^2 &\equiv& \frac{1}{N} \sum_{n=1}^N (o_n - \mu)^2 .
\end{eqnarray}
A convenient way to update $\sigma^2 \mid \mu, \bfm{o}$ is to sample a random variate $y$ from the chi-square distribution with $N-1$ degrees of freedom,
\begin{eqnarray}
y &\sim& \chi^2(N-1)
\end{eqnarray}
and then update $\sigma^2$ as
\begin{eqnarray}
%{\sigma'}^2 &=& \frac{N \hat{\sigma}^2}{y} .
{\sigma'}^2 &=& N \hat{\sigma}^2 / y .
\end{eqnarray}
Note that $\mu$ and $\sigma^2$ can be updated in either order, but the updated values of $\mu$ or $\sigma^2$ must be used in sampling the not-yet-updated $\sigma^2$ or $\mu$, and vice-versa.

Other output probabilities, such as mixtures of normal distributions or other distributions, can be substituted by simply changing $P(\bfm{E} \mid \bfm{O}, \bfm{S})$ and the scheme by which $\bfm{E}$ is updated.

\end{document}

### 2015-02-16 manuscripts/biophysical-journal-cover-letter/cover-letter.tex
Initial import from http://simtk.org/home/bhmm Adding everything but experimental datasets.

\documentclass[ucb,qb3,10pt,fullfrom]{ucletter}

\usepackage{palatino}

\name{John D. Chodera}
\telephone{415.867.7384}
\email{jchodera@berkeley.edu}
%\department{Department of Chemistry}
%\fromaddress{QB3 Institute}
%260J Stanley Hall #3220\\
%University of California\\
%Berkeley, CA 94720-3220}
%\mailcode{5447}
\date{\today}

%\documentclass[10pt]{letter}
%\address{
%John D. Chodera\\
%QB3 Institute\\
%260J Stanley Hall \#3220\\
%University of California\\
%Berkeley, CA 94720-3220, USA \\
%phone: 415.867.7384\\
%email: \href{mailto:jchodera@berkeley.edu}{jchodera@berkeley.edu}\\
%\name{John D. Chodera}
%}

\usepackage{fancyhdr}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}

\def\bs{$\backslash$}

\pagestyle{fancy}
\fancyhf{}
\rfoot{\small\textsc{John D. Chodera - Cover Letter - \thepage}}

\signature{
\resizebox{2.5in}{!}{\includegraphics{JDC-signature}} \\
John D.~Chodera,\\corresponding author
}

\usepackage{times}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}

% We will be using PDF figures and generating PDF output.
\pdfoutput=1
\pdfcompresslevel=9
\usepackage[pdftex]{graphicx}
\DeclareGraphicsExtensions{.pdf}

%        \addtolength{\oddsidemargin}{-.875in}
%        \addtolength{\evensidemargin}{-.875in}
%        \addtolength{\textwidth}{1.75in}
%
%        \addtolength{\topmargin}{-.875in}
%        \addtolength{\textheight}{1.75in}
%
%\begin{document}
%
%\begin{letter}{To:\\MFPL Faculty Search Committee}
%
%\opening{Dear Search Committee,}

\begin{document}
%\small
\begin{letter}{}
\opening{Dear \emph{Biophysical Journal} Editors,}

We are submitting the following manuscript for consideration for publication as an {\bf Article} in \emph{Biophysical Journal}:

{\bf Bayesian hidden Markov model analysis of single-molecule force spectroscopy: Characterizing kinetics under measurement uncertainty}, by John D.~Chodera, Phillip Elms, Frank No\'{e}, Bettina Keller, Christian M.~Kaiser, Aaron Ewall-Wice,  Susan Marqusee, Carlos Bustamante, and Nina Singhal Hinrichs.

In this manuscript, we describe a new technique for the analysis of single-molecule force spectroscopy experiments.
Our approach utilizes a \emph{Bayesian} extension of hidden Markov models that allows information about kinetic and equilibrium properties to be extracted simultaneously from force or extension data collected in single-molecule optical or atomic force microscopy experiments.
In addition to providing a way to reconstruct hidden state sequences (like standard hidden Markov models), our approach provides much more accurate estimates of characteristics of conformational states and transition rates than previous methods based on simple segmentation of the observed force range using force thresholds.
Our approach also provides an excellent assessment of the \emph{confidence intervals} with which various kinetic and mechanical properties are known can be computed, allowing the experimenter to easily judge whether a particular mechanistic hypothesis is borne out by the data.
We validate the method on synthetic data, and illustrate the approach applied to real experimental data by characterizing the folding kinetics under force of an RNA hairpin with three distinct conformational states using an optical trap.
All of the Matlab code and data used here will be made freely available online as a companion to this work.

We believe this work will be of high interest to the \emph{Biophysical Journal} readership, as the \emph{Journal} has featured a number of papers describing important advancements in  practical statistical analysis of single-molecule data.
%A number of articles on single-molecule force spectroscopy have appeared in previous issues---a testament to the interest in these tools:
%\begin{itemize}
%  \item Evanko D. \emph{Optimizing your optical tweezers.} Nature Methods 3:584, 2006.
%  \item Dufr\^{e}ne YF. \emph{Atomic force microscopy of membrane proteins separating two aqueous compartments.} Nature Methods 3:973, 2006.
%  \item Walter NG, Huang C-Y, Manzo AJ, and Sobhy MA. \emph{Do-it-yourself guide: how to use the modern single-molecule toolkit.} Nature Methods 5:475, 2008.
%  \item Neuman KC and Nagy A. \emph{Single-molecule force spectroscopy: optical tweezers, magnetic tweezers and atomic force microscopy.} Nature Methods 5:491, 2008.
%  \item Min TL, Mears PJ, Chubiz LM, Rao CV, Golding I, and Chemla YR. \emph{High-resolution, long-term characterization of bacterial motility using optical tweezers.} Nature Methods 6:831, 2009.
%  \item Lipfert J, Kerssemakers JWJ, Jager T, and Deller NH. \emph{Magnetic torque tweezers: measuring torsional stiffness in DNA and RecA-DNA filaments.} Nature Methods: 7:977, 2010.
%  \item Dufr\^{e}ne YF, Evans E, Engel A, Helenius J, Gaub HE, and M\"{u}ller DJ. \emph{Five challenges to bringing single-molecule force spectroscopy into living cells.} Nature Methods 8:123, 2011.
%\end{itemize}

%We believe the primary audience for our manuscript will be similar to that of several recent \emph{Nature Methods} papers:
%\begin{itemize}
%  \item Ballard AJ and Jarzynski C. {\it Replica exchange with nonequilibrium switches.} PNAS 106:12224, 2009.
%  \item Ytreberg FM and Zuckerman DM. {\it A black-box re-weighting analysis can correct flawed simulation data.} PNAS 105:7982, 2008.  
%  \item Frenkel D. {\it Speed-up of Monte Carlo simulations by sampling rejected states.} PNAS 101:17571, 2004.  
%\end{itemize}

%\eject

Appropriate reviewers of this work include the following scientists working in the field of single-molecule force spectroscopy:
\begin{itemize}
  \item Yann R.~Chemla (UIUC) - \url{ychemla@illinois.edu}
  \item Matthias Rief (TU Munich) - \url{mrief@ph.tum.de}
  \item Gerhard Hummer (NIH) - \url{gerhard.hummer@nih.gov}
  \item Steven M Block (Stanford) - \url{sblock@stanford.edu}
  \item Felix Ritort (University of Barcelona) - \url{ritort@ffn.ub.es}  
  \item Michael Woodside (University of Alberta) - \url{michael.woodside@nrc-cnrc.gc.ca}
  \item Robert Best (University of Cambridge) - \url{rbb24@cam.ac.uk}
  \item Jane Clarke (Cambridge) - \url{jc162@cam.ac.uk}
  \item Olga Dudko (UCSD) - \url{dudko@physics.ucsd.edu}
  \item Taekjip Ha (UIUC) - \url{taekjipha@physics.illinois.edu}
  \item Alexander Dunn (Stanford) - \url{alex.dunn@stanford.edu}
\end{itemize}

\closing{Kind regards,}
\end{letter}
\end{document}

### 2015-02-16 manuscripts/nature-methods-cover-letter/cover-letter.tex
Initial import from http://simtk.org/home/bhmm Adding everything but experimental datasets.

--- 2015-02-16 manuscripts/biophysical-journal-cover-letter/cover-letter.tex
@@
 \begin{letter}{}
-\opening{Dear \emph{Biophysical Journal} Editors,}
+\opening{Dear \emph{Nature Methods} Editors,}
 
-We are submitting the following manuscript for consideration for publication as an {\bf Article} in \emph{Biophysical Journal}:
+We are submitting the following manuscript for consideration for publication as an {\bf Article} in \emph{Nature Methods}:
 
@@
 
-We believe this work will be of high interest to the \emph{Biophysical Journal} readership, as the \emph{Journal} has featured a number of papers describing important advancements in  practical statistical analysis of single-molecule data.
-%A number of articles on single-molecule force spectroscopy have appeared in previous issues---a testament to the interest in these tools:
-%\begin{itemize}
-%  \item Evanko D. \emph{Optimizing your optical tweezers.} Nature Methods 3:584, 2006.
-%  \item Dufr\^{e}ne YF. \emph{Atomic force microscopy of membrane proteins separating two aqueous compartments.} Nature Methods 3:973, 2006.
-%  \item Walter NG, Huang C-Y, Manzo AJ, and Sobhy MA. \emph{Do-it-yourself guide: how to use the modern single-molecule toolkit.} Nature Methods 5:475, 2008.
-%  \item Neuman KC and Nagy A. \emph{Single-molecule force spectroscopy: optical tweezers, magnetic tweezers and atomic force microscopy.} Nature Methods 5:491, 2008.
-%  \item Min TL, Mears PJ, Chubiz LM, Rao CV, Golding I, and Chemla YR. \emph{High-resolution, long-term characterization of bacterial motility using optical tweezers.} Nature Methods 6:831, 2009.
-%  \item Lipfert J, Kerssemakers JWJ, Jager T, and Deller NH. \emph{Magnetic torque tweezers: measuring torsional stiffness in DNA and RecA-DNA filaments.} Nature Methods: 7:977, 2010.
-%  \item Dufr\^{e}ne YF, Evans E, Engel A, Helenius J, Gaub HE, and M\"{u}ller DJ. \emph{Five challenges to bringing single-molecule force spectroscopy into living cells.} Nature Methods 8:123, 2011.
-%\end{itemize}
+We believe this work will be of high interest to the \emph{Nature Methods} readership.
+A number of articles on single-molecule force spectroscopy have appeared in previous issues---a testament to the interest in these tools:
+\begin{itemize}
+  \item Evanko D. \emph{Optimizing your optical tweezers.} Nature Methods 3:584, 2006.
+  \item Dufr\^{e}ne YF. \emph{Atomic force microscopy of membrane proteins separating two aqueous compartments.} Nature Methods 3:973, 2006.
+  \item Walter NG, Huang C-Y, Manzo AJ, and Sobhy MA. \emph{Do-it-yourself guide: how to use the modern single-molecule toolkit.} Nature Methods 5:475, 2008.
+  \item Neuman KC and Nagy A. \emph{Single-molecule force spectroscopy: optical tweezers, magnetic tweezers and atomic force microscopy.} Nature Methods 5:491, 2008.
+  \item Min TL, Mears PJ, Chubiz LM, Rao CV, Golding I, and Chemla YR. \emph{High-resolution, long-term characterization of bacterial motility using optical tweezers.} Nature Methods 6:831, 2009.
+  \item Lipfert J, Kerssemakers JWJ, Jager T, and Deller NH. \emph{Magnetic torque tweezers: measuring torsional stiffness in DNA and RecA-DNA filaments.} Nature Methods: 7:977, 2010.
+  \item Dufr\^{e}ne YF, Evans E, Engel A, Helenius J, Gaub HE, and M\"{u}ller DJ. \emph{Five challenges to bringing single-molecule force spectroscopy into living cells.} Nature Methods 8:123, 2011.
+\end{itemize}
 
@@
 
-%\eject
+\eject
 

### 2015-02-16 matlab/attic/analyze-multiple-spring-constants.m

@@ -1,0 +1,80 @@
+% Analyze data from multiple spring constants.
+
+% Clear all
+clear all;
+clear java;
+
+% Add paths to BHMM package.
+addpath /Users/jchodera/Documents/latex/doc/papers/2009/bhmm/matlab/bhmm
+javaaddpath /Users/jchodera/Documents/latex/doc/papers/2009/bhmm/matlab/bhmm
+
+% PARAMETERS
+matfilename = '10khz/fibers_l_10kHz.mat'; % file to analyze
+nsplit = 10; % number of traces to split into.
+
+% Set BHMM options.
+nstates = 2;
+nsamples = 100; % number of samples to generate
+
+options = bhmm_default_options();
+options.maximumIterations = 3; % only need 30 iterations of EM for mlhmm
+options.reversible = true; % infer reversible transition matrices
+options.diagonally_dominant = false; % don't enforce diagonally-dominant transition matrices
+options.verbosity = 2; % set verbosity level
+options.equilibrium = false; % trajectory data is not initially drawn from equilibrium
+options.tau = 1.0 / (10.0e3); % seconds between observations
+
+% Get list of fibers in each file.
+disp(sprintf('querying fibers in "%s"...', matfilename));
+fiber_names = who('-file', matfilename);
+nfibers = length(fiber_names);
+
+for fiber_index = 1:nfibers
+  fiber_name = fiber_names{fiber_index};
+  disp(sprintf('processing fiber "%s"...', fiber_name));
+
+  % Load fiber data.
+  contents = load(matfilename, fiber_name);
+  fiber = getfield(contents, fiber_name);
+  clear contents;
+
+  % Get number of traces.
+  ntraces = length(fiber);
+  for trace_index = 1:ntraces
+    disp(sprintf('trace %d / %d', trace_index, ntraces));
+
+    % Extract trace.
+    o_t = fiber{trace_index};
+    T = length(o_t);
+
+    % Break trace into smaller parts that are less sensitive to drift.
+    data = {};
+    for i = 1:nsplit
+      indices = ((i-1)*floor(T/nsplit)+1):(i*floor(T/nsplit));
+      data{i} = o_t(indices)';
+    end
+
+    % Generate an initial model from data.
+    initial_model = generate_initial_model(data, nstates, options);
+    
+    % Fit a maximum-likelihood HMM.
+    %disp('Determining maximum-likelihood model...');
+    %mlmodel = mlhmm(data, nstates, options);
+    
[…]

### 2015-02-16 matlab/attic/analyze.m

@@ -1,0 +1,195 @@
+
+%***************************************************************************************************************************
+%TAKES THE DATA AND MAKES PRETTY PICTURES AND STUFF
+%PROVIDE AS AN ARGUMENT THE TRAJECTORY NUMBER, FOR EXAMPLE, to analyze trajectory 19 type 'analyze(10)' 
+%***************************************************************************************************************************
+
+
+function analyze(traj_index)
+
+opt = 1
+
+%***************************************************************************************************************************
+%Load the trajectory to analyze
+%***************************************************************************************************************************
+
+	ind = int2str(traj_index); 
+	s = strcat('load trajectory_', ind);
+	eval(s); 
+
+%***************************************************************************************************************************
+%load the actual parameters and initialize (hopefully small) all state matrices if we are working with contrived data
+%***************************************************************************************************************************
+	
+	if opt==0,
+    		s = strcat('load actual_parameters_', int2str(origin_index)); 
+    		eval(s);
+                size(allstates)
+		meanstates = mean(int16(allstates));
+		stdstates = std(int16(allstates));
+	end
+%***************************************************************************************************************************
+%Create means of mus, sigmas, standard deviations, transitions, and sigmas
+%***************************************************************************************************************************
+
+
+	meanmu = mean(allmu);
+	stdmu = std(allmu);
+	meansigma = mean(allsigma);
+	stdsigma = std(allsigma);
+	meantransition = squeeze(mean(alltransitions));
+	stdtransition = squeeze(std(alltransitions));
+	transgraph = zeros(size(alltransitions,3),size(alltransitions,2),size(alltransitions,1));
+
+%************************************************************************************************************************************************************
+	%format all transition matrices to be a readable plot   	
+%************************************************************************************************************************************************************
+
+
+for i = 1:numstates,
+    for j = 1:numstates
+        for k = 1:sample_num,
+               transgraph(k,i,j) = alltransitions(j,i,k); 
+        end
+    end
+end
+
+
+
+%************************************************************************************************************************************************************
+	%plot transition matrices    	
+%************************************************************************************************************************************************************
+
+
+
+
+
+
+for i = 1:numstates,
+	  figure(i);  plot(transgraph(:,:,i)); 
+          xlabel('sample count') ;
+          ylabel('transition probability');
[…]

### 2015-02-16 matlab/attic/initialize.m

@@ -1,0 +1,154 @@
+
+%************************************************************************************************************************************************************
+%FUNCTION WILL INITIALIZE A SAMPLING TRAJECTORY AND GRANT IT A UNIQUE ID NUMBER, ALL TRAJECTORIES WILL BE STORED IN trajectories.m 
+%takes in data, initializes trajectery in output_file_name which should have been generated by makefretdata, makefrets or raw data
+%************************************************************************************************************************************************************
+
+%************************************************************************************************************************************************************
+%arguments are as follows:
+%data_list_index = index of the parent job list
+%out_index = the index of the trajectory being initialized,
+%state_count = the number of states in the run,
+%opt (to be moved to data files) = whether you wish to sample contrived or actual data. 0 = contrived, any other real (and probably complex knowing matlab)number = actual 
+%************************************************************************************************************************************************************
+
+
+
+function initialize(data_list_index, out_index, state_count,opt)
+
+%************************************************************************************************************************************************************
+% Set the number of states and initialize the variables
+% numstates: number of states to be sampled
+% muguess: a vector of guessed means for each state. Each index value corresponds to a numbered state
+% sigmaguess: a vector of instantaneous sigma guesses. Each index corresponds to a numbered state
+% transitionguess: a transition matrix representing the transition probabilities between each state
+%************************************************************************************************************************************************************
+
+
+numstates = state_count;
+muguess = zeros(1,numstates);
+sigmaguess = zeros(1,numstates); 
+transitionguess = zeros(numstates,numstates);	
+
+%************************************************************************************************************************************************************
+% Load in the fret data from a list file
+%************************************************************************************************************************************************************
+
+inid = strcat('data_list_', int2str(data_list_index)); 
+s = inid;  
+eval(s); 
+%indid = strcat('data_set_', int2str(data_list_index));
+
+%************************************************************************************************************************************************************
+%We now have a list of data files. 
+%We now go to each data set in the list and append it to the observations list, each seperated by negative one thousands 
+%In the process we eliminate ludicrous fret values with the filter function (values greater then one or less then zero that are not delimiters)
+%************************************************************************************************************************************************************
+
+clear fretdata
+s = strcat('load ', ' data_set_', int2str(data_list(1)));
+eval(s);
+observations = fretdata; 
+% observation_sets(1).data = filterfrets(fretdata); 
+observation_sets(1).data = fretdata; 
+
+len = length(data_list); 
+
+for i = 2:len,
[…]

### 2015-02-16 matlab/attic/makefrets.m

@@ -1,0 +1,75 @@
+% THIS FILE GENERATES SAMPLE OBSERVATIONS ASSUMING THERE ARE N STATES AND AN NxN TRANSITION MATRIX.  EACH STATE IS DEFINED BY A NORMAL DISTRIBUTION WITH MEAN MU AND STANDARD DEVIATION SIGMA
+
+%accepts d_length = length of fret trajectory
+%accepts data_index = index of data file to save to
+
+
+function makefretdata(d_length,data_index)
+
+
+outid = strcat(' data_set_', int2str(data_index)); 
+outactid = strcat(' actual_parameters_', int2str(data_index)); 
+
+% Definitions for mu, sigma, and the transition matrix
+numstates = 4;
+fretmu = [0.1 0.3 0.6 0.9];
+fretsigma = [0.1 0.1 0.1 0.1];
+% transitionmatrix = [0.6 0.1 0.2 0.1; 0.1 0.2 0.6 0.1; 0.5 0.1 0.3 0.1; 0.1 0.7 0.1 0.1];
+transitionmatrix = [0.98 0.01 0.005 0.005; 0.01 0.975 0.01 0.05; 0.05 0.01 0.975 0.01; 0.05 0.05 0.01 0.98];
+datalength = d_length;
+
+% calculate the eigenvalues and eigenvectors
+[fretevec, freteval] = eig(transitionmatrix');
+[sortedeigs, sortedorder] = sort(diag(freteval), 'descend');
+
+% Calculate the stationary eigenvector and the cumulative probabilities of the stationary eigenvector
+stationary = fretevec(:, sortedorder(1));
+stationary = stationary ./ sum(stationary);
+cumstationary = stationary;
+for i = 2:numstates,
+  cumstationary(i) = cumstationary(i-1) + cumstationary(i);
+end
+
+% Calculate the cumulative transition probabilities
+cumprobs = transitionmatrix;
+for i = 1:numstates,
+  for j = 2:numstates,
+    cumprobs(i,j) = cumprobs(i,j-1) + cumprobs(i,j);
+  end
+end
+
+%fid = fopen(strcat(filename, '.m', 'w');
+
+
+fretdata = zeros(1, datalength); 
+
+
+i=1;
+% Define the initial position
+q = rand();
+[n, mybin] = histc(q, cumstationary);
+mystate(i) = mybin+1;
+% Pick the datapoint from the normal distribution corresponding to the state
[…]

### 2015-02-16 matlab/attic/normalprobability.m

@@ -1,0 +1,4 @@
+
+function probability = normalprobability(value, mu, sigma)
+
+  probability = 1/((2 * pi)^(1/2)*sigma) * exp(- (value - mu)^2/(2*sigma^2));

### 2015-02-16 matlab/attic/order_states.m

@@ -1,0 +1,28 @@
+%GIVEN NEW DEFFINITIONS ORDERING OF STATES, RESORTS THE MU, SIGMA, AND TRANSITION MATRIX TO REFLECT. 
+
+
+function [sorted_mu sorted_sigma sorted_transition] = order_states(new_indices, mu, sigma, transition)
+
+
+
+numstates = size(new_indices,2); 
+
+%sort mu values, sigma values, and columns of transition matrix 
+
+for i = 1:numstates,
+	new_index = new_indices(i); 
+	sorted_mu(new_index) = mu(i); 
+	sorted_sigma(new_index) = sigma(i); 
+	sorted_transition(:,new_index) = transition(:,i); 
+end
+
+%sort rows of transition matrix
+
+temp = sorted_transition; 
+
+for i = 1:numstates,
+	new_index = new_indices(i); 
+	sorted_transition(new_index,:) = temp(i,:); 
+end
+
+

### 2015-02-16 matlab/attic/ratematrix/hummer_rate_matrix_update.m

@@ -1,0 +1,114 @@
+function Tij = hummer_rate_matrix_update(Tij, Nij, tau)
+% Produce a (correlated) sample of the transition matrix corresponding to a true rate matrix that satisfies detailed balance using scheme of Gerhard Hummer.
+%
+% ARGUMENTS
+%  Tij (MxM matrix) - current transition matrix sample 
+%    This transition matrix must correspond to a valid rate matrix, such that Tij = expm(Kij * tau)
+%  Nij (MxM matrix) - transition conditional count matrix
+%    Nij(i,j) is number of observed counts terminating in j at time tau given that they initially started in i at time zero.
+%  tau - evolution time to relate rate matrix to transition matrix through Tij = expm(Kij * tau)
+%
+% RETURN VALUES
+%  Tij_new (MxM matrix) - updated sample of transition matrix, may be correlated with previous sample
+%
+% NOTES
+%  The algorithm described in [1] and modified in [2] (especially with regard to the prior) is implemented.
+%  Transitions between all states are allowed -- no constraints on the sparisty of Kij are imposed.
+%
+% REFERENCES
+%  [1] Sriraman S, Kevrekidis IG, and Hummer G. Coarse master equation from Bayesian analysis of replica molecular dynamics simulations. JPC B 109(14):6479-6484, 2005.
+%  [2] Buchete N-V and Hummer G. Coarse master equations for peptide folding dynamics. JPC B 112(19):6057-6069, 2008.
+%
+% TODO
+%  Check to be sure we don't need a correction for asymmetric proposal probability.
+
+% PARAMETERS
+nsteps = 1000; % number of update steps (set to give approximately uncorrelated new sample)
+delta = 1.0; % move width in logarithmic parameterization (set to give acceptance probability of ~ 50%)
+
+% Determine size of transition matrix.
+M = size(Tij,1);
+
+% Compute compressed vector from initial transition matrix.
+x = pack_vector(Tij, tau);
+x0 = x'; % store initial vector
+
+% Compute initial log-likelihood.
+logL = log_likelihood(Tij, Nij);
+
+% Storage
+x_ni = zeros(nsteps,M-1 + M*(M-1)/2);
+
+% Update sample using Hummer scheme.
+naccept = 0; % number of accepted moves
+for step = 1:nsteps
[…]

### 2015-02-16 matlab/attic/ratematrix/pack_vector.m

@@ -1,0 +1,49 @@
+function x = pack_vector(Tij, tau)
+% Convert a transition matrix into a packed version of Hummer's parameterization of the rate matrix.
+% The transition matrix Tij must be determined from a proper rate matrix such that
+%
+% Tij = exp(Kij * tau)
+%
+% ARGUMENTS
+%  Tij (MxM) - row-stochastic transition matrix that satsifes Tij = exp(Kij * tau) for rate matrix Kij with positive off-diagonal elements
+%  tau - lag time relating Kij and Tij
+%
+% RETURN VALUES
+%  x (M-1 + M(M-1)/2 vector) - packed form of Hummer's parameterization of a rate matrix satisfying detailed balance
+%    stationary probabilities Pi(i) = x(i) for i = 1..(M-1)
+%    upper triangle of rates K(i,j) are derived from x(M:end)   
+
+% Determine number of states.
+M = size(Tij,1);
+
+% Determine number of unique elements for dimension of x.
+Nunique = M-1 + M*(M-1)/2; % number of unique elements not determined by detailed balance
+
+% Allocate storage for x.
+x = zeros(Nunique, 1); % allocate storage for vector
+
+% Determine rate matrix from Tij.
+Kij = logm(Tij) / tau;
+
+% Compute stationary probability.
+%Pi = null(Kij');
+options = optimset;
+options.disp = 0;
+[Pi,lambda] = eigs(Tij',1,'LM',options);
+Pi = Pi / sum(Pi);
+
+% Compute elementwise logarithms for working parameters.
+logPi = log(Pi);
+logKij = log(Kij - 2*diag(diag(Kij))); % take care to ensure diagonal of Kij is positive
+
+% Store elements 1..(M-1) of logPi.
+x(1:(M-1)) = logPi(1:(M-1));
+
+% Store upper right triangle of logKij.
+index = M;
+for i = 1:M
+  x(index:index+(M-i-1)) = logKij(i,i+1:end);
+  index = index + M-i;
+end
+
+return

### 2015-02-16 matlab/attic/ratematrix/test_transition_matrix_update.m

@@ -1,0 +1,45 @@
+% Test transition matrix updating scheme.
+
+% Clear all data
+clear;
+
+% Define a test count matrix.
+%
+% This count matrix for the alanine dipeptide six-state decomposition comes from Eq. 27 of:
+% Hinrichs NS and Pande VS. Calculation of the distribution of eigenvalues and eigenvectors in Markovian state models for molecular dynamics. JCP 126:244101, 2007.
+
+Nij = [ 
+    4380  153   15    2    0    0;
+     211 4788    1    0    0    0;
+     169    1 4604  226    3    0;
+       3   13  158 4823    3    0;
+       0    0    0    4 4978   18;
+       7    5    0    0   62 4926;
+    ]
+
+tau = 0.1; % lag time (ps)
+
+M = size(Nij, 1); % number of states
+
+% Simple estimate of Tij that does not obey detailed balance.
+Tij_reference = zeros(size(Nij));
+for i = 1:M
+  Tij_reference(i,:) = Nij(i,:) / sum(Nij(i,:));
+end
+
+% Generate an initial guess.
+disp('Computing maximum-likelihood estimate of Tij...');
+Tij = generate_transition_matrix_guess(Nij, tau);
+
+% Compare the result.
+disp('Tij estimate from counting (without detailed balance):')
+Tij_reference
+disp('Tij estimate from likelihood maximization (with detailed balance constraint)')
+Tij
+
+% Produce samples.
+disp('Sampling with Hummer method...');
+for i = 1:10
+  Tij = hummer_rate_matrix_update(Tij, Nij, tau)
+end
+

### 2015-02-16 matlab/attic/sortBeta.m

@@ -1,0 +1,21 @@
+function [sorted_mu sorted_sigma sorted_transition] = sortBeta(mu, sigma, transition)
+
+%WILL SORT STATES IN ACCORDANCE TO RULES OF THE COMPARESTATES FUNCTION, ASCENDING
+%RETURNS SORTED MU, SIGMA, AND TRANSITION MATRIX 
+
+numstates = size(mu,2); 
+
+states = 1:numstates;
+
+sorted_mu = mu; 
+sorted_sigma = sigma; 
+sorted_transition = transition;
+
+
+%MERGE SORT STATE DEFFINITIONS
+new_indices = state_merge_sort(states, mu, sigma, transition);
+
+
+%REORDER MU SIGMA AND TRANSITION MATRIX ACCORDING TO NEW STATE DEFFINITIONS 
+
+[sorted_mu sorted_sigma sorted_transition] = order_states(new_indices, mu, sigma, transition);

### 2015-02-16 matlab/attic/state_merge.m

@@ -1,0 +1,33 @@
+%MERGES TWO LISTS OF STATE INDICES IN ASCENDING SORTED ORDER BASED ON COMPARE STATES SEE http://en.wikipedia.org/wiki/Merge_sort 
+
+
+function merged_indices = state_merge(left, right, mu, sigma, transition)
+
+
+    merged_indices = zeros(1); 
+    merged_indices(1,:) = []; 
+
+
+while size(left,2) > 0 && size(right, 2) > 0,
+	
+    
+	if comparestates(mu(left(1)), sigma(left(1)), transition(left(1),left(1)), mu(right(1)), sigma(right(1)), transition(right(1),right(1)))>-1,
+	
+		merged_indices = [merged_indices left(1)] ;
+		left(:,1) = []; 
+
+	else
+		merged_indices = [merged_indices right(1)]; 
+		right(:,1) = []; 
+	end
+
+end
+
+if size(left,2) > 0,
+	merged_indices = [merged_indices, left];
+end
+
+if size(right,2) > 0,
+	merged_indices = [merged_indices, right]; 
+end	
+

### 2015-02-16 matlab/attic/writeout.m

@@ -1,0 +1,11 @@
+%***********************************************************************************************************************************************************
+%Script will write out a trajectory 
+%************************************************************************************************************************************************************
+
+outid = strcat(' trajectory_',int2str(out_index));
+
+s = strcat('save', outid, ' origin_file opt origin_index numstates stateguess muguess sigmaguess transitionguess allmu allsigma alltransitions allstates data_list_index correct observation_sets observations guess_counts sample_num datalength'); 
+
+
+eval(s); 
+

### 2015-02-16 matlab/bhmm/Contents.m

@@ -1,0 +1,86 @@
+% Bayesian hidden Markov model (BHMM) analysis of timeseries data
+%
+% The data is assumed to come from a fixed number of states each emitting real-valued signals corresponding to 
+% one-dimensional i.i.d. normal random variates.  Each state is characterized by a mean and variance, inferred 
+% from the data.  Transitions between the states are governed by a transition matrix, and dynamics is assumed
+% to be Markovian.
+%
+%
+% EXAMPLE USE
+%
+% % Find maximum-likelihood model for data.
+% maximum_likelihood_model = mlhmm(data, nstates);
+%
+% % Generate samples from Bayesian HMM.
+% bayesian_model_samples = bhmm(data, maximum_likelihod_model);
+%
+%
+% PROVIDED FUNCTIONS
+%
+% * Test harnesses
+%
+% test_bhmm                     - test Bayesian HMM on synthetic data
+%
+% * Hidden Markov model construction and sampling:
+%
+% bhmm_default_options          - generate an options structure for BHMM containing default parameters
+% mlhmm                         - determine maximum-likelihood HMM using EM algorithm
+% bhmm                          - generate samples from Bayesian HMM
+%
+% * Analysis of BHMM output:
+%
+% analyze_rates                 - analyze transition rate distributions and confidence intervals
+% analyze_states                - analyze state emission probabilities and confidence intervals
+%
+% * Construction of other classes of models:
+%
+% generate_segmentation_model   - construct model based on segmentation of the data using Gaussian mixture model EM fit
+%
+% * Auxiliary functions:
+%
+% generate_initial_model        - generate initial valid HMM model
+% em_gaussian_mixture           - fit a mixture of Gaussians using EM algorithms (used for initialization)
+%
+% plot_state_assignments        - plot observed trajectory with colored state assignments from a particular model
+%
+% update_emission_probabilities - update state emission parameters by sampling E ~ P(E | S,O)
+% update_transition_matrix      - update transition matrix by sampling T ~ P(T | S)
+% update_state_trajectories     - update state tranjectories by sampling S ~ P(S | T,E,O)
+% rate_matrix_sample            - generate an uncorrelated sample from the rate matrix
+% transition_matrix_mle         - generate maximum-likelihood estimate of transition matrix
+% transition_matrix_sample      - generate a new (potentially correlated) sample from the transition matrix according to specified option structure
+% transition_matrix_sample_reversible - generate a new reversible transition matrix sample
[…]

### 2015-02-16 matlab/bhmm/analyze_transition_matrix.m

@@ -1,0 +1,38 @@
+function analyze_transition_matrix(models, options)
+% Analyze transition probability or rate constant distribution.
+%
+% TODO: Allow options.nbins to override hard-coded nbins default below.
+
+nstates = models(1).nstates;
+nmodels = length(models);
+nbins = 40; % number of histogram bins for plotting
+ci = 0.95; % confidence interval to report
+
+Tijn = zeros(nstates, nstates, nmodels);
+for n = 1:nmodels
+  Tijn(:,:,n) = models(n).Tij;
+end
+
+% Plot.
+clf;
+for i = 1:nstates
+  for j = 1:nstates
+    subplot(nstates,nstates,nstates*(i-1)+j);
+    Tn = squeeze(Tijn(i,j,:));
+    Tmean = mean(Tn);
+    [Tlow, Thigh] = confidence_interval(Tn, 0.5 - ci/2, 0.5 + ci/2);
+    [N,X] = hist(Tn, nbins);
+    bar(X,N,'grouped','k');
+
+    % Label plot.
+    title(sprintf('%.5f [%.5f, %.5f]', Tmean, Tlow, Thigh), 'FontSize', 7);
+    xlabel('');
+    ylabel(sprintf('T_{%d %d}', i, j));
+    set(gca, 'YTick', []);    
+
+    % Adjust axes.
+    axis([-0.05 1.05 0 1.15*max(N)]);
+  end
+end
+
+return

### 2015-02-16 matlab/bhmm/baum_welch.m

@@ -1,0 +1,71 @@
+function [log_xi_tij, log_gamma_ti] = baum_welch(o_t, model, options)
+% Baum-Welch algorithm for computing expected transition and state observation counts.
+% 
+% ARGUMENTS
+%   o_t (T) [real array] - trajectory of real-valued observations
+%   model [structure] - model structure
+%   options [structure] - options structure
+%
+% RETURNS
+%   log_xi_tij (T,N,N) [real array] - log-probability of observing transition  - xi_tij(t,i,j) = P(X_t = i, X_{t+1} = j | O, model)
+%   log_gamma_ti (T,N) [real array] - log-probability observing state i at time t - gamma_ti(t,i) = P(X_t = i | O, model)
+
+% Constants.
+T = length(o_t); % number of observations in this trajectory
+N = model.nstates; % number of states
+
+if options.equilibrium
+  % Experiment starts in equilibrium.
+  log_rho_i = model.logPi;
+elseif isfield(options, 'log_rho_i')
+  % Use specified nonequilibrium starting conditions.
+  log_rho_i = options.log_rho_i;
+else
+  % Use uniform initial starting conditions, which should be equivalent to complete ignorance.
+  log_rho_i = model.logPi * 0.0;
+  log_rho_i = log_rho_i - log(sum(exp(log_rho_i)));
+end
+
+% Use Java code.
+if options.use_java
+  % Prepare data for fast Java helper code.
+  mu = zeros(model.nstates,1);
+  sigma = zeros(model.nstates,1);
+  for i = 1:model.nstates
+    mu(i) = model.states{i}.mu;
+    sigma(i) = model.states{i}.sigma;
+  end  
+  % Update model parameters.
+  model.logTij = log(model.Tij); % elementwise logarithm - can we use a more numerically stable approach later?
+  model.logPi = log(model.Pi);
+
+  % Perform Baum-Welch.
+  % log-probability of observing transition i to j at time t : xi_tij(t,i,j) = P(X_t = i, X_{t+1} = j | O, model)
+  log_alpha_ti = bhmm_helper.forwardAlgorithm(o_t, mu, sigma, model.logPi, model.logTij, log_rho_i);  
+  log_beta_ti = bhmm_helper.backwardAlgorithm(o_t, mu, sigma, model.logPi, model.logTij, log_rho_i);  
[…]

### 2015-02-16 matlab/bhmm/bayesian_gaussian_mixture.m

@@ -1,0 +1,162 @@
+function model = bayesian_gaussian_mixture(data, nstates, options)
+% Fit a Gaussian mixture model to observed data using the EM algorithm using random initialization.
+%
+% model = bayesian_gaussian_mixture(data, nstates, options)
+%
+% ARGUMENTS
+%   data (cell array of 1D arrays) - observed trajectories of some real-valued signal
+%   nstates - number of Gaussian components to fit
+%   options - options data structure
+%
+% RETURNS
+%   model (structure) - initial model parameters
+%
+% NOTES
+%
+%
+% REFERENCES
+%   
+
+if (options.verbosity >= 1)
+  disp('******************************************************************************');
+  disp('Gaussian mixture model fitting by expectation maximization (EM) algorithm');
+end
+
+% DEBUG
+options.enforce_state_ordering = true; % enforce ordering of states at this stage
+RELATIVE_TOLERANCE = options.convergenceTolerance;
+  
+% Create model structure.
+model = struct();
+
+% Store number of states.
+model.nstates = nstates;
+
+% Determine number of trajectories.
+nTrajectories = length(data);
+
+% Concatenate all observations.
+observations = [];
+for trajectoryIndex = 1:nTrajectories
+  % Extract trajectory.
+  o_t = data{trajectoryIndex};
+  % Make into a row vector.
+  o_t = reshape(o_t, 1, length(o_t));
+  % Concatenate observations.
+  observations = [observations o_t];  
+end
+nobservations = length(observations);
+
+%
+% Generate initial guess by randomly sampling from the data
+%
+
+nobservations;
+random_permutation = randperm(nobservations);
+random_centers = random_permutation(1:nstates);
+distances = abs(observations'*ones(1,nstates) - ones(nobservations,1)*observations(random_centers))
+[mindistance, initialstatelabels] = min(distances')
+
+% Allocate storage for model parameters.
+mu = zeros(nstates,1); % mu(i) is the Gaussian mean for state i
+sigma = zeros(nstates,1); % sigma(i) is the Guassian standard deviation for state i
+Pi = zeros(nstates,1); % Pi(i) is the normalized weight for Gaussian state i
+
+% Determine mean and std dev of each cluster.
+for i = 1:nstates
[…]

### 2015-02-16 matlab/bhmm/bhmm_default_options.m

@@ -1,0 +1,42 @@
+function options = bhmm_default_options()
+% Generate structure containing default options for Bayesian HMM.
+%
+% RETURNS
+%  options (struct) - structure containing default options for Bayesian HMM
+%
+% NOTES
+%
+% The options structure contains the following fields:
+%
+% updateMethod : {'bayesian', 'maximum-likelihood'} Choice of method to update HMM each iteration
+
+options = struct();
+
+options.updateMethod = 'bayesian'; % method for updates - one of {'bayesian', 'maximum-likelihood'}
+options.tau = 1.0; % time between observations
+
+options.maximumIterations = 1000; % maximum number of allowed iterations
+options.convergenceTolerance = 1.0e-4; % relative convergence tolerance for EM
+
+% Options for transition matrix estimation.
+options.reversible = true; % infer reversible transition matrices
+options.diagonally_dominant = true; % enforce diagonally-dominant transition matrices
+options.verbosity = 10; % set verbosity level (0 = silent, 1 = minimal, 2 = more output, ..., 10)
+options.equilibrium = true; % trajectory data is initially drawn from equilibrium
+
+options.observable_name = 'force'; % label for obserable axis
+options.observable_units = 'pN'; % units for observable axis
+options.time_units = 's'; % units for time axis
+
+options.sample_states = 0; % controls whether number of hidden states is to be sampled
+options.minstates = 2; % minimum number of states if sample_states is set
+options.maxstates = 100; % maximum number of states if sample_states is set
+
+% Options for state observables.
+options.enforce_state_ordering = true; % enforce states are ordered in increasing order
+options.enforce_state_separation = false; % if nonzero, will enforce state means to be separated by specified amount
+
+options.use_java = 1; % use Java acceleration
+
+return
+

### 2015-02-16 matlab/bhmm/combine_confidence_intervals.m

@@ -1,0 +1,34 @@
+function [low, high] = combine_confidence_intervals(xlow, xhigh, interval)
+% Combine multiple confidence intervals using Gaussian posterior assumption.
+%
+% [low, high] = combine_confidence_intervals(xlow, xhigh, interval)
+%
+% ARGUMENTS
+%   xlow (1D array) - low end of symmetric confidence intervals about the mean to be combined
+%   xhigh (1D array) - high end of symmetric confidence intervals about the mean to be combined
+%   interval - fraction of probability encompassed by symmetric confidence intervals about the mean (e.g. 0.95 for 95% CI)
+%
+% RETURNS
+%   low - low end of combined confidence interval
+%   high - high end of combined confidence interval
+
+% Determine how many sigma the given confidence interval is equal to.
+nsigma = sqrt(2) * erfinv(interval); % 'interval' represents 'nsigma' std devs of gaussian
+
+% Compute means and standard deviations.
+mu_n = (xlow + xhigh) / 2.0;
+sigma_n = (xhigh - xlow) / 2.0 / nsigma;
+
+% Combine means and standard deviations.
+sigma = sum(sigma_n.^(-2))^(-1/2);
+mu = sum(sigma_n.^(-2) .* mu_n) / sum(sigma_n.^(-2));
+
+% DEBUG
+[mu, sigma]
+
+% Compute new confidence interval.
+low = mu - nsigma * sigma;
+high = mu + nsigma * sigma;
+
+return
+

### 2015-02-16 matlab/bhmm/compute_autocorrelation.m

@@ -1,0 +1,29 @@
+function [tvec, C_t] = compute_autocorrelation(o_t, tmax, options)
+% Estimate normalized fluctuation autocorrelation function for observed timeseries data.
+%
+% [tvec, C_t] = compute_autocorrelation(o_t, tmax, options)
+%
+
+% Determine timeseries length.
+T = length(o_t);
+
+% Compute fluctuation timeseries.
+o_mean = mean(o_t);
+do_t = o_t - o_mean;
+
+% Compute unnormalized correlation function.
+C_t = zeros(1,tmax+1);
+for t = 0:tmax
+  C_t(t+1) = mean( do_t(1:T-t) .* do_t(1+t:T) );
+end  
+
+% Normalize correlation function.
+C_t = C_t / C_t(1);
+
+% Compute times.
+tvec = (0:tmax) * options.tau;
+
+return;
+
+
+

### 2015-02-16 matlab/bhmm/compute_log_likelihood.m

@@ -1,0 +1,53 @@
+function log_likelihood = compute_log_likelihood(model, data, options, use_state_assignments)
+% Compute the log-likelihood of the HMM given the data.
+%
+% log_likelihood = compute_log_likelihood(model, data, options, use_state_assignments)
+%
+% ARGUMENTS
+%   model (structure) - model parameters, containing state trajectories
+%   data (cell array of 1D arrays) - observed trajectories of some real-valued signal
+%  options - optional options structure
+%    if options.reversible is true, then reversible transition matrix sampling will be used 
+%    if options.diagonally_dominant is true, then only diagonally-dominant transition matrices will be accepted
+%    if options.verbosity sets level of verbose output (0 = none, 1 = summary report, 2 = output each iteration)
+%    if options.equilibrium is true, trajectories are initially in equilibrium
+%  use_state_assignments - if true, will use state assignments; otherwise, will marginalize them out
+%
+% RETURNS
+%   log_likelihood (double) - log-likelihood of the HMM 'model' given the data 'data', or P(O | \Theta)
+%
+% NOTE
+%   We compute the probability of a trajectory conditioned on the first observation, so that the trajectory may not necessarily be in equilibrium.
+
+% Calculate the stationary probability distribution from dominant eigenvector of transition matrix.
+model.logTij = log(model.Tij); % elementwise logarithm - can we use a more numerically stable approach later?
+
+% Initialize log likelihood.
+log_likelihood = 0.0;
+
+% Compute log likelihood summed over all trajectories.
+for trajectory_index = 1:length(data)
+  % Extract trajectory of observations.
+  o_t = data{trajectory_index}; % o_t(t) is the observation at time t \in 1...T.
+
+  % Aggregate model parameters into vectors.
+  mu = zeros(model.nstates,1);
+  sigma = zeros(model.nstates,1);
+  for i = 1:model.nstates
+    mu(i) = model.states{i}.mu;
+    sigma(i) = model.states{i}.sigma;
+  end  
+
+  % Accumulate log-likelihood.
+  if use_state_assignments
+    % Extract hidden state trajectory.
+    s_t = model.state_trajectories{trajectory_index};  
+  
+    log_likelihood = log_likelihood + bhmm_helper.computeTrajectoryLogLikelihood(s_t, o_t, mu, sigma, model.logPi, model.logTij, options.equilibrium);
+  else
[…]

### 2015-02-16 matlab/bhmm/compute_log_posterior.m

@@ -1,0 +1,61 @@
+function log_posterior = compute_log_posterior(model, data, options, use_state_assignments)
+% Compute the log posterior of the HMM given the data.
+%
+% log_posterior = compute_log_posterior(model, data, options, use_state_assignments)
+%
+% ARGUMENTS
+%   model (structure) - model parameters, containing state trajectories
+%   data (cell array of 1D arrays) - observed trajectories of some real-valued signal
+%  options - optional options structure
+%    if options.reversible is true, then reversible transition matrix sampling will be used 
+%    if options.diagonally_dominant is true, then only diagonally-dominant transition matrices will be accepted
+%    if options.verbosity sets level of verbose output (0 = none, 1 = summary report, 2 = output each iteration)
+%    if options.equilibrium is true, trajectories are initially in equilibrium
+%  use_state_assignments - if true, will use state assignments; otherwise, will marginalize them out
+%
+% RETURNS
+%   log_posterior (double) - log-likelihood of the HMM 'model' given the data 'data', or P(O | \Theta)
+%
+% NOTE
+%   We compute the probability of a trajectory conditioned on the first observation, so that the trajectory may not necessarily be in equilibrium.
+
+% Calculate the stationary probability distribution from dominant eigenvector of transition matrix.
+model.logTij = log(model.Tij); % elementwise logarithm - can we use a more numerically stable approach later?
+
+% Initialize log posterior.
+log_posterior = 0.0;
+
+% Compute transition matrix prior.
+% TODO: Check detailed balance.
+
+% Compute emission prior.
+for i = 1:model.nstates
+  log_posterior = log_posterior - log(model.states{i}.sigma);
+end
+
+% Compute log likelihood summed over all trajectories.
+for trajectory_index = 1:length(data)
+  % Extract trajectory of observations.
+  o_t = data{trajectory_index}; % o_t(t) is the observation at time t \in 1...T.
+
+  % Aggregate model parameters into vectors.
+  mu = zeros(model.nstates,1);
+  sigma = zeros(model.nstates,1);
+  for i = 1:model.nstates
+    mu(i) = model.states{i}.mu;
+    sigma(i) = model.states{i}.sigma;
+  end  
+
+  % Accumulate log-likelihood.
+  if use_state_assignments
+    % Extract hidden state trajectory.
+    s_t = model.state_trajectories{trajectory_index};  
+  
[…]

### 2015-02-16 matlab/bhmm/draw.m

@@ -1,0 +1,30 @@
+function outcome = draw(Pi)
+% Draw a random outcome given vector of outcome probabilities.
+%
+% ARGUMENTS
+%   Pi (N vector) - vector of outcome probabilities
+%
+% RETURNS
+%   outcome - integer outcome \in 1..N
+
+if any(Pi < 0)
+  Pi  
+end
+
+% return random outcome
+outcome = randsample(length(Pi), 1, true, Pi);
+return
+
+% Compute cumulative probability.
+CMF = cumsum(Pi);
+
+% Normalize just in case.
+CMF = CMF / CMF(end);
+
+% Draw an outcome.
+r = rand;
+[n,bin] = histc(r, CMF);
+outcome = bin + 1;
+
+return
+

### 2015-02-16 matlab/bhmm/extract_model_parameters.m

@@ -1,0 +1,17 @@
+function [Tij, mu, sigma] = extract_model_parameters(model)
+% Extract model parameters, returning them as a list of matrices and vectors.
+%
+% [Tij, mu, sigma] = extract_model_parameters(model)
+
+Tij = model.Tij;
+nstates = model.nstates;
+mu = zeros(nstates,1);
+sigma = zeros(nstates,1);
+for i = 1:nstates
+  mu(i) = model.states{i}.mu;
+  sigma(i) = model.states{i}.sigma;
+end
+
+return
+
+  

### 2015-02-16 matlab/bhmm/generate_initial_model.m

@@ -1,0 +1,73 @@
+function model = generate_initial_model(data, nstates, options)
+% Generate a (poor) initial model for hidden Markov model (HMM).
+%
+% model = generate_initial_model(data, nstates, options)
+%
+% ARGUMENTS
+%   data (cell array of 1D arrays) - observed trajectories of some real-valued signal
+%   nstates - number of states
+%   options - options data structure
+%
+% RETURNS
+%   model (structure) - initial model parameters
+%
+% NOTES
+%   A (poor) guess at the initial model parameters is generated.
+%   This guess must be subsequently refined by iterations of either the maximum-likeihood or Bayesian HMM algorithms.
+%
+% TODO
+%   Implement initial guess via EM algorithm
+%   Guess number of states by multiple-normal density fitting?
+
+if (options.verbosity >= 1)
+  disp('******************************************************************************');
+  disp('Bayesian HMM initial model generation');
+end
+  
+% Create initial model from Gaussian mixture model.
+model = em_gaussian_mixture(data, nstates, options);
+
+% Generate initial transition matrix.
+% Compute expected transition counts using only fractional state assignemnts.
+Nij = zeros(nstates, nstates); % Nij(i,j) is fractional expected transition counts from i to j
+% Construct transition count matrix from state trajectories.
+mu = zeros(nstates, 1);
+sigma = zeros(nstates, 1);
+Pi = zeros(nstates, 1);
+for i = 1:nstates
+  mu(i) = model.states{i}.mu;
+  sigma(i) = model.states{i}.sigma;
+  Pi(i) = model.Pi(i);
+end
+for trajectory_index = 1:length(data)
+  % Extract trajectory of observables.
+  o_t = data{trajectory_index};
+  T = length(o_t);
+
+  % Compute fractional assignment of samples to states
+  p_ti = bhmm_helper.computeStateProbabilities(o_t, mu, sigma, Pi);
+
+  % Accumulate transition counts from this trajectory.
+  for t = 1:(T-1)
+    Nij = Nij + p_ti(t,:)' * p_ti(t+1,:);
+  end
+end
+% Update transition matrix estimate.
+model.Tij = transition_matrix_mle(Nij, options);
+clear mu sigma Pi Nij;
+
+% Compute stationary probability.
+model.Pi = stationary_probability(model.Tij);
+model.logPi = log(model.Pi);
+
+% Assign samples based on probability alone.
+%disp('Generating initial guess of state trajectory...');
+%mloptions = options;
[…]

### 2015-02-16 matlab/bhmm/hardcoded_model.m

@@ -1,0 +1,4 @@
+function model = hardcoded_model()
+
+
+return

### 2015-02-16 matlab/bhmm/mlhmm.m

@@ -1,0 +1,143 @@
+function mlmodel = mlhmm(data, nstates, options, initial_model)
+% Construct maximum-likelihod hidden Markov model (HMM), useful for initializing a Bayesian HMM.
+%
+% mlmodel = mlhmm(data, nstates, options, [initial_model])
+%
+% ARGUMENTS
+%   data (cell array of 1D arrays) - observed trajectories of some real-valued signal (e.g. trap extension, FRET efficiency)
+%   nstates (int) - number of states to fit
+%   options - options structure to control model generation
+%
+% OPTIONAL ARGUMENTS
+%   initial_model - model to initialized MLHMM procedure
+%
+% RETURNS
+%   mlmodel (structure) - maximum likelihood model after fit to data
+%
+% NOTES
+%   Trajectories provided in 'data' cell array may be of different lengths. 
+%   The expectation-maximization (EM) algorithm is used for updates; this is a variation on Baum-Welch.
+%   This may not produce the global maximum-likelihood model.
+%
+% TODO
+%   Add ability for user to pass in optional 'options' structure.
+
+% Import fast Java helper code.
+import bhmm_helper.*;
+
+% DEBUG
+RELATIVE_TOLERANCE = options.convergenceTolerance;
+
+if (nargin == 3)
+  % Generate a valid initial model (which may be a poor guess).
+  model = generate_initial_model(data, nstates, options);
+else
+  % Use specified initial model.
+  model = initial_model;
+end
+
+% Determine number of trajectories provided in 'data'.
+ntrajectories = length(data);
+
+if (options.verbosity >= 1)
+  disp('Fitting maximum-likelihood HMM with EM algorithm');
+end
+
+% Run cycles of maximum-likelihood updating.
+for iteration = 1:options.maximumIterations
+
+  if (options.verbosity >= 2)
+    disp('******************************************************************************');
+    disp(sprintf('EM iteration %d', iteration));
+  end
+
+  % Store old model.
+  [old_Tij, old_mu, old_sigma] = extract_model_parameters(model);
+
+  % Update emission probability functions associated with each state.
+  for i = 1:model.nstates
+    state = model.states{i};
+    % Define normal density for emission probability.
+    model.states{i}.emission_probability = @(o) (2*pi)^(-1/2) * state.sigma^(-1) * exp(-(1/2)*((o - state.mu) / state.sigma)^2); 
[…]

### 2015-02-16 matlab/bhmm/noe_transition_matrix_update.m

@@ -1,0 +1,123 @@
+function Tij = noe_transition_matrix_update(Tij, Nij, tau)
+% Produce a (correlated) sample of the transition matrix corresponding to a true rate matrix that satisfies detailed balance using scheme of Gerhard Hummer.
+%
+% Tij = hummer_rate_matrix_update(Tij, Nij, tau)
+%
+% ARGUMENTS
+%  Tij (MxM matrix) - current transition matrix sample 
+%    This transition matrix must correspond to a valid rate matrix, such that Tij = expm(Kij * tau)
+%  Nij (MxM matrix) - transition conditional count matrix
+%    Nij(i,j) is number of observed counts terminating in j at time tau given that they initially started in i at time zero.
+%  tau - evolution time to relate rate matrix to transition matrix through Tij = expm(Kij * tau)
+%
+% RETURN VALUES
+%  Tij_new (MxM matrix) - updated sample of transition matrix, may be correlated with previous sample
+%
+% NOTES
+%  The algorithm described in [1] and modified in [2] (especially with regard to the prior) is implemented.
+%  Transitions between all states are allowed -- no constraints on the sparisty of Kij are imposed.
+%
+% REFERENCES
+%  [1] Sriraman S, Kevrekidis IG, and Hummer G. Coarse master equation from Bayesian analysis of replica molecular dynamics simulations. JPC B 109(14):6479-6484, 2005.
+%  [2] Buchete N-V and Hummer G. Coarse master equations for peptide folding dynamics. JPC B 112(19):6057-6069, 2008.
+%
+% TODO
+%  * Check to be sure we don't need a correction for asymmetric proposal probability.
+%  * Automatically adjust delta to provide ~ 50% acceptance before fixing delta.
+%  * Automatically determine 'nsteps' to provide an uncorrelated sample.
+%  * Speed up sampling with Java methods?
+%  * Consider starting sampling from maximum-likelihood estimate if we're going to produce a completely decorrelated estimate.
+
+% PARAMETERS
+nsteps = 1000; % number of update steps (set to give approximately uncorrelated new sample)
+delta = 1.0; % move width in logarithmic parameterization (set to give acceptance probability of ~ 50%)
+
+% Determine size of transition matrix.
+M = size(Tij,1);
+
[…]

### 2015-02-16 matlab/bhmm/pack_vector2.m

@@ -1,0 +1,15 @@
+function x = pack_vector2(logKij, logPi)
+
+M = length(logPi);
+
+% Store elements 1..(M-1) of logPi.
+x(1:(M-1)) = logPi(1:(M-1));
+
+% Store upper right triangle of logKij.
+index = M;
+for i = 1:M
+  x(index:index+(M-i-1)) = logKij(i,i+1:end);
+  index = index + M-i;
+end
+
+return

### 2015-02-16 matlab/bhmm/plot_correlated_model_parameters.m

@@ -1,0 +1,77 @@
+function plot_correlated_model_parameters(models, options)
+% Generate LaTeX table showing mean and confidence intervals for all properties.
+%
+% ARGUMENTS
+%   models - a set of BHMM-sampled models
+
+% Augment models with mu and sigma vectors.
+nmodels = length(models);
+nstates = models(1).nstates;
+clear augmented_models;
+for model_index = 1:nmodels
+  model = models(model_index);
+  model.mu_i = zeros(model.nstates,1);
+  model.sigma_i = zeros(model.nstates,1);
+  for i = 1:model.nstates
+    model.mu_i(i) = model.states{i}.mu;
+    model.sigma_i(i) = model.states{i}.sigma;      
+  end
+  augmented_models(model_index) = model;
+end
+models = augmented_models;
+
+% Augment models with rate matrices and lifetimes.
+clear augmented_models;
+for model_index = 1:nmodels
+  model = models(model_index);
+  model.Kij = real(logm(model.Tij)) / options.tau;
+  model.tau_i = options.tau ./ (1 - diag(model.Tij));
+  augmented_models(model_index) = model;
+end
+models = augmented_models;
+
+properties = {'Pi', 'Tij', 'Kij', 'tau_i', 'mu_i', 'sigma_i'};
+nproperties = length(properties);
+% Compute model parameters.
+model_parameters = struct()
+for property_index = 1:nproperties
+  property = properties{property_index};
+  statistics = struct();
+  
+  % Collect statistic into vector.
+  nmodels = length(models);
+  x = getfield(models(1), property);
+  property_size = size(x);
+  property_dim = length(property_size);
+  if property_size(end) == 1
+    property_dim = property_dim - 1;
+  end
+  x_n = zeros([nmodels size(x)]);
+  for n = 1:nmodels
+    model = models(n);
+    x = getfield(model, property);
+    if (property_dim==1)
+      x_n(n,:) = x;
+    else
+      x_n(n,:,:) = x;
+    end
+  end    
+
+  model_parameters = setfield(model_parameters, property, x_n);
+
+end
+
+% Generate plot.
+clf;
+%plot(squeeze(model_parameters.Kij(:,3,1)), squeeze(model_parameters.Kij(:,3,2)), '.');
[…]

### 2015-02-16 matlab/bhmm/rate_matrix_sample.m

@@ -1,0 +1,189 @@
+function model = rate_matrix_sample(model, Nij, mode)
+% Generate an uncorrelated transition matrix sample using the method of Gerhard Hummer.
+%
+% Tij = sample_transition_matrix(Tij, Nij, tau)
+%
+% ARGUMENTS
+%  Tij (MxM matrix) - current transition matrix sample 
+%    This transition matrix must correspond to a valid rate matrix, such that Tij = expm(Kij * tau)
+%  Nij (MxM matrix) - transition conditional count matrix
+%    Nij(i,j) is number of observed counts terminating in j at time tau given that they initially started in i at time zero.
+%  tau - evolution time to relate rate matrix to transition matrix through Tij = expm(Kij * tau)
+%
+% OPTIONAL ARGUMENTS
+%  mode - if mode is 'sample', sampling will take place; if mode is 'optimize', then MC optimization is used
+%
+% RETURN VALUES
+%  Tij_new (MxM matrix) - updated sample of transition matrix, may be correlated with previous sample
+%
+% NOTES
+%  The transition matrix is guaranteed to satisfy detailed balance and correspond to a rate matrix with positive off-diagonal entries.
+%
+%  The algorithm described in [1] and modified in [2] (especially with regard to the prior) is implemented.
+%  Transitions between all states are allowed -- no constraints on the sparisty of Kij are imposed.
+%
+% REFERENCES
+%  [1] Sriraman S, Kevrekidis IG, and Hummer G. Coarse master equation from Bayesian analysis of replica molecular dynamics simulations. JPC B 109(14):6479-6484, 2005.
+%  [2] Buchete N-V and Hummer G. Coarse master equations for peptide folding dynamics. JPC B 112(19):6057-6069, 2008.
+%
+% TODO
+%  * Check to be sure we don't need a correction for asymmetric proposal probability.
+%  * Automatically adjust delta to provide ~ 50% acceptance before fixing delta.
+%  * Automatically determine 'nsteps' to provide an uncorrelated sample.
+%  * Speed up sampling with Java methods?
+%  * Consider starting sampling from maximum-likelihood estimate if we're going to produce a completely decorrelated estimate.
+
+% Handle optional arguments
[…]

### 2015-02-16 matlab/bhmm/save_models.m

@@ -1,0 +1,17 @@
+function save_models(filename, models)
+% Save critical model data (without state trajectories) to specified file.
+%
+% save_models(filename)
+%
+% ARGUMENTS
+%  filename - file to save models to
+
+% Determine number of models.
+nmodels = length(models);
+
+% Create storage for compressed models.
+packed_models = [];
+
+for model = 1:nmodels
+end
+

### 2015-02-16 matlab/bhmm/show_trajectory.m

@@ -1,0 +1,7 @@
+function show_trajectory(s_t)
+
+disp(sprintf('%d', s_t));
+
+return
+
+

### 2015-02-16 matlab/bhmm/stationary_probability.m

@@ -1,0 +1,24 @@
+function Pi = stationary_probability(Tij)
+% Compute stationary probability of a row-stochastic transition matrix.
+%
+% Pi = stationary_probability(Tij)
+%
+% ARGUMENTS
+%  Tij (N x N matrix) - row-stochastic transition matrix
+%
+% RETURNS
+%  Pi (N vector) - vector of stationary probabilities
+%
+% TODO
+%  * Use a more numerically stable / efficient method here, such as null(Tij' - eye(N))?
+
+% Compute stationary eigenvector.
+options = struct();
+options.disp = 0;
+options.issym = false;
+[Pi,lambda] = eigs(Tij',1,'LR',options);
+
+% Normalize to obtain stationary probability.
+Pi = Pi / sum(Pi); 
+
+return

### 2015-02-16 matlab/bhmm/test_bhmm.m

@@ -1,0 +1,37 @@
+% Test Bayesian HMM with synthetic data.
+
+% Clear workspace.
+clear;
+
+% PARAMETERS
+T = 10000; % observation length
+
+% Generate test model and data.
+[true_model, data] = generate_test_model(T);
+true_model.Pi
+
+% Set parameters.
+options = bhmm_default_options(); % get default options structure
+options.maximumIterations = 10; % maximum number of allowed iterations
+options.convergenceTolerance = 1.0e-4; % relative convergence tolerance in likelihood
+options.tau = 1;
+
+% Options for transition matrix estimation.
+options.reversible = true; % infer reversible transition matrices
+options.diagonally_dominant = false; % don't enforce diagonally-dominant transition matrices
+options.verbosity = 2; % set verbosity level
+options.equilibrium = false; % trajectory data is not initially drawn from equilibrium
+
+% Determine maximum-likelihood HMM.
+disp('Determining maximum-likelihood model...');
+mlmodel = mlhmm(data, true_model.nstates, options);
+
+% Sample from Bayesian HMM starting from maximum-likelihood model.
+disp('Sampling from Bayesian posterior...');
+nsamples = 100;
+models = bhmm(data, mlmodel, nsamples, options);
+
+% Plot
+plot_state_assignments(data, models(end), options);
+
+

### 2015-02-16 matlab/bhmm/transition_matrix_mle.m

@@ -1,0 +1,125 @@
+function Tij = transition_matrix_mle(cij, options)
+% Generate maximum-likelihood estimate of row-stochastic transition matrix using iterative procedure of Ref. [1].
+%
+% WARNING
+%  There is currently no constraint ensuring sum(Pi(1:N-1)) < 1 during likelihood maximization.  This needs to be fixed before this code is usable.
+%
+% Tij = transition_matrix_mle(cij, options)
+%
+% ARGUMENTS
+%  cij (MxM matrix) - transition conditional count matrix (row form), including any prior pseudocounts
+%    cij(i,j) is number of observed counts terminating in j at time tau given that they initially started in i at time zero.
+%  options - optional options structure
+%    if options.reversible is true, then reversible transition matrix sampling will be used 
+%    if options.diagonally_dominant is true, then only diagonally-dominant transition matrices will be accepted
+%    if options.verbosity sets level of verbose output (0 = none, 1 = summary report, 2 = output each iteration)
+%    if options.equilibrium is true, trajectories are initially in equilibrium
+%
+% RETURN VALUES
+%  Tij (MxM matrix) - maximum likelihood row-stochastic transition matrix
+%
+% REFERENCES
+%
+% [1] Prinz JH, Wu H, Sarich M, Keller B, Fischbach M, Held M, Chodera JD, Schuette C, and Noe F. 
+% Markov models of molecular kinetics: Generation and validation. Submitted.
+%
+% TODO
+%  * Enforce positivity in xij elements.
+
+% Parameters.
+MAXITS = 1000; % maximum number of iterations
+RELATIVE_TOLERANCE = 1.0e-5; % relative convergence tolerance
+EPSILON = 1.0e-3; % fractional counts to add to each transition to avoid numerical issues
+
+if (options.verbosity >= 2)
+  disp('Computing maximum-likelihood estimate of transition matrix');
+  disp('cij = ');
+  disp(cij);
+end
+
+% Determine size of transition count matrix.
+M = size(cij,1);
+
+% Cast count matrix to doubles.
[…]

### 2015-02-16 matlab/bhmm/transition_matrix_sample.m

@@ -1,0 +1,33 @@
+function Tij = transition_matrix_sample(Tij, Nij, options)
+% Generate an updated transition matrix sample T ~ P(T | N).
+%
+% Tij = sample_transition_matrix(Tij, Nij, tau)
+%
+% ARGUMENTS
+%  Tij (MxM matrix) - current transition matrix sample 
+%    This transition matrix must be reversible.
+%  Nij (MxM matrix) - transition conditional count matrix
+%    Nij(i,j) is number of observed counts terminating in j at time tau given that they initially started in i at time zero.
+%  options - optional options structure
+%    if options.reversible is true, then reversible transition matrix sampling will be used 
+%    if options.diagonally_dominant is true, then only diagonally-dominant transition matrices will be accepted
+%    if options.verbosity sets level of verbose output (0 = none, 1 = summary report, 2 = output each iteration)
+%    if options.equilibrium is true, trajectories are initially in equilibrium
+%
+% RETURN VALUES
+%  Tij_new (MxM matrix) - updated sample of transition matrix, may be correlated with previous sample
+%
+% NOTES
+%  If options.reversible = true, the transition matrix is guaranteed to satisfy detailed balance.
+%  The transition matrix will not necessarily correspond to a rate matrix.
+%
+%  The algorithm described in [1] and modified in [2] (especially with regard to the prior) is implemented.
+%  Transitions between all states are allowed -- no constraints on the sparisty of Kij are imposed.
+
+if (options.reversible)
+  Tij = transition_matrix_sample_reversible(Tij, Nij, options);
+else
+  Tij = transition_matrix_sample_nonreversible(Tij, Nij, options);
+end
+
+return

### 2015-02-16 matlab/bhmm/transition_matrix_sample_nonreversible.m

@@ -1,0 +1,101 @@
+function Tij = transition_matrix_sample_nonreversible(Tij, Nij, options)
+% Generate an updated nonreversible transition matrix sample T ~ P(T | N) using Dirichlet sampling.
+%
+% Tij = sample_transition_matrix(Tij, Nij, tau)
+%
+% ARGUMENTS
+%  Tij (MxM matrix) - current transition matrix sample 
+%    This transition matrix must be reversible.
+%  Nij (MxM matrix) - transition conditional count matrix
+%    Nij(i,j) is number of observed counts terminating in j at time tau given that they initially started in i at time zero.
+%  options - optional options structure
+%    if options.diagonally_dominant is true, then only diagonally-dominant transition matrices will be accepted
+%    if options.verbosity sets level of verbose output (0 = none, 1 = summary report, 2 = output each iteration)
+%    if options.equilibrium is true, trajectories are initially in equilibrium
+%    if options.prior_transition_pseudocounts exists (a KxK matrix where the (i,j) element represents transitions from i to j), these pseudocounts are included; all ones is uniform prior
+%
+% RETURN VALUES
+%  Tij_new (MxM matrix) - updated sample of nonreversible transition matrix, may be correlated with previous sample
+%
+% NOTES
+%  If options.reversible = true, the transition matrix is guaranteed to satisfy detailed balance.
+%  The transition matrix will not necessarily correspond to a rate matrix.
+%
+%  The algorithm described in [1] and modified in [2] (especially with regard to the prior) is implemented.
+%  Transitions between all states are allowed -- no constraints on the sparisty of Kij are imposed.
+% REFERENCES
+%  [1] Singhal
+
+% PARAMETERS
+ROW_SUM_TOLERANCE = 1.0e-3; % row sum tolerance
+
+% Include prior transition pseudocounts.
+if isfield(options, 'prior_transition_pseudocounts')
+  Nij = double(Nij) + double(options.prior_transition_pseudocounts);
+else
+  % DEBUG: Add pseudocounts.
+  Nij = double(Nij) + 1.0;  
+end
+
+% Determine size of transition matrix.
+M = size(Tij,1);
+
+% Promote Nij to double.
+Nij = double(Nij);
+
+% Compute total number of observed transitions out of each state.
+Ni = sum(Nij,2);
[…]

### 2015-02-16 matlab/bhmm/unpack_vector2.m

@@ -1,0 +1,42 @@
+function [logKij, logPi, Kij, Tij] = unpack_vector2(x, tau)
+
+% Determine dimension of packed vector.
+Nunique = length(x);
+
+% Determine number of states.
+M = round(-(1/2) + sqrt(1/4 + 2*(Nunique+1)));
+
+% Allocate storage for Hummer's parameterization.
+logPi = zeros(M,1);
+logKij = zeros(M,M);
+
+% Extract logarithms of stationary probabilities.
+logPi(1:M-1) = x(1:M-1);
+logPi(M) = log(1.0 - sum(exp(logPi(1:M-1))));
+
+% Extract logarithms of unique rate matrix entries.
+index = M;
+for i = 1:M  
+  logKij(i,i+1:end) = x(index:index+(M-i-1));
+  index = index + M-i;
+end
+
+% Use detailed balance to determine lower triangle of rate matrix entries.
+for i = 1:M
+  for j = 1:i-1
+    logKij(i,j) = logKij(j,i) + logPi(j) - logPi(i);
+  end
+end
+
+% Compute rate matrix.
+Kij = exp(logKij);
+for i = 1:M
+  Kij(i,i) = 0.0;
+  Kij(i,i) = - sum(Kij(i,:));
+end
+
+% Compute transition matrix.
+Tij = expm(Kij * tau);
+
+return
+

### 2015-02-16 matlab/bhmm/update_number_of_states.m

@@ -1,0 +1,348 @@
+function model = update_number_of_states(data, model, options)
+% Update number of hidden states using reversible-jump scheme.
+%
+% model = update_number_of_states(data, model, options)
+%
+% ARGUMENTS
+%   data (cell array of 1D arrays) - observed trajectories of some real-valued signal
+%   model (structure) - model parameters, containing state trajectories
+%   options (structure) - options structure
+%
+% RETURNS
+%   model (structure) - model with updated number of states
+%
+% NOTES
+%   The hidden state assignment trajctory must be redrawn immediately afterwards.
+%   If options.minstates is specified, the number of states will not be allowed to fall below this.
+%   If options.maxstates is specified, the number of states will not be allowed to rise above this.
+%
+% TODO
+% - Instead of just rejecting moves that create too few or too many states, only propose
+%   new models with valid numbers of states.
+
+% DEBUG
+options.verbosity = 10;
+
+% Do nothing if sample_states is not set, or if minstates == maxstates.
+if (options.sample_states == 0) || (options.minstates == options.maxstates)
+  return
+end
+
+% DEBUG
+show_model(model)
+
+% Choose split or merge move.
+% Here, we split or merge with equal probability, unless we're at min or max states.
+move_type = '';
+if (model.nstates == options.maxstates)
+  move_type = 'merge';
+elseif (model.nstates == options.minstates)
+  move_type = 'split';
+elseif (rand < 0.5)
+  move_type = 'merge';
+else
+  move_type = 'split';
+end
+
+% Compute model likelihood, ignoring hidden state assignments.
+if (options.verbosity >= 5)
+  disp('Computing log-likelihood...');
+end
+log_likelihood = compute_log_likelihood(model, data, options, 0)
+if (options.verbosity >= 6)
+  disp(sprintf('log P(O | \\Theta_old) = %f', log_likelihood));
+end
+
+% Attempt move
+if (move_type == 'merge')
+  %
+  % Merge model states a and b into newmodel state i.
+  %
[…]

### 2015-02-16 matlab/bhmm/update_state_trajectories.m

@@ -1,0 +1,130 @@
+function model = update_state_trajectories(data, model, options)
+% Update state trajectories given emission probabilities and transition matrix.
+%
+% model = update_state_trajectories(data, model, options)
+%
+% ARGUMENTS
+%   data (cell array of 1D arrays) - observed trajectories of some real-valued signal
+%   model (structure) - model parameters, containing state trajectories
+%   options (struct) - options for run - see bhmm_default_options()
+%
+% RETURNS
+%   model (structure) - model with updated state trajectories.
+%
+% NOTE
+%   We compute the probability of a trajectory conditioned on the first observation, so that the trajectory may not necessarily be in equilibrium.
+%   A modified version of the forward-backward algorithm (due to N. S. Hinrichs) is used in the Bayesian sampling.
+
+% Associate emission probabilities with states.
+% TODO: Move this elsewhere?
+for i = 1:model.nstates
+  state = model.states{i};
+  % Define normal density for emission probability.
+  model.states{i}.emission_probability = @(o) (2*pi)^(-1/2) * state.sigma^(-1) * exp(-(1/2)*((o - state.mu) / state.sigma)^2); 
+  model.states{i}.log_emission_probability = @(o) - (1/2)*log(2*pi) - log(state.sigma) - (1/2)*((o - state.mu) / state.sigma)^2;
+end
+
+% Calculate the stationary probability distribution from dominant eigenvector of transition matrix.
+model.logTij = log(model.Tij); % elementwise logarithm - can we use a more numerically stable approach later?
+model.logPi = log(model.Pi);
+
+% Update state trajectories.
+for trajectory_index = 1:length(data)
+  % Extract trajectory of observations.
+  o_t = data{trajectory_index}; % o_t(t) is the observation at time t \in 1...T.
+
+  switch (options.updateMethod)
+   case 'bayesian'
+    % Sample a state trajectory.
+
+    % Use fast Java helper code.
+    mu = zeros(model.nstates,1);
+    sigma = zeros(model.nstates,1);
+    for i = 1:model.nstates
+      mu(i) = model.states{i}.mu;
+      sigma(i) = model.states{i}.sigma;
+    end  
[…]

### 2015-02-16 matlab/bhmm/viterbi.m

@@ -1,0 +1,64 @@
+function [s_t, log_trajectory_probability] = viterbi(o_t, model)
+% Viterbi algorithm for computing most probable state trajectory given trajectory of observations.
+%
+% ARGUMENTS
+%   o_t (T array) - trajectory of observations
+%   model - model structure
+%
+% RETURNS
+%   s_t (T array of int32) - most probable state trajectory (s_t(t) \in 1..model.nstates)
+%   log_trajectory_probability - log trajectory probability
+
+% Determine trajectory length.
+T = length(o_t);
+
+%
+% Initialization
+%
+
+s_t = zeros(1,T,'int32'); % s_t(t) is the most likely state trajectory time t
+%delta_it = zeros(model.nstates, T, 'double');
+log_delta_it = zeros(model.nstates, T, 'double');
+Phi_it = zeros(model.nstates, T, 'int32');
+
+for i = 1:model.nstates
+  %delta_it(i,1) = model.Pi(i) .* model.states{i}.emission_probability(o_t(1));
+  log_delta_it(i,1) = model.logPi(i) + model.states{i}.log_emission_probability(o_t(1));  
+  Phi_it(i,1) = 0;
+end
+
+%
+% Recursion
+%
+
+for t = 2:T
+  for j = 1:model.nstates
+    %[valmax, argmax] = max(delta_it(:,t-1) .* model.Tij(:,j));
+    %delta_it(j,t) = valmax * model.states{j}.emission_probability(o_t(t));
+    
+    [log_valmax, argmax] = max(log_delta_it(:,t-1) + model.logTij(:,j));
+    log_delta_it(j,t) = log_valmax + model.states{j}.log_emission_probability(o_t(t));    
+    
+    Phi_it(j,t) = argmax;
+  end
+end
+
+%
+% Termination
+%
+
+%[valmax, argmax] = max(delta_it(:,T));
+%trajectory_probability = valmax;
+[log_valmax, argmax] = max(log_delta_it(:,T));
+log_trajectory_probability = log_valmax;
+s_t(T) = argmax;
+
+%
+% Reconstruction
+%
+
+for t = (T-1):-1:1
+  s_t(t) = Phi_it(s_t(t+1),t+1);
+end
+
+return

### 2015-02-16 matlab/validation/analyze_with_bhmm.m

@@ -1,0 +1,84 @@
+% Analyze traces with Bayesian HMM to compute average forces and lifetimes for different trap extensions.
+
+clear;
+
+addpath ../../matlab/bhmm
+javaaddpath ../../matlab/bhmm
+
+fibername = 'fiber3';
+fiber_data = sprintf('calibrated/%s.mat', fibername); % file to load, containing 'trace' cellarray with multiple traces to be analyzed.
+subsample = 50; % subsampling fraction -- subsample to 1 kHz
+sampletime = (1/50000) * subsample; % time between samples, in seconds
+nstates = 3; % number of states -- fixed beforehand
+nsamples = 1000; % number of models to sample with Bayesian HMM
+nburnin = 50; % number of samples to discard to equilibration
+
+% Load data.
+load(fiber_data);
+ntraces = length(trace);
+
+% Allocate storage for estimates
+estimates = cell(ntraces,1); % estimates{trace_index} is a struct containing estimates of state properties for each trace
+
+for trace_index = 1:ntraces  
+  % Extract data.
+  o_t = trace{trace_index}';
+
+  % Subsample data.
+  o_t = o_t(1:subsample:end);
+
+  % Pack just this trace into a cell array.
+  data = {};
+  data{1} = o_t;
+
+  % Determine maximum-likelihood HMM.
+  %disp('Determining maximum-likelihood model...');
+  %mlmodel = mlhmm(data, model.nstates);
+  
+  options = bhmm_default_options();
+  options.maximumIterations = 10; % maximum number of allowed iterations
+  options.convergenceTolerance = 1.0e-4; % relative convergence tolerance in likelihood
+
+  % Options for transition matrix estimation.
+  options.reversible = true; % infer reversible transition matrices
+  options.diagonally_dominant = false; % don't enforce diagonally-dominant transition matrices
+  options.verbosity = 2; % set verbosity level
+  options.equilibrium = false; % trajectory data is not initially drawn from equilibrium
+  options.maximumIterations = 2; % maximum number of EM iterations for maximum-likelihood HMM (WARNING: THESE ITERATIONS ARE SLOW)
+
+  options.tau = sampletime; % seconds between observations
+
+  % Generate an initial (poor) model from data.
+  %initial_model = generate_initial_model(data, nstates, options);
+  
+  % Determine maximum-likelihood HMM.
+  disp('Determining maximum-likelihood model...');
[…]

### 2015-02-16 matlab/validation/empirical_confidence_interval.m

@@ -1,0 +1,23 @@
+function [low, high] = empirical_confidence_interval(sample, interval)
+% Compute specified symmetric confidence interval for empirical sample.
+%
+% ARGUMENTS
+%  sample - empirical sample
+%  interval - desired confidence interval (0 < interval < 1) 
+%    e.g. 0.68 for 68% confidence interval, 0.95 for 95% confidence interval
+
+% Sort sample in increasing order.
+sample = sort(sample);
+
+% Determine sample size.
+N = length(sample);
+
+% Compute low and high indices.
+low_index = round((N-1) * (0.5 - interval/2)) + 1;
+high_index = round((N-1) * (0.5 + interval/2)) + 1;
+
+% Compute low and high.
+low = sample(low_index);
+high = sample(high_index);
+
+return

### 2015-02-16 matlab/validation/four_state_model.m

@@ -1,0 +1,31 @@
+function [model] = four_state_model()
+% Generate a four-state test model.
+%
+%  [model] = four_state_model()
+%
+% RETURNS
+%  model - a four-state test model
+
+% Define a model.
+model = struct();
+
+% Define number of states.
+model.nstates = 4;
+
+% Define state emission probabilities.
+model.states{1} = struct('mu', 0.1, 'sigma', 0.1);
+model.states{2} = struct('mu', 0.3, 'sigma', 0.1);
+model.states{3} = struct('mu', 0.6, 'sigma', 0.1);
+model.states{4} = struct('mu', 0.9, 'sigma', 0.1);
+
+% Define a transition matrix that satisfies detailed balance and comes from a true rate matrix.
+model.Tij = [0.98 0.01  0.005 0.005; 
+	     0.01 0.975 0.01  0.05; 
+	     0.05 0.01  0.975 0.01; 
+	     0.05 0.05  0.01  0.98];
+model.Pi = stationary_probability(model.Tij);
+
+% Store model parameters.
+model.tau = 1.0;
+
+return

### 2015-02-16 matlab/validation/generate_problematic_test_model.m

@@ -1,0 +1,74 @@
+function [model, data] = generate_test_model(T)
+% Generate a test model and synthetic data.
+%
+%  [model, data] = generate_test_model()
+%
+% RETURNS
+%  model - the model from which the data is simulated
+%  data  - the synthetic test data generated from the model
+%
+% NOTE
+%  The random number seed is set deterministically.
+
+% Set up synthetic test system.
+disp('******************************************************************************');
+disp('Creating synthetic test system...');
+
+% DEBUG Seed random number generator deterministically.
+rand('seed', 1);
+
+% Define a model.
+model = struct();
+
+% Define number of states.
+model.nstates = 3;
+
+% Define state emission probabilities.
+model.states{1} = struct('mu', -30, 'sigma', 1.0); % unfolded
+model.states{2} = struct('mu', -29, 'sigma', 1.0); % intermediate
+model.states{3} = struct('mu', 0, 'sigma', 0.2); % folded
+
+% Define a transition matrix that satisfies detailed balance and comes from a true rate matrix.
+model.Tij = [    0.980 0.019 0.001;
+		 0.05  0.90  0.05;
+		 0.001 0.009 0.990];
+        
+model.Pi = stationary_probability(model.Tij);
+
+% Ensure that Tij satisfies detailed balance.
+for i = 1:model.nstates
+  for j = 1:(i-1)
+    model.Tij(i,j) = model.Pi(j) * model.Tij(j,i) / model.Pi(i);
+  end
+end
+model.Pi = stationary_probability(model.Tij);
+
+% Store model parameters.
+model.tau = 0.001;
+
+% Generate synthetic data.
+%T = 50000; % trajectory length
+
+% Generate state trajectory.
+s_t = zeros(1,T,'int32'); % s_t(t) is state visited at time t
+s_t(1) = draw(model.Pi);
+for t = 2:T
+  s_t(t) = draw(model.Tij(s_t(t-1),:));
+end
+model.state_trajectories{1} = s_t; 
+
+% Generate observation trajectory.
+o_t = zeros(1,T,'single'); % o_t(t) is observation at time t
+for t = 1:T
+  state = model.states{s_t(t)};
+  o_t(t) = randn() * state.sigma + state.mu;
+end
+
+% Pack data into cellarray.
+data{1} = o_t;
+
+show_model(model);
+
+disp('******************************************************************************');
+
+return

### 2015-02-16 matlab/validation/generate_test_model.m

@@ -1,0 +1,74 @@
+function [model, data] = generate_test_model(T)
+% Generate a test model and synthetic data.
+%
+%  [model, data] = generate_test_model()
+%
+% RETURNS
+%  model - the model from which the data is simulated
+%  data  - the synthetic test data generated from the model
+%
+% NOTE
+%  The random number seed is set deterministically.
+
+% Set up synthetic test system.
+disp('******************************************************************************');
+disp('Creating synthetic test system...');
+
+% DEBUG Seed random number generator deterministically.
+rand('seed', 1);
+
+% Define a model.
+model = struct();
+
+% Define number of states.
+model.nstates = 3;
+
+% Define state emission probabilities.
+model.states{1} = struct('mu', 3.0, 'sigma', 1.0); % unfolded
+model.states{2} = struct('mu', 4.7, 'sigma', 0.3); % intermediate
+model.states{3} = struct('mu', 5.6, 'sigma', 0.2); % folded
+
+% Define a transition matrix that satisfies detailed balance and comes from a true rate matrix.
+model.Tij = [    0.980 0.019 0.001;
+		 0.05  0.90  0.05;
+		 0.001 0.009 0.990];
+        
+model.Pi = stationary_probability(model.Tij);
+
+% Ensure that Tij satisfies detailed balance.
+for i = 1:model.nstates
+  for j = 1:(i-1)
+    model.Tij(i,j) = model.Pi(j) * model.Tij(j,i) / model.Pi(i);
+  end
+end
+model.Pi = stationary_probability(model.Tij);
+
+% Store model parameters.
+model.tau = 0.001;
+
+% Generate synthetic data.
+%T = 50000; % trajectory length
+
+% Generate state trajectory.
+s_t = zeros(1,T,'int32'); % s_t(t) is state visited at time t
+s_t(1) = draw(model.Pi);
+for t = 2:T
+  s_t(t) = draw(model.Tij(s_t(t-1),:));
+end
+model.state_trajectories{1} = s_t; 
+
+% Generate observation trajectory.
+o_t = zeros(1,T,'single'); % o_t(t) is observation at time t
+for t = 1:T
+  state = model.states{s_t(t)};
+  o_t(t) = randn() * state.sigma + state.mu;
+end
+
+% Pack data into cellarray.
+data{1} = o_t;
+
+show_model(model);
+
+disp('******************************************************************************');
+
+return

### 2015-02-16 matlab/validation/plot_confidence_intervals.m

@@ -1,0 +1,71 @@
+% Plot confidence interval data.
+
+clf;
+
+% Pi
+subplot(2,2,1);
+plot([0 1], [0 1], 'k-');
+hold on;
+[Plow, Phigh] = beta_confidence_intervals(ci_Pi, ntrials, 0.95);
+%errorbar(repmat(cis',1,nstates), ci_Pi, sqrt(ci_Pi .* (1-ci_Pi))/sqrt(ntrials),'.');
+errorbar(repmat(cis',1,nstates), ci_Pi, ci_Pi - Plow, Phigh - ci_Pi, '.');
+axis square;
+set(gca, 'XTick', [0 1]);
+set(gca, 'YTick', [0 1]);
+%title('\pi_i');
+text(0.1, 0.85, '\pi_i');
+
+% Tij
+
+ci_X = zeros(ncis, nstates*nstates);
+for i = 1:nstates
+  for j = 1:nstates
+    ci_X(:,i+nstates*(j-1)) = ci_Tij(:,i,j);
+  end
+end
+
+% TODO: Compute more accurate confidence intervals for these estimates using Bayesian inference.  See earlier work.
+
+subplot(2,2,2);
+plot([0 1], [0 1], 'k-');
+hold on;
+[Plow, Phigh] = beta_confidence_intervals(ci_X, ntrials, 0.95);
+%errorbar(repmat(cis',1,nstates*nstates), ci_X, sqrt(ci_X .* (1-ci_X))/sqrt(ntrials),'.');
+errorbar(repmat(cis',1,nstates*nstates), ci_X, ci_X - Plow, Phigh - ci_X,'.');
+axis square;
+set(gca, 'XTick', [0 1]);
+set(gca, 'YTick', [0 1]);
+%title('T_{ij}');
+text(0.1, 0.85, 'T_{ij}');
+
+% mu
+subplot(2,2,3);
+plot([0 1], [0 1], 'k-');
+hold on;
+[Plow, Phigh] = beta_confidence_intervals(ci_mu_i, ntrials, 0.95);
+%errorbar(repmat(cis',1,nstates), ci_mu_i, sqrt(ci_mu_i .* (1-ci_mu_i))/sqrt(ntrials),'.');
+errorbar(repmat(cis',1,nstates), ci_mu_i, ci_mu_i - Plow, Phigh - ci_mu_i, '.');
+axis square;
+set(gca, 'XTick', [0 1]);
+set(gca, 'YTick', [0 1]);
[…]

### 2015-02-16 matlab/validation/test_confidence_intervals_random.m

@@ -1,0 +1,128 @@
+% Generate synthetic data to test the confidence intervals genereated from BHMM sampling.
+
+% Clear workspace.
+clear;
+
+% Add BHMM paths.
+addpath ../../matlab/bhmm
+javaaddpath ../../matlab/bhmm
+
+% Test parameters.
+ntrials = 50; % number of independent realizations of the experiment to generate
+cis = 0.05:0.05:0.95; % confidence intervals to evaluate
+ncis = length(cis);
+observation_lengths = [10000]; % length of observation trajectories
+nmodels = 1000; % number of models to sample with Bayesian HMM
+nburnin = 50; % number of samples to discard to equilibration
+
+% Set parameters.
+options = bhmm_default_options(); % get default options structure
+options.maximumIterations = 10; % maximum number of allowed EM iterations
+options.convergenceTolerance = 1.0e-4; % relative convergence tolerance in EM iterations
+options.tau = 1; % time interval between observations
+
+% Options for transition matrix estimation.
+options.reversible = true; % infer reversible transition matrices
+options.diagonally_dominant = false; % don't enforce diagonally-dominant transition matrices
+options.verbosity = 3; % set verbosity level
+options.equilibrium = true; % trajectory data is initially drawn from equilibrium
+
+% Options for plots.
+options = rmfield(options, 'observable_units'); % no units for observable axis
+options = rmfield(options, 'time_units'); % no units for time axis
+options.observable_name = 'obs';
+
+%% Run many uncorrelated experiments to test confidence intervals.
+
+% Allocate data.
+ci_Pi = zeros(ncis,1);
+ci_Pi_denom = zeros(ncis,1);
+ci_Tij = zeros(ncis,1);
+ci_Tij_denom = zeros(ncis,1);
+ci_mu_i = zeros(ncis,1);
+ci_mu_i_denom = zeros(ncis,1);
+ci_sigma_i = zeros(ncis,1);
+ci_sigma_i_denom = zeros(ncis,1);
+
+for trial = 1:ntrials
+  % Generate synthetic model and observation data.
[…]

### 2015-02-16 matlab/validation/test_multistate_sampling.m

@@ -1,0 +1,120 @@
+% Generate synthetic data to test the confidence intervals genereated from BHMM sampling.
+
+% Clear workspace.
+clear;
+
+% Add BHMM paths.
+addpath ../../matlab/bhmm
+javaaddpath ../../matlab/bhmm
+
+% Test parameters.
+observation_lengths = [10000]; % length of observation trajectories
+sigma = 0.2;
+true_model = three_state_model(sigma); % test model to use.
+nstates = true_model.nstates;
+nmodels = 1000; % number of models to sample with Bayesian HMM
+nburnin = 50; % number of samples to discard to equilibration
+
+% Set parameters.
+options = bhmm_default_options(); % get default options structure
+options.maximumIterations = 10; % maximum number of allowed EM iterations
+options.convergenceTolerance = 1.0e-4; % relative convergence tolerance in EM iterations
+options.tau = 1; % time interval between observations
+
+% Options for transition matrix estimation.
+options.reversible = true; % infer reversible transition matrices
+options.diagonally_dominant = false; % don't enforce diagonally-dominant transition matrices
+options.verbosity = 3; % set verbosity level
+options.equilibrium = true; % trajectory data is not initially drawn from equilibrium
+
+% Options for plots.
+options = rmfield(options, 'observable_units'); % no units for observable axis
+options = rmfield(options, 'time_units'); % no units for time axis
+options.observable_name = 'obs';
+
+% Turn on state sampling.
+options.enforce_state_ordering = 0; % don't enforce state ordering for now
+options.sample_states = 1; % sample over number of states
+
+% Show the true model.
+show_model(true_model);
+
+%% Generate initial data and model to illustrate.
+
+% Generate synthetic data from the true model.
+data = generate_synthetic_data(true_model, observation_lengths);
+
+% Determine MLHMM.
+disp('Determining MLHMM...');
+mlmodel = mlhmm(data, true_model.nstates, options);
+
+% DEBUG
+models = [mlmodel];
+for i = 1:100
+  model = models(end);
+  models = bhmm(data, model, 1, options);
+  plot_state_assignments(data, models(end), options);
+  pause
+end
+
+% Initialize Bayesian HMM.
[…]

### 2015-02-16 matlab/validation/test_table_segmentation.m

@@ -1,0 +1,75 @@
+% Generate LaTeX table.
+nlengths = NL;
+
+string = '\begin{tabular}{|c|c|';
+for length_index = 1:NL
+  string = sprintf('%sc', string);
+end
+string = sprintf('%s|}',string);
+disp(string)
+disp('\hline');
+
+% header line
+disp(sprintf('property & true value & \\multicolumn{%d}{c|}{observation length} \\\\', nlengths));
+string = ' &';
+for length_index = 1:NL
+  string = sprintf('%s & $10^%d$', string, round(log10(L_n(length_index))));
+end
+string = sprintf('%s \\\\ \\hline', string);
+disp(string)
+
+% \pi_i
+property = 'Pi';
+for i = 1:nstates
+  string = sprintf('$\\pi_{%d}$ & $%.3f$', i, true_model.Pi(i));
+  for length_index = 1:NL
+    string = sprintf('%s & $%.3f$', string, model_store{length_index}.Pi(i));
+  end
+  string = sprintf('%s \\\\', string);
+  disp(string)
+end  
+
+disp('\hline');
+
+% T_ij
+property = 'Tij';
+for i = 1:nstates
+  for j = 1:nstates
+    string = sprintf('$T_{%d%d}$ & $%.3f$', i, j, true_model.Tij(i,j));
+    for length_index = 1:NL
+      string = sprintf('%s & $%.3f$', string, model_store{length_index}.Tij(i,j));
+    end
+    string = sprintf('%s \\\\', string);
+    disp(string)
+  end
+end  
+
+disp('\hline');
+
+% mu_i
+property = 'mu_i';
+for i = 1:nstates
+  string = sprintf('$\\mu_{%d}$ & $%.3f$', i, true_model.states{i}.mu);
+  for length_index = 1:NL
+    string = sprintf('%s & $%.3f$', string, model_store{length_index}.states{i}.mu);
+  end
+  string = sprintf('%s \\\\', string);
+  disp(string)
+end  
+
+disp('\hline');
+
+% sigma_i
+property = 'sigma_i';
+for i = 1:nstates
+  string = sprintf('$\\sigma_{%d}$ & $%.3f$', i, true_model.states{i}.sigma);
+  for length_index = 1:NL
+    string = sprintf('%s & $%.3f$', string, model_store{length_index}.states{i}.sigma);
+  end
+  string = sprintf('%s \\\\', string);
+  disp(string)
+end  
+
+disp('\hline');
+
+disp('\end{tabular}');

### 2015-02-16 matlab/validation/test_uncertainty_prediction2.m

@@ -1,0 +1,113 @@
+% Test prediction of uncertainty.
+
+generate_bhmm = true;
+if generate_bhmm
+  clear;
+
+  % Add BHMM paths.
+  addpath ../../matlab/bhmm
+  javaaddpath ../../matlab/bhmm
+
+  load true_uncertainties.mat;
+
+  % Generate true model and initial observation data for first trace.
+  [true_model, data] = generate_test_model(observation_length);
+  nstates = true_model.nstates;
+
+  % Build MLHMM.
+  mlhmm_model = mlhmm(data, true_model.nstates, options);
+
+  % Generate Bayesian HMM and store predicted models.
+  nburnin = 100;
+  nmodels = 10000;
+  bhmm_models = bhmm(data, mlhmm_model, nburnin+nmodels, options);    
+  
+  % Discard initial models to burn-in.
+  bhmm_models = bhmm_models(nburnin+1:end);
+
+  % Remove hidden state sequences.
+  clear pruned_models;
+  for i = 1:nmodels
+    model = bhmm_models(i);
+    model = rmfield(model, 'state_trajectories');
+    pruned_models(i) = model;
+  end
+  bhmm_models = pruned_models;
+  clear pruned_models;
+  
+  save uncertainty_prediction_models data bhmm_models;
+else
+  load uncertainty_prediction_models;
+end
+
+% Compute original log probabilities of sampled models.
+log_sampled_probabilities = zeros(nmodels,1);
+for model_index = 1:nmodels
+  log_sampled_probabilities(model_index) = compute_log_posterior(bhmm_models(model_index), data, options, false);
+end    
+
+% Define property to examine.
+%property = @(model) model.Tij(1,2) / model.Tij(1,3);
+property = @(model) model.Tij(1,2);
+%property = @(model) model.Pi(2);
+%property = @(model) 1.0 / (1 - model.Tij(2,2)); % lifetime of intermediate state
+
+% Compute property for all models.
+property_n = zeros(nmodels,1); % storage for property
+for n = 1:nmodels
+  property_n(n) = property(bhmm_models(n));
+end
+
+% Predict uncertainty with BHMM modeling.
+predicted_uncertainty = zeros(max_traces,1);
+predicted_uncertainty_error = zeros(max_traces,1);
+for ntraces = 1:max_traces
+  % Sample a certain number of synethetic experiments for prediction.
+  mlhmm_estimates = zeros(nsynthetic,1); 
+  bhmm_predicted_estimates = zeros(nsynthetic,1); 
[…]

### 2015-02-16 manuscripts/bayesian-hmm-method/bhmm.tex
Updated manuscript font to sans-serif. Updated Christian Kaiser affiliation.

@@
 
-\usepackage{palatino}
+%\usepackage{palatino}
+% Change to a sans serif font.
+\usepackage{sourcesanspro}
+\renewcommand*\familydefault{\sfdefault} %% Only if the base font of the document is to be sans serif
+\usepackage[T1]{fontenc}
+%\usepackage[font=sf,justification=justified]{caption}
+\usepackage[font=sf]{floatrow}
+
+% Rework captions to use sans serif font.
+\makeatletter
+\renewcommand\@make@capt@title[2]{%
+ \@ifx@empty\float@link{\@firstofone}{\expandafter\href\expandafter{\float@link}}%
+  {\textbf{#1}}\sf\@caption@fignum@sep#2\quad
+}%
+\makeatother
+
 \usepackage{amsmath}
@@
 %% TITLE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Bayesian hidden Markov model analysis of single-molecule force spectroscopy: Characterizing kinetics under measurement uncertainty}
+\title{Bayesian hidden Markov model analysis of single-molecule force spectroscopy:\\
+Characterizing kinetics under measurement uncertainty}
 %\title{Characterizing the uncertainty in equilibrium kinetics from single-molecule force spectroscopy measurements using Bayesian hidden Markov models}
@@
 \author{Christian M.~Kaiser}
- \email{cmkaiser@berkeley.edu}
- \affiliation{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720, USA}
- \affiliation{Department of Physics, University of California, Berkeley, CA 94720, USA}
+ %\email{cmkaiser@berkeley.edu}
+ \email{kaiser.jhu.bio@gmail.com}
+% \affiliation{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720, USA}
+ %\affiliation{Department of Physics, University of California, Berkeley, CA 94720, USA}
+ \affiliation{Department of Biology, Krieger School of Arts and Sciences, Johns Hopkins University, Baltimore, MD 21218}
 

### 2015-02-16 manuscripts/bayesian-hmm-method/bhmm.tex
Corrections by FN

@@
 Recent advances in biophysical measurement have led to an unprecedented ability to monitor the dynamics of single biological macromolecules, such as proteins and nucleic acids~\cite{ritort:j-phys:2006:single-molecule-review}.
-As a new approach to probing the behavior of biological macromolecules, these experiments promise to change the way we study folding, dynamics, catalysis, association, transcription, translation, and motility, providing otherwise-inaccessible information about microscopic kinetics, energetics, mechanism, and the stochastic heterogeneity inherent in these processes.
+%As a new approach to probing the behavior of biological macromolecules, these experiments promise to change the way we study folding, dynamics, catalysis, association, transcription, translation, and motility, providing otherwise-inaccessible information about microscopic kinetics, energetics, mechanism, and the stochastic heterogeneity inherent in these processes.
 Advances in instrumentation for optical force spectroscopy in particular have produced instruments of extraordinary stability, precision, and temporal resolution~\cite{block:rev-sci-instrum:2004:optical-trapping,bustamante:annu-rev-biochem:2008:optical-tweezers} that have already demonstrated great utility in the study of biomolecules in the presence of externally perturbative forces~\cite{bustamante:nature:2003:dna-mechanics,block:curr-opin-chem-biol:2008:rna,bustamante:cell:2011:central-dogma}.
@@
 {\color{red}The instrument can generally be operated in several modes: a force ramp mode, in which the trap is translated rapidly enough to potentially carry the system {\color{red}transiently} out of equilibrium; a passive mode, in which the trap is held fixed and the equilibrium fluctuations of the system are observed; and a constant force-feedback mode, in which the trap is continually repositioned to maintain a specified average constant force on the fiber.}
-Here, we concern ourselves with the latter two classes of experiment, though nonequilibrium experiments remain an exciting topic of active research~\cite{ritort:adv-chem-phys:2008:nonequilibrium-review}.
+Here, we concern ourselves with passive-mode force probe experiments, though both nonequilibrium experiments and constant force-feedback experiments and their analysis remain an exciting topic of active research~\cite{ritort:adv-chem-phys:2008:nonequilibrium-review}
+{\color{magenta}FN: add additional citations for both modes / analyses}.
 
@@
 {\color{red} As a result, a histogram of the observed temporal trace can have strongly overlapping state distributions, as in Fig.~\ref{figure:p5ab-observed-trajectory}, right.}
-Attempting to separate these states by simply dividing the observed force or extension range into regions, following current practice~\cite{woodside:science:2006:dna-hairpin-optical-trap,ritort:biophys-j:2011:short-handles}, can often lead to a high degree of state mis-assignment that results in the estimated rate constants and state distributions containing a significant amount of error~\cite{chodera:rate-theory} (see \emph{Supplementary Material: Comparison with threshold model}).
+Attempting to separate these states by simply dividing the observed force or extension range into regions, following current practice~\cite{woodside:science:2006:dna-hairpin-optical-trap,ritort:biophys-j:2011:short-handles}, can often lead to a high degree of state mis-assignment that results in the estimated rate constants and state distributions containing a significant amount of error~\cite{chodera:rate-theory,PrinzChoderaNoe_PRX14_RateTheory} (see \emph{Supplementary Material: Comparison with threshold model}).
 
-Hidden Markov models (HMMs)~\cite{rabiner:ieee-proceedings:1989:hmm-tutorial}, which use temporal information in addition to the instantaneous value of the observable (force or extension) to determine which conformational states the system has visited during the experiment, have provided an effective solution to the hidden state problem in many other classes of single-molecule experiments, such as ion channel currents~\cite{becker:pfluegers-arch:1994:ion-channel-hmm,sachs:biophys-j:2000:ion-channel-hmm,degunst-kuensch-schouten:j-am-stat-assoc:2001:ion-channel-hmm,qin:biophys-j:2004:ion-channel-hmm}, single-molecule FRET~\cite{andrec-levy-talaga:jpca:2003:photon-arrival-hmm,mckinney-joo-ha:biophys-j:2006:hmm-fret,lee:jpcb:2009:hmm-fret,scherer-dinner:jpcb:2009:rna-fret-hmm,gopich-szabo:jpcb:2009:decoding-fret}, and the stepping of motor proteins~\cite{smith:biophys-j:2001:hmm-actomyosin,sachs:biophys-j:2006:motor-hmm,sigworth:biophys-j:2010:hmm-motors}.
+Hidden Markov models (HMMs)~\cite{rabiner:ieee-proceedings:1989:hmm-tutorial}, which use temporal information in addition to the instantaneous value of the observable (force or extension) to determine which conformational states the system has visited during the experiment, have provided an effective solution to the hidden state problem in many other classes of single-molecule experiments, such as ion channel currents~\cite{becker:pfluegers-arch:1994:ion-channel-hmm,sachs:biophys-j:2000:ion-channel-hmm,degunst-kuensch-schouten:j-am-stat-assoc:2001:ion-channel-hmm,qin:biophys-j:2004:ion-channel-hmm}, single-molecule FRET~\cite{andrec-levy-talaga:jpca:2003:photon-arrival-hmm,mckinney-joo-ha:biophys-j:2006:hmm-fret,lee:jpcb:2009:hmm-fret,scherer-dinner:jpcb:2009:rna-fret-hmm,gopich-szabo:jpcb:2009:decoding-fret,KellerNoe_JACS14_HMM-FRET}, and the stepping of motor proteins~\cite{smith:biophys-j:2001:hmm-actomyosin,sachs:biophys-j:2006:motor-hmm,sigworth:biophys-j:2010:hmm-motors}.
 In applying hidden Markov modeling to the analysis of single-molecule force spectroscopy data, the observed force or extension trace is assumed to come from a realization of an underlying Markov chain, where the system makes history-independent transitions among a set of discrete conformational states with probabilities governed by a transition or rate matrix.
-Data, in the form of force or bead-to-bead extension measurements, is sampled at an interval that ensures that sequential observations satisfy the Markov property of history-independence, though the appropriate interval depends on the properties of the experimental configuration.
+%Data, in the form of force or bead-to-bead extension measurements, is sampled at an interval that ensures that sequential observations satisfy the Markov property of history-independence, though the appropriate interval depends on the properties of the experimental configuration.
 Under a given set of external force conditions, each state has a distribution of forces or extensions associated with it.
-Given observed timeseries data for forces or extensions, the maximum likelihood estimate (MLE) of the model parameters (transition rates and state force or extension distributions) and sequence of hidden states corresponding to the observed data {\color{red} (used to color the points in Fig.~\ref{figure:p5ab-observed-trajectory})} can be determined by standard methods~\cite{baum:1970:ann-math-statist:baum-welch,viterbi:1967:IEEE-trans-info-theory:viterbi-algorithm}, as demonstrated in recent work~\cite{kruithof-vannoort:biophys-j:2009:hmm-force}.
 
@@
 
+Given observed timeseries data for forces or extensions, the maximum likelihood estimate (MLE) of the model parameters (transition rates and state force or extension distributions) and sequence of hidden states corresponding to the observed data {\color{red} (used to color the points in Fig.~\ref{figure:p5ab-observed-trajectory})} can be determined by standard methods~\cite{baum:1970:ann-math-statist:baum-welch,viterbi:1967:IEEE-trans-info-theory:viterbi-algorithm}, as demonstrated in recent work~\cite{kruithof-vannoort:biophys-j:2009:hmm-force}.
 Unfortunately, this approach has a number of significant drawbacks.
-Due to technical limitations, experiments often suffer from limited statistics---the events of interest (transitions between states or visits to rare states) may occur only a few times during the course of the measurement, and data for additional fibers is time-consuming to collect.
-As a result, while the MLE yields the most likely set of model parameters, there may be enormous uncertainty in some of these parameters, and the uncertainty in multiple parameters may be correlated in complex nonlinear ways.
-While methods exist for estimating the standard error or confidence intervals from 
-%\mbox{MLHMMs}
-{\color{red} hidden Markov models}
-~\cite{aittokallio-uusipaikka:technical-report:2000:hmm-standard-error}, these schemes can be prohibitively costly for long traces, and may still significantly underestimate the statistical error for short traces due to the normally-distributed error approximation inherent in the approach.
-{\color{blue}[JDC: Should we say something about bootstrap here since we will compare to it later?]}
-The high cost (both in terms of instrument and experimenter time) of collecting additional data also means that it is not a simple task to judge how much data need be collected to test a particular hypothesis in a statistically meaningful way.
+%Due to technical limitations, experiments often suffer from limited statistics---the events of interest (transitions between states or visits to rare states) may occur only a few times during the course of the measurement, and data for additional fibers is time-consuming to collect.
+%As a result, while the MLE yields the most likely set of model parameters, there may be 
+{\color{magenta}The parameters estimated by the MLE might be subject to enormous uncertainty. 
+Methods for estimating the standard error or confidence intervals from hidden Markov models~\cite{aittokallio-uusipaikka:technical-report:2000:hmm-standard-error} make a number of approximations that can lead to a significant underestimation of the error.}
+%, these schemes can be prohibitively costly for long traces, and may still significantly underestimate the statistical error for short traces due to the normally-distributed error approximation inherent in the approach.
+%{\color{blue}[JDC: Should we say something about bootstrap here since we will compare to it later?]}
+%The high cost (both in terms of instrument and experimenter time) of collecting additional data also means that it is not a simple task to judge how much data need be collected to test a particular hypothesis in a statistically meaningful way.
 Worse yet, the standard algorithms employed to find the MLE may not even find the true maximum likelihood solution, instead converging to a local maximum in likelihood that is far from optimal~\cite{merialdo:ieee-trans-speech-audio-proc:1993:baum-welch-localit}.
@@
 
-In the following sections, we briefly describe the theory behind maximum-likelihood hidden Markov models of force spectroscopy data (\emph{Theory}), the validation of our methodology on synthetic data for which the computed confidence intervals can be checked for correctness (\emph{Validation}), and our specific applications to real force spectroscopy data (\emph{Applications}).
-We follow with a discussion of future extensions of this approach, and include a complete description of the computational implementation in \emph{Algorithms}.
+%In the following sections, we briefly describe the theory behind maximum-likelihood hidden Markov models of force spectroscopy data (\emph{Theory}), the validation of our methodology on synthetic data for which the computed confidence intervals can be checked for correctness (\emph{Validation}), and our specific applications to real force spectroscopy data (\emph{Applications}).
+%We follow with a discussion of future extensions of this approach, and include a complete description of the computational implementation in \emph{Algorithms}.
 A Matlab implementation of the approach described here is also made freely available online [\url{http://github.com/choderalab/bhmm}].
@@
 
-Suppose the temporal behavior of some observable $O(x)$ that is a function of molecular configuration $x$---here, generally force or molecular extension---is observed at temporal intervals $\Delta t$ to produce a timeseries $o_t$, where $t = 0, 1, \ldots, L$.
+Suppose the temporal behavior of some observable $O(x)$ that is a function of molecular configuration $x$---here, the force or molecular extension---is observed at temporal intervals $\Delta t$ to produce a timeseries $o_t$, where $t = 0, 1, \ldots, L$.
 An instantaneous observation $o_t$ does not necessarily contain enough information to unambiguously identify the current conformational state the molecule occupies; to infer the hidden state, we must also make use of the temporal information in the observed trace.
@@
 State transitions $(s^{}_{t-1} \rightarrow s^{}_{t})$ are governed by the discrete transition probability $T_{s^{}_{t-1} s^{}_{t}}$.
-The Markov property of HMMs prescribes that the probability that a system originally in state $i$ at time $t$ is later found in state $j$ at time $t+1$ is dependent only on knowledge of the state $i$, and given by the corresponding matrix element $T_{ij}$ of the (row-stochastic) transition matrix $\bfm{T}$.
-Alternatively, one could instead use the rate matrix $\bfm{K}$, related to the transition matrix $\bfm{T}$ through the equation $\bfm{T} = e^{\bfm{K} \Delta t}$.  
-If the processes described by $\bfm{T}$ or $\bfm{K}$ are slow compared to the observation interval $\Delta t$, then we can easily estimate the rate matrix from the associated transition matrix in a way that avoids the matrix logarithm, through the expansion $\bfm{K} \approx (\bfm{T} - \bfm{I}) / \Delta t$, where $\bfm{I}$ denotes the $M \times M$ identity matrix.
+The Markov property of HMMs prescribes that the probability that a system originally in state $i$ at time $t$ is later found in state $j$ at time $t+\Delta t$ is dependent only on knowledge of the state $i$, and given by the corresponding matrix element $T_{ij}$ of the (row-stochastic) transition matrix $\bfm{T}$.
+{\color{magenta}The choice of $\Delta t$ is critical for the accuracy of the model \cite{NoeEtAl_PMMHMM_JCP13} --- $\Delta t$ must
+be chosen to reside in a time-scale gap between the $M$ slow relaxation processes that are resolved and faster relaxation
+processes that are not resolved by the model.}
+% not sure if we need this discussion here. We can still say how a rate matrix can be computed at the end.
+%Alternatively, one could instead use the rate matrix $\bfm{K}$, related to the transition matrix $\bfm{T}$ through the equation $\bfm{T} = e^{\bfm{K} \Delta t}$.  
+%If the processes described by $\bfm{T}$ or $\bfm{K}$ are slow compared to the observation interval $\Delta t$, then we can easily estimate the rate matrix from the associated transition matrix in a way that avoids the matrix logarithm, through the expansion $\bfm{K} \approx (\bfm{T} - \bfm{I}) / \Delta t$, where $\bfm{I}$ denotes the $M \times M$ identity matrix.
 
-The probabilistic ``emission'' of observables from each state $(s^{}_{t} \rightarrow o^{}_{t})$ is governed by the continuous emission probability $\varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}})$, parametrized by observable emission parameters $\bfm{e}$.
+{\color{magenta}The dependence of the experimental observable on the hidden state}
+%The probabilistic ``emission'' of observables from each state $(s^{}_{t} \rightarrow o^{}_{t})$ 
+is governed by the emission probability $\varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}})$, parametrized by observable emission parameters $\bfm{e}$.
 For example, in the force spectroscopy applications described here, $\varphi(o \mid \bfm{e}_s)$ is taken to be a univariate normal (Gaussian) distribution parameterized by a mean $\mu$ and variance $\sigma^2$ that characterize each state, such that $\bfm{e}_i \equiv \{\mu_i, \sigma^2_i\}$.
@@
 
-Instead of simply determining the model that maximizes the likelihood of observing the data $\bfm{O}$ given the model parameters $\bfm{\Theta}$, we can make use of Bayes' theorem to compute the posterior distribution of model parameters given the observed data:
+Instead of simply determining the model that maximizes the likelihood of observing the data $\bfm{O}$ given the model parameters $\bfm{\Theta}$, {\color{magenta} we here propose to sample to full posterior distribution of model parameters given the observed data.
+Towards this end, we emply Bayes' theorem to compute the posterior distribution:}
 \begin{eqnarray}
@@
 When only the model parameters $\bfm{\Theta} \equiv \{ \bfm{T}, \bfm{E} \}$ or the hidden state histories $\bfm{S}$ are of interest, we can simply marginalize out the uninteresting variables by sampling from the augmented joint posterior for $\{\bfm{T}, \bfm{E}, \bfm{S}\}$ and examine only the variables of interest.
-In addition, the structure of the Gibbs sampling scheme above allows individual components (such as the observable distribution model $\varphi(o \mid \bfm{e})$ or transition probability matrix $\bfm{T}$) to be modified without affecting the structure of the remainder of the calculation.
+%In addition, the structure of the Gibbs sampling scheme above allows individual components (such as the observable distribution model $\varphi(o \mid \bfm{e})$ or transition probability matrix $\bfm{T}$) to be modified without affecting the structure of the remainder of the calculation.
 
@@
 The appropriate observation interval (1 ms) for BHMM analysis was determined by examination of force autocorrelation functions, which suggested that intrastate motion had completely decayed by 1 ms (see \emph{Supplementary Material: Choice of observation interval}).
+{\color{magenta}[FN: Update the selection of the HMM lag time.
+I suggest fixing the lag time $\tau$ first, running a maximum-likelihood HMM with N states, and then select $M<N$ based on how many relaxation timescales are larger than $\tau$]}
 The observed 50 kHz force traces were therefore subsampled  to 1 kHz for analysis by retaining every 50th sample.
@@
 
+{\color{magenta}[Describe how/why you selected three over two states in this model]}
 Table~\ref{table:p5ab-bhmm-confidence-intervals} lists the BHMM posterior means and confidence intervals characterizing the three-state model extracted from this single 60 s observed force trace.
@@
 \color{black}
-FN acknowledges DFG Grant 825/2.
+FN acknowledges funding by DFG Grant 825/2 and ERC starting grant ``pcCell''.
 This work was supported in part by a grant from the NSF (SM).
@@
 Note that the equilibrium probability vector $\bfv{\pi}$ computed during the Gaussian mixture model fitting is not respected during this step.
+{\color{magenta}[FN: clarify which of the two $\pi$-vectors is used in the subsequent MLHMM]}
 
@@
 The HMM parameters $\bfm{\Theta} \equiv \{ \bfm{T}, \bfm{E} \}$ are fit to the observed data $\bfm{O}$ through use of the expectation-maximization (EM) algorithm~\cite{dempster-laird-rubin:1977:j-royal-statist-soc-b:em-algorithm}.
-This is an iterative procedure, where the model parameters are subsequently refined through successive iterations.
+This is an iterative procedure, where the model parameters are subsequently refined through successive iterations, according to the following
+scheme:
+\color{magenta}
+\begin{eqnarray}
+\mathbf{S'} &=& \arg \max_S P(S \mid \mathbf{T}, \mathbf{E}, \mathbf{O})\\
+\mathbf{T'} &=& \arg \max_T P(T \mid \mathbf{S}, \mathbf{O})\\
+\mathbf{E'} &=& \arg \max_E P(E \mid \mathbf{S}, \mathbf{O})
+\end{eqnarray}
+\color{black}
 The initial HMM is usually quick to compute, and can give the experimenter a rough idea of the model parameters, as well as providing a useful starting point for sampling models from the Bayesian posterior.
@@
 Using the adjusted count matrix $\bfm{C}$, a Metropolis-Hastings Monte Carlo procedure~\cite{metropolis-hastings} is used to update the matrix and produce a new sample from $P(\bfm{T}' \mid \bfm{S}')$.
-Two move types are attempted, selected with equal probability, and 1000 moves are attempted to generate a new sample $\bfm{T}'$ that is approximately uncorrelated from the previous $\bfm{T}$.
+Two move types are attempted, selected with equal probability, and 1000 trial moves are attempted to generate a new sample $\bfm{T}'$ that is approximately uncorrelated from the previous $\bfm{T}$.
 Prior to starting the Monte Carlo procedure, the vector of equilibrium probabilities for all states $\bfv{\pi}$ is computed according to 
@@
 This behavior might be undesirable, as the states may switch labels during the posterior sampling procedure; this will require any analysis of the models sampled from the posterior to account for the possible permutation symmetry in the states.
-On the other hand, breaking this symmetry (e.g., by enforcing an ordering on the state mean observables) can artificially restrict the confidence intervals of the states, which might additionally complicate data analysis.
+{\color{magenta}Here we choose to order the states according to their value in the experimental observable (force or extension), thus removing
+the ambiguity due to exchangeable state labels. Note that this practice can artificially restrict the confidence intervals of the states when 
+the distributions of the experimental observable overlap strongly in neighboring states}
+%On the other hand, breaking this symmetry (e.g., by enforcing an ordering on the state mean observables) can artificially restrict the confidence intervals of the states, which might additionally complicate data analysis.
 

### 2015-02-16 manuscripts/bayesian-hmm-method/bhmm.tex
Edits to abstract.

@@
 While maximum-likelihood schemes such as hidden Markov models have solved this problem for other classes of single-molecule experiments by using temporal information to aid in the inference of a sequence of distinct conformational states, these methods do not give a clear picture of how precisely the model parameters are determined by the data due to instrument noise and finite-sample statistics, both significant problems in force spectroscopy.
-We solve this problem through a Bayesian extension that allows the experimental uncertainties to be directly quantified, and build in detailed balance to further reduce uncertainty through physical constraints.
+{\color{blue}We solve this problem through a Bayesian extension that allows the experimental uncertainties to be directly quantified, and allow uncertainty to be further reduced through the imposition of physical priors such as the requirement that dynamics obey detailed balance.}
 We illustrate the utility of this approach in characterizing the {\color{red}observed multistate kinetics of two molecules in a stationary optical trap: an RNA hairpin that exhibits three-state behavior and a protein system (apomyglobin at pH 5) that exhibits two-state behavior with highly overlapping force distributions characterizing each state.}
 
-%\emph{Keywords: force spectroscopy; optical tweezers; single-molecule experiments; hidden Markov model (HMM); Bayesian analysis; statistical error; statistical uncertainty}
+\emph{Keywords: force spectroscopy; optical tweezers; single-molecule experiments; hidden Markov model (HMM); Bayesian analysis; statistical error; statistical uncertainty}
 

### 2015-02-16 manuscripts/bayesian-hmm-method/bhmm.tex
revised Fig. 2
Added reference; clean up of bibtex references.

@@
 {\color{red}The instrument can generally be operated in several modes: a force ramp mode, in which the trap is translated rapidly enough to potentially carry the system {\color{red}transiently} out of equilibrium; a passive mode, in which the trap is held fixed and the equilibrium fluctuations of the system are observed; and a constant force-feedback mode, in which the trap is continually repositioned to maintain a specified average constant force on the fiber.}
-Here, we concern ourselves with passive-mode force probe experiments, though both nonequilibrium experiments and constant force-feedback experiments and their analysis remain an exciting topic of active research~\cite{ritort:adv-chem-phys:2008:nonequilibrium-review}
+Here, we concern ourselves with passive-mode force probe experiments, though both nonequilibrium experiments and constant force-feedback experiments and their analysis remain an exciting topic of active research~\cite{ritort:adv-chem-phys:2008:nonequilibrium-review,elms:biophys-j:2012:force-feedback}
 {\color{magenta}FN: add additional citations for both modes / analyses}.

### 2015-02-16 manuscripts/bayesian-hmm-method/bhmm.tex
Fix to make figure captions in sans serif.

@@
  \@ifx@empty\float@link{\@firstofone}{\expandafter\href\expandafter{\float@link}}%
-  {\textbf{#1}}\sf\@caption@fignum@sep#2\quad
+  {\sf\textbf{#1}}\sf\@caption@fignum@sep#2\quad
 }%
@@
 \noindent
-\resizebox{\textwidth}{!}{\includegraphics{algorithm-workflow/Fig_ProbabilitySpace_trimmed.pdf}}
+\resizebox{0.8\textwidth}{!}{\includegraphics{algorithm-workflow/Fig_ProbabilitySpace.pdf}}
 \caption{{\bf Illustration of the Bayesian hidden Markov model (BHMM) sampling algorithm.}

### 2015-02-16 manuscripts/bayesian-hmm-method/bhmm.tex
Updated figure 2 caption.

@@
 \caption{{\bf Illustration of the Bayesian hidden Markov model (BHMM) sampling algorithm.}
-During each BHMM update, a modified forward-backward algorithm is used to sample a new hidden state trace $\mathbf{S}$, which is then used to generate a new sample of the reversible transition matrix $\mathbf{T}$ using MCMC sampling, followed by sapling of the state emission probabilities $\mathbf{E}$.
-The cycle then repeats for each subsequent iteration, which together sample from the Bayesian posterior of the HMM and fully characterize the uncertainty in the kinetic model given the data.
+During each BHMM sampling iteration, a modified forward-backward algorithm is used to sample a new hidden state trace $\mathbf{S}$, which is then used to generate a new sample of the reversible transition matrix $\mathbf{T}$ using MCMC sampling, followed by sampling new state emission probabilities $\mathbf{E}$.
+Collectively, these updates sample from the Bayesian posterior of the HMM and fully characterize the uncertainty in the kinetic model given the data.
 \label{figure:bhmm-workflow}}

### 2015-02-16 manuscripts/bayesian-hmm-method/bhmm.tex
Updated probability P(...) representation to be consistent throughout.

@@
 \newcommand{\expect}[1]{\left \langle #1 \right \rangle}                % <.> for denoting expectations over realizations of an experiment or thermal averages
+
+% Probability representation
+\renewcommand{\Pr}{\mathrm{P}}
 
@@
 %o^{}_t &\sim& \varphi(\cdot \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-process}
-\mathbb{P}(s^{}_0) &=& \rho^{}_{s^{}_0} \nonumber \\
-\mathbb{P}(s^{}_t \mid s^{}_{t-1}, \bfm{T}) &=& T_{s^{}_{t-1} s^{}_{t}} , \:\: t \ge 1 \nonumber \\
-\mathbb{P}(o^{}_t \mid s^{}_t, \bfm{e}_{s^{}_{t}}) &=& \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-process} .
+\Pr(s^{}_0) &=& \rho^{}_{s^{}_0} \nonumber \\
+\Pr(s^{}_t \mid s^{}_{t-1}, \bfm{T}) &=& T_{s^{}_{t-1} s^{}_{t}} , \:\: t \ge 1 \nonumber \\
+\Pr(o^{}_t \mid s^{}_t, \bfm{e}_{s^{}_{t}}) &=& \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-process} .
 \end{eqnarray}
@@
 \begin{eqnarray}
-P(\bfm{O} \mid \bfm{\Theta}) &=& \sum_{\bfm{S}}  \rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{t-1} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-likelihood} ,
+\Pr(\bfm{O} \mid \bfm{\Theta}) &=& \sum_{\bfm{S}}  \rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{t-1} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-likelihood} ,
 \end{eqnarray}
@@
 \end{eqnarray}
-If multiple independent traces $\{o_t\}$ are available, the probability $P(\bfm{O} \mid \bfm{\Theta})$ is simply the product of Eq.~\ref{equation:hmm-likelihood} for the independent traces.
+If multiple independent traces $\{o_t\}$ are available, the probability $\Pr(\bfm{O} \mid \bfm{\Theta})$ is simply the product of Eq.~\ref{equation:hmm-likelihood} for the independent traces.
 
@@
 \begin{eqnarray}
-\hat{\bfm{\Theta}} &=& \arg\max_\bfm{\Theta} P(\bfm{O} \mid \bfm{\Theta}) ,
+\hat{\bfm{\Theta}} &=& \arg\max_\bfm{\Theta} \Pr(\bfm{O} \mid \bfm{\Theta}) ,
 \end{eqnarray}
@@
 \begin{eqnarray}
-\hat{\bfm{S}} &=& \arg\max_\bfm{S} P(\bfm{S} \mid \bfm{O}, \hat{\bfm{\Theta}}) .
+\hat{\bfm{S}} &=& \arg\max_\bfm{S} \Pr(\bfm{S} \mid \bfm{O}, \hat{\bfm{\Theta}}) .
 \end{eqnarray}
@@
 \begin{eqnarray}
-P(\bfm{\Theta} \mid \bfm{O}) &\propto& P(\bfm{O} \mid \bfm{\Theta}) P(\bfm{\Theta}) \label{equation:bayes-theorem}.
+\Pr(\bfm{\Theta} \mid \bfm{O}) &\propto& \Pr(\bfm{O} \mid \bfm{\Theta}) \Pr(\bfm{\Theta}) \label{equation:bayes-theorem}.
 \end{eqnarray}
-Here, $P(\bfm{\Theta})$ denotes a prior distribution that encodes any \emph{a priori} information we may have about the model parameters $\bfm{\Theta}$.
+Here, $\Pr(\bfm{\Theta})$ denotes a prior distribution that encodes any \emph{a priori} information we may have about the model parameters $\bfm{\Theta}$.
 This prior information might include, for example, physical constraints (such as ensuring the transition matrix satisfies detailed balance) or prior rounds of inference from other independent experiments.
@@
 \begin{eqnarray}
-P(\bfm{\Theta} \mid \bfm{O}) &\propto& P(\bfm{\Theta}) \sum_{\bfm{S}}\rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{t-1} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}})  \label{equation:full-posterior} .
+\Pr(\bfm{\Theta} \mid \bfm{O}) &\propto& \Pr(\bfm{\Theta}) \sum_{\bfm{S}}\rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{t-1} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}})  \label{equation:full-posterior} .
 \end{eqnarray}
 Drawing samples of $\bfm{\Theta}$ from this distribution will, in principle, allow the confidence with which individual parameters and combinations thereof are known, given the data (and subject to the validity of the model of Eq.~\ref{equation:hmm-process} in correctly representing the process by which the observed data is generated).
-While the posterior $P(\bfm{\Theta}|\bfm{O})$ is complex, we could in principle use a Markov chain Monte Carlo (MCMC) approach~\cite{jun-s-liu:mcmc} to sample it.
+While the posterior $\Pr(\bfm{\Theta}|\bfm{O})$ is complex, we could in principle use a Markov chain Monte Carlo (MCMC) approach~\cite{jun-s-liu:mcmc} to sample it.
 In its current form, however, this would be extremely expensive due to the sum over all hidden state histories $\bfm{S}$ appearing in ratios involving Eq.~\ref{equation:full-posterior}.
@@
 \begin{eqnarray}
-P(\bfm{\Theta}, \bfm{S} \mid \bfm{O}) &\propto& \left[ \rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{t-1} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \right] P(\bfm{\Theta}) .
+\Pr(\bfm{\Theta}, \bfm{S} \mid \bfm{O}) &\propto& \left[ \rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{t-1} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \right] \Pr(\bfm{\Theta}) .
 \nonumber \\\label{equation:augmented-posterior}
@@
 
-If we presume the prior is separable, such that $P(\bfm{\Theta}) \equiv P(\bfm{T}) P(\bfm{E})$, we can sample from the augmented posterior (Eq.~\ref{equation:augmented-posterior}) using the framework of Gibbs sampling~\cite{jun-s-liu:mcmc}, in which the augmented model parameters are updated by cycles of sampling from conditional distributions,
+If we presume the prior is separable, such that $\Pr(\bfm{\Theta}) \equiv \Pr(\bfm{T}) \Pr(\bfm{E})$, we can sample from the augmented posterior (Eq.~\ref{equation:augmented-posterior}) using the framework of Gibbs sampling~\cite{jun-s-liu:mcmc}, in which the augmented model parameters are updated by cycles of sampling from conditional distributions,
 \begin{eqnarray}
-\bfm{S}' &\sim& P(\bfm{S} \mid \bfm{T}, \bfm{E}, \bfm{O}) \propto \rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{t-1} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \nonumber \\
-\bfm{T}' &\sim& P(\bfm{T} \mid \bfm{E}, \bfm{S}', \bfm{O}) = P(\bfm{T} \mid \bfm{S}') \propto P(\bfm{T}) \prod_{t=1}^{L^{}} T_{{s'}^{}_{t-1} {s'}^{}_{t}} \label{equation:conditional-gibbs} \\
-\bfm{E}' &\sim& P(\bfm{E} \mid \bfm{S}', \bfm{T}', \bfm{O}) = P(\bfm{E} \mid \bfm{S}', \bfm{O}) \propto P(\bfm{E}) \prod_{t=0}^{L^{}} \varphi(o^{}_{t} \mid \bfm{e}_{{s'}^{}_{t}}) \nonumber .
+\bfm{S}' &\sim& \Pr(\bfm{S} \mid \bfm{T}, \bfm{E}, \bfm{O}) \propto \rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{t-1} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \nonumber \\
+\bfm{T}' &\sim& \Pr(\bfm{T} \mid \bfm{E}, \bfm{S}', \bfm{O}) = \Pr(\bfm{T} \mid \bfm{S}') \propto \Pr(\bfm{T}) \prod_{t=1}^{L^{}} T_{{s'}^{}_{t-1} {s'}^{}_{t}} \label{equation:conditional-gibbs} \\
+\bfm{E}' &\sim& \Pr(\bfm{E} \mid \bfm{S}', \bfm{T}', \bfm{O}) = \Pr(\bfm{E} \mid \bfm{S}', \bfm{O}) \propto \Pr(\bfm{E}) \prod_{t=0}^{L^{}} \varphi(o^{}_{t} \mid \bfm{e}_{{s'}^{}_{t}}) \nonumber .
 \end{eqnarray}
@@
 Other choices for the observable distribution model $\phi(o | \mathbf{e})$ can easily be substituted should they be more appropriate for analyzing particular classes of experiments, and only require substituting an appropriate parameter update step.
-We note that marginal posterior distributions of each mean $P(\mu_i | \bfm{O})$ reflect the statistical uncertainty in how well the mean force or position is determined, and need not correspond to the standard deviation $\sigma_i$, which may be much broader (or narrower, depending on the situation).
+We note that marginal posterior distributions of each mean $\Pr(\mu_i | \bfm{O})$ reflect the statistical uncertainty in how well the mean force or position is determined, and need not correspond to the standard deviation $\sigma_i$, which may be much broader (or narrower, depending on the situation).
 
@@
 To verify that our BHMM posterior sampling scheme reflects the true uncertainty in the model parameters, we can test that the scheme produces the expected estimates for synthetic data generated from a model with known parameters $\bfm{\Theta}^*$.
-Given observed data $\bfm{O}$ generated from $P(\bfm{O} \mid \bfm{\Theta}^*)$, sampling from the posterior $P(\bfm{\Theta} \mid \bfm{O})$ using the scheme described in \color{blue} \emph{Sampling from the posterior of the BHMM} \color{black} will provide us with confidence intervals $[\theta_\mathrm{low}, \theta_\mathrm{high}]$ for a specified confidence interval $\alpha \in [0,1]$.
+Given observed data $\bfm{O}$ generated from $\Pr(\bfm{O} \mid \bfm{\Theta}^*)$, sampling from the posterior $\Pr(\bfm{\Theta} \mid \bfm{O})$ using the scheme described in \color{blue} \emph{Sampling from the posterior of the BHMM} \color{black} will provide us with confidence intervals $[\theta_\mathrm{low}, \theta_\mathrm{high}]$ for a specified confidence interval $\alpha \in [0,1]$.
 If these computed confidence intervals are accurate, we should find that the true model parameter $\theta^*$ lies in the computed confidence interval $[\theta^{(\alpha)}_\mathrm{low}, \theta^{(\alpha)}_\mathrm{high}]$ with probability $\alpha$.
 A 95\% confidence interval ($\alpha = 0.95$), for example, should contain the BHMM estimate in 95\% of such trials.
-This can be tested by generating synthetic observed data $\bfm{O}$ from $P(\bfm{O} \mid \bfm{\Theta}^*)$ and verifying that we find $\theta^* \in [\theta^{(\alpha)}_\mathrm{low}, \theta^{(\alpha)}_\mathrm{high}]$ in a fraction $\alpha$ of these synthetic experiments.
+This can be tested by generating synthetic observed data $\bfm{O}$ from $\Pr(\bfm{O} \mid \bfm{\Theta}^*)$ and verifying that we find $\theta^* \in [\theta^{(\alpha)}_\mathrm{low}, \theta^{(\alpha)}_\mathrm{high}]$ in a fraction $\alpha$ of these synthetic experiments.
 
@@
 
-As a more rigorous test, we sampled 50 random models from the prior $P(\bfm{\Theta})$ with two to six states, generated a 10 000 observation synthetic trajectory for each, and accumulated statistics on the observed fraction of time the true model parameters were within the BHMM confidence intervals for various values of the confidence interval width $\alpha$.
+As a more rigorous test, we sampled 50 random models from the prior $\Pr(\bfm{\Theta})$ with two to six states, generated a 10 000 observation synthetic trajectory for each, and accumulated statistics on the observed fraction of time the true model parameters were within the BHMM confidence intervals for various values of the confidence interval width $\alpha$.
 The results of this test are depicted in \emph{Supplementary Figure 1}. %~\ref{figure:validation-confidence-intervals}.
@@
 
-To initialize either computation of the MLHMM or sampling from the posterior for the BHMM, an initial model that respects any constraints imposed in the model prior $P(\bfm{\Theta})$ must be selected.
+To initialize either computation of the MLHMM or sampling from the posterior for the BHMM, an initial model that respects any constraints imposed in the model prior $\Pr(\bfm{\Theta})$ must be selected.
 Here, we employ a Gaussian observable distribution model for $\varphi(o \mid \bfm{e})$ (Eq.~\ref{equation:gaussian-observable}) and enforce that the transition matrix $\bfm{T}$ satisfy detailed balance.
@@
 \begin{eqnarray}
-P(\bfm{O} \mid \bfv{\pi}, \bfm{E}) &=& \prod_{t=0}^{L^{}} \sum_{m=1}^M \pi_m \varphi(o^{}_{t} \mid \mu_m, \sigma^2_m) ,
+\Pr(\bfm{O} \mid \bfv{\pi}, \bfm{E}) &=& \prod_{t=0}^{L^{}} \sum_{m=1}^M \pi_m \varphi(o^{}_{t} \mid \mu_m, \sigma^2_m) ,
 \label{equation:gaussian-mixture-likelihood}
@@
 \color{red}
-The iterative procedure described in Algorithm 1 of \cite{noe:jcp:2011:msm-review} is then applied to update the transition matrix in a manner that preserves the detailed balance constraint while maximizing the likelihood function $P(\bfm{T} | \bfm{S})$.
+The iterative procedure described in Algorithm 1 of \cite{noe:jcp:2011:msm-review} is then applied to update the transition matrix in a manner that preserves the detailed balance constraint while maximizing the likelihood function $\Pr(\bfm{T} | \bfm{S})$.
 This algorithm proceeds as follows.
@@
 \begin{eqnarray}
-\mathbf{S'} &=& \arg \max_S P(S \mid \mathbf{T}, \mathbf{E}, \mathbf{O})\\
-\mathbf{T'} &=& \arg \max_T P(T \mid \mathbf{S}, \mathbf{O})\\
-\mathbf{E'} &=& \arg \max_E P(E \mid \mathbf{S}, \mathbf{O})
+\mathbf{S'} &=& \arg \max_S \Pr(S \mid \mathbf{T}, \mathbf{E}, \mathbf{O})\\
+\mathbf{T'} &=& \arg \max_T \Pr(T \mid \mathbf{S}, \mathbf{O})\\
+\mathbf{E'} &=& \arg \max_E \Pr(E \mid \mathbf{S}, \mathbf{O})
 \end{eqnarray}
@@
 \begin{array}{lcl}
-\bfm{S}' \mid \bfm{T}, \bfm{E}, \bfm{O} &\sim& P(\bfm{S}' \mid \bfm{T}, \bfm{E}, \bfm{O}) \nonumber \\
-\bfm{T}' \mid \bfm{S}' &\sim& P(\bfm{T}' \mid \bfm{S}') \nonumber \\
-\bfm{E}' \mid \bfm{S}', \bfm{O} &\sim& P(\bfm{E}' \mid \bfm{S}', \bfm{O}) 
+\bfm{S}' \mid \bfm{T}, \bfm{E}, \bfm{O} &\sim& \Pr(\bfm{S}' \mid \bfm{T}, \bfm{E}, \bfm{O}) \nonumber \\
+\bfm{T}' \mid \bfm{S}' &\sim& \Pr(\bfm{T}' \mid \bfm{S}') \nonumber \\
+\bfm{E}' \mid \bfm{S}', \bfm{O} &\sim& \Pr(\bfm{E}' \mid \bfm{S}', \bfm{O}) 
 \end{array}
@@
 We use a modified form of the Viterbi process to generate an independent sample of the hidden state history $\bfm{S}$ given the transition probabilities $\bfm{T}$, state observable distribution parameters $\bfm{E}$, and observed data $\bfm{O}$.
-Like the Viterbi scheme, a forward recursion is applied to each observation trace $\bfm{o}^{}$, but instead of computing the most likely state history on the reverse pass, a new hidden state history $\bfm{S}$ is drawn from the distribution $P(\bfm{S} \mid \bfm{O}, \bfm{T}, \bfm{E})$.
+Like the Viterbi scheme, a forward recursion is applied to each observation trace $\bfm{o}^{}$, but instead of computing the most likely state history on the reverse pass, a new hidden state history $\bfm{S}$ is drawn from the distribution $\Pr(\bfm{S} \mid \bfm{O}, \bfm{T}, \bfm{E})$.
 The forward recursion uses the same forward algorithm as used in Baum-Welch~\cite{baum:1970:ann-math-statist:baum-welch}, 
@@
 \end{eqnarray}
-In the reverse recursion, we now sample a state sequence by sampling each hidden state from the conditional distribution $s_t \sim P(s_t \mid s_{t+1}, \ldots, s_L)$ starting from $t = L$ and proceeding down to $t = 0$, where the conditional distribution is given by,
+In the reverse recursion, we now sample a state sequence by sampling each hidden state from the conditional distribution $s_t \sim \Pr(s_t \mid s_{t+1}, \ldots, s_L)$ starting from $t = L$ and proceeding down to $t = 0$, where the conditional distribution is given by,
 \begin{eqnarray}
-\lefteqn{P(s_t = i \mid s_{t+1}, \ldots, s_L)} \\
+\lefteqn{\Pr(s_t = i \mid s_{t+1}, \ldots, s_L)} \\
 &\propto& \begin{cases}
@@
 \end{eqnarray} 
-It is straightforward to show the result of these sampling steps reconstitutes the probability distribution $P(\bfm{S} | \bfm{T}, \bfm{E}, \bfm{O})$ (see \emph{Supplementary Material: Proof of state history sampling scheme}).
+It is straightforward to show the result of these sampling steps reconstitutes the probability distribution $\Pr(\bfm{S} | \bfm{T}, \bfm{E}, \bfm{O})$ (see \emph{Supplementary Material: Proof of state history sampling scheme}).
 
@@
 
-%If no detailed balance constraint is used and the prior $P(\bfm{T})$ is Dirichlet in each row of the transition matrix $\bfm{T}$, it is possible to generate an independent sample of the transition matrix from the conditional distribution $P(\bfm{T}' \mid \bfm{S}')$ by sampling each row of the transition matrix from the conjugate Dirichlet posterior using the transition counts from the sampled state sequence $\bfm{S}'$~\cite{noe:jcp:2008:transition-matrix-sampling}.
-{\color{red} If we assume each row of the transition matrix $\bfm{T}$ is independent with Dirichlet prior $P(\bfm{T})$, then it is possible to sample from the transition matrix posterior Dirichlet distribution $P(\bfm{T}' \mid \bfm{S}')$, given the transition counts calculated from the sampled state sequence $\bfm{S}'$~\cite{noe:jcp:2008:transition-matrix-sampling}.}
+%If no detailed balance constraint is used and the prior $\Pr(\bfm{T})$ is Dirichlet in each row of the transition matrix $\bfm{T}$, it is possible to generate an independent sample of the transition matrix from the conditional distribution $\Pr(\bfm{T}' \mid \bfm{S}')$ by sampling each row of the transition matrix from the conjugate Dirichlet posterior using the transition counts from the sampled state sequence $\bfm{S}'$~\cite{noe:jcp:2008:transition-matrix-sampling}.
+{\color{red} If we assume each row of the transition matrix $\bfm{T}$ is independent with Dirichlet prior $\Pr(\bfm{T})$, then it is possible to sample from the transition matrix posterior Dirichlet distribution $\Pr(\bfm{T}' \mid \bfm{S}')$, given the transition counts calculated from the sampled state sequence $\bfm{S}'$~\cite{noe:jcp:2008:transition-matrix-sampling}.}
 However, because physical systems in the absence of energy input through an external driving force should satisfy detailed balance, we make use of this constraint in updating our transition probabilities, since this has been demonstrated to substantially reduce parameter uncertainty in the data-limited regime~\cite{noe:jcp:2008:transition-matrix-sampling}.
@@
 where the Kronecker $\delta_{ij} = 1$ if $i = j$ and zero otherwise, and $\bfm{B} \equiv (b_{ij})$ is a matrix of prior pseudocounts, which we take to be zero following the work of No\'{e} et al.~\cite{noe:pnas:2009:ww-domain}.
-Using the adjusted count matrix $\bfm{C}$, a Metropolis-Hastings Monte Carlo procedure~\cite{metropolis-hastings} is used to update the matrix and produce a new sample from $P(\bfm{T}' \mid \bfm{S}')$.
+Using the adjusted count matrix $\bfm{C}$, a Metropolis-Hastings Monte Carlo procedure~\cite{metropolis-hastings} is used to update the matrix and produce a new sample from $\Pr(\bfm{T}' \mid \bfm{S}')$.
 Two move types are attempted, selected with equal probability, and 1000 trial moves are attempted to generate a new sample $\bfm{T}'$ that is approximately uncorrelated from the previous $\bfm{T}$.
@@
 
-Following the update of the transition matrix $\bfm{T}$, the observable distribution parameters $\bfm{E}$ are updated by sampling $\bfm{E}$ from the conditional probability $P(\bfm{E}' \mid \bfm{S}', \bfm{O})$.
+Following the update of the transition matrix $\bfm{T}$, the observable distribution parameters $\bfm{E}$ are updated by sampling $\bfm{E}$ from the conditional probability $\Pr(\bfm{E}' \mid \bfm{S}', \bfm{O})$.
 The conditional probability for the observable distribution parameters for state $m$, denoted $\bfm{e}_m$, is given in terms of the output model $\varphi(o \mid \bfm{e})$ by Bayes' theorem,
 \begin{eqnarray}
-P(\bfm{E} \mid \bfm{O}, \bfm{S}) &=& \left[\prod_{t=0}^{L^{}} \varphi(o_{t} \mid \bfm{e}_{s^{}_{t}}) \right] P(\bfm{E}) \label{equation:conditional-emission-probability} .
+\Pr(\bfm{E} \mid \bfm{O}, \bfm{S}) &=& \left[\prod_{t=0}^{L^{}} \varphi(o_{t} \mid \bfm{e}_{s^{}_{t}}) \right] \Pr(\bfm{E}) \label{equation:conditional-emission-probability} .
 \end{eqnarray}
 
-An important choice must be made with regards to the prior, $P(\bfm{E})$.
+An important choice must be made with regards to the prior, $\Pr(\bfm{E})$.
 If the prior is chosen to be composed of independent priors for each state, as in
 \begin{eqnarray}
-P(\bfm{E}) &=& \prod_{m=1}^M P(\bfm{e}_m) \label{equation:separable-emission-prior} ,
+\Pr(\bfm{E}) &=& \prod_{m=1}^M \Pr(\bfm{e}_m) \label{equation:separable-emission-prior} ,
 \end{eqnarray}
@@
 \begin{eqnarray}
-P(\bfm{e} \mid \bfm{o}) &=& P(\bfm{o} \mid \bfm{e}) P(\bfm{e}) = \left[ \prod_{n=1}^{N_m} \varphi(o_n \mid \bfm{e}) \right] P(\bfm{e}) .
+\Pr(\bfm{e} \mid \bfm{o}) &=& \Pr(\bfm{o} \mid \bfm{e}) \Pr(\bfm{e}) = \left[ \prod_{n=1}^{N_m} \varphi(o_n \mid \bfm{e}) \right] \Pr(\bfm{e}) .
 \end{eqnarray}
@@
 In the application presented here, we use a Gaussian output model (Eq.~\ref{equation:gaussian-observable}) 
-for the state observable distributions $P(o \mid \bfm{e})$, where $\bfm{e} \equiv \{\mu, \sigma^2\}$, with $\mu$ the state mean observable and $\sigma^2$ the variance (which will include both the distribution of the observable characterizing the state and any broadening from measurement noise).
-Other models (including multidimensional or multimodal observation models) are possible, and require replacing only the observation model $\varphi(o \mid \bfm{e})$ and corresponding prior $P(\bfm{e})$.
+for the state observable distributions $\Pr(o \mid \bfm{e})$, where $\bfm{e} \equiv \{\mu, \sigma^2\}$, with $\mu$ the state mean observable and $\sigma^2$ the variance (which will include both the distribution of the observable characterizing the state and any broadening from measurement noise).
+Other models (including multidimensional or multimodal observation models) are possible, and require replacing only the observation model $\varphi(o \mid \bfm{e})$ and corresponding prior $\Pr(\bfm{e})$.
 
@@
 \begin{eqnarray}
-P(\bfm{e}) &\propto& \sigma^{-1} \label{equation:emission-parameter-prior} ,
+\Pr(\bfm{e}) &\propto& \sigma^{-1} \label{equation:emission-parameter-prior} ,
 \end{eqnarray}
@@
 \begin{eqnarray}
-P(\bfm{e} \mid \bfm{o}) &\propto& \sigma^{-(N+1)} \exp\left[-\frac{1}{2 \sigma^2} \sum_{n=1}^N (o_n - \mu)^2 \right] ,
+\Pr(\bfm{e} \mid \bfm{o}) &\propto& \sigma^{-(N+1)} \exp\left[-\frac{1}{2 \sigma^2} \sum_{n=1}^N (o_n - \mu)^2 \right] ,
 \end{eqnarray}
@@
 \begin{eqnarray}
-\mu &\sim& P(\mu \mid \sigma^2, \bfm{o}) \nonumber \\
-\sigma^2 &\sim& P(\sigma^2 \mid \mu, \bfm{o})
+\mu &\sim& \Pr(\mu \mid \sigma^2, \bfm{o}) \nonumber \\
+\sigma^2 &\sim& \Pr(\sigma^2 \mid \mu, \bfm{o})
 \end{eqnarray}
@@
 \begin{eqnarray}
-%P(\mu \mid \sigma^2, \bfm{o}) &\equiv& \frac{P(\mu, \sigma^2 \mid \bfm{o})}{\int d\mu \, P(\mu, \sigma^2 \mid \bfm{o})} \nonumber \\
+%\Pr(\mu \mid \sigma^2, \bfm{o}) &\equiv& \frac{\Pr(\mu, \sigma^2 \mid \bfm{o})}{\int d\mu \, \Pr(\mu, \sigma^2 \mid \bfm{o})} \nonumber \\
 %&\propto& \exp\left[-\frac{1}{2 (\sigma^2/N)} (\mu - \hat{\mu})^2 \right] 
-P(\mu \mid \sigma^2, \bfm{o}) &\propto& \exp\left[-\frac{1}{2 (\sigma^2/N)} (\mu - \hat{\mu})^2 \right] 
+\Pr(\mu \mid \sigma^2, \bfm{o}) &\propto& \exp\left[-\frac{1}{2 (\sigma^2/N)} (\mu - \hat{\mu})^2 \right] 
 \end{eqnarray}
@@
 \begin{eqnarray}
-%P(\sigma^2 \mid \mu, \bfm{o}) &=& \frac{p(\mu, \sigma^2 \mid \bfm{o})}{\int d\sigma^2 \, p(\mu, \sigma^2 \mid \bfm{o})} \nonumber \\
+%\Pr(\sigma^2 \mid \mu, \bfm{o}) &=& \frac{\Pr(\mu, \sigma^2 \mid \bfm{o})}{\int d\sigma^2 \, \Pr(\mu, \sigma^2 \mid \bfm{o})} \nonumber \\
 %&\propto& \sigma^{-(N+1)} \exp\left[-\frac{1}{2 \sigma^2} \sum_{n=1}^N (o_n - \mu)^2 \right] \nonumber \\
 %&\propto& \sigma^{-(N+1)} \exp\left[-\frac{N \hat{\sigma}^2}{2 \sigma^2} \right] 
-P(\sigma^2 \mid \mu, \bfm{o}) &\propto& \sigma^{-(N+1)} \exp\left[-\frac{N \hat{\sigma}^2}{2 \sigma^2} \right] 
+\Pr(\sigma^2 \mid \mu, \bfm{o}) &\propto& \sigma^{-(N+1)} \exp\left[-\frac{N \hat{\sigma}^2}{2 \sigma^2} \right] 
 \end{eqnarray}
@@
 
-Other output probabilities, such as mixtures of normal distributions or other distributions, can be substituted by simply changing $P(\bfm{E} \mid \bfm{O}, \bfm{S})$ and the scheme by which $\bfm{E}$ is updated.
+Other output probabilities, such as mixtures of normal distributions or other distributions, can be substituted by simply changing $\Pr(\bfm{E} \mid \bfm{O}, \bfm{S})$ and the scheme by which $\bfm{E}$ is updated.
 

### 2015-02-18 manuscripts/bayesian-hmm-method/bhmm.tex
wrote T-matrix sampling description in manuscript appendix

@@
 \usepackage{amssymb}
+\usepackage[]{algorithm2e}
 
@@
 
-The transition matrix is updated using the reversible transition matrix sampling scheme of No\'{e}~\cite{noe:jcp:2008:transition-matrix-sampling,chodera-noe:jcp:2011:observable-sampling}.
-Here, an adjusted count matrix $\bfm{C} \equiv (c_{ij})$ is computed using the updated hidden state sequence $\bfm{S}'$,
-\begin{eqnarray}
-c_{ij} &=& b_{ij} +\sum_{t=1}^{L} \delta_{is_{t-1}} \delta_{js_{t}} ,
-\end{eqnarray}
-where the Kronecker $\delta_{ij} = 1$ if $i = j$ and zero otherwise, and $\bfm{B} \equiv (b_{ij})$ is a matrix of prior pseudocounts, which we take to be zero following the work of No\'{e} et al.~\cite{noe:pnas:2009:ww-domain}.
-Using the adjusted count matrix $\bfm{C}$, a Metropolis-Hastings Monte Carlo procedure~\cite{metropolis-hastings} is used to update the matrix and produce a new sample from $\Pr(\bfm{T}' \mid \bfm{S}')$.
-Two move types are attempted, selected with equal probability, and 1000 trial moves are attempted to generate a new sample $\bfm{T}'$ that is approximately uncorrelated from the previous $\bfm{T}$.
-Prior to starting the Monte Carlo procedure, the vector of equilibrium probabilities for all states $\bfv{\pi}$ is computed according to 
-\begin{eqnarray}
-\bfm{T}^\T \bfv{\pi} = \bfv{\pi} .
-\end{eqnarray}
-
-The first move type is a reversible element shift.
-A pair of states $(i,j)$, $i \ne j$, are selected with uniform probability, and a random number $\Delta$ is selected uniformly over the interval,
-\begin{equation*}
-\Delta \in[\max(-T_{ii},-\frac{\pi_{j}}{\pi_{i}}T_{jj}),T_{ij}] .
-\end{equation*}
-The changed elements in the proposed transition matrix $\bfm{T}'$ are then given by:
-\begin{align*}
-T'_{ij} & =T_{ij}-\Delta \:\: ; \:\: T'_{ji}  =T_{ji}-\frac{\pi_{i}}{\pi_{j}}\Delta \\
-T'_{ii} & =T_{ii}+\Delta \:\: ; \:\: T'_{jj}  =T_{jj}+\frac{\pi_{i}}{\pi_{j}}\Delta  .
-\end{align*}
-This move is accepted with probability
-\begin{eqnarray}
-\lefteqn{P_\mathrm{accept}(\bfm{T}' | \bfm{T} ) = \mathrm{min}\left\{1, \sqrt{\frac{(T'_{ij})^{2}+(T'_{ji})^{2}}{(T_{ij})^{2}+(T_{ji})^{2}}} \right. }\\
-&\times& \left.
+%%%%%%%%%%%%%%
+
+\color{magenta}
+
+Here we employ the reversible transition matrix sampling algorithm derived in [{\bf FN: reference to reversible T-matrix estimation preprint when available}].
+We first define a coordinate transform to unconditional transition probabilities:
+\begin{equation}
+  \label{eq:transformation_T_X}
+  x_{ij} = \pi_{i} T_{ij}
+\end{equation}
+and sample from their posterior 
+\begin{equation}
+  \label{eq:X_posterior}
+   p(\mathbf{X} \mid \mathbf{C} ) = p(\mathbf{C} \mid \mathbf{X}) p(\mathbf{X})
+\end{equation}
+using the prior
+\begin{equation}
+  \label{eq:X_prior}
+   p(\mathbf{X}) = \prod_{i \ge j} x_{ij}^{-1}.
+\end{equation}
+This choice of prior ensures that the maximum likelihood estimator, i.e. the maximum of $p(\mathbf{C} \mid \mathbf{X})$ coincides with
+the mean of the posterior $p(\mathbf{X} \mid \mathbf{C} )$. This behavior is desirable in order to have uncertainty intervals envelop the
+maximum likelihood estimator. Without a suitable choice of prior this property is lost, as described in~\cite{noe:pnas:2009:ww-domain}.
+
+It turns out that the posterior (\ref{eq:X_posterior}) can be conveniently sampled by considering yet another set of
+variables $\mathbf{V}=[v_{ij}]$, which are related to $\mathbf{X}$ by the scaling $x_{ij} = v_{ij} / \sum_{k,l} v_{kl}$. Thus, we can still recover 
+transition matrices from the $v_{ij}$ variables by row normalization:
+\begin{equation}
+  \label{eq:T_from_V}
+  T_{ij} = \frac{v_{ij}}{\sum_k v_{ik}}
+\end{equation}
+
+Using the $v_{ij}$ variables, the sampling of posterior (\ref{eq:X_posterior}) is implemented by iterating the following Gibbs sampling
+procedure. Initially, when given an initial transition matrix $\mathbf{T}^{(0)}$, we set 
+\begin{equation} 
+	v_{ij}^{(0)} = x_{ij}^{(0)} = \pi_{i} T_{ij}^{(0)}
+\end{equation} 
+When starting from scratch we initialize 
+\begin{equation} 
+	v_{ij}^{(0)} = (c_ij + c_ji) / 2 \sum_{k,l} c_{kl}.
+\end{equation} 
+In each Gibbs sampling cycle we iterate through all $i \ge j$. If $c_{ij} + c_{ji} = 0$, the conditional probability density is concentrated at
+$v_{ij} = 0$ and we do not perform any action. If  $c_{ij} + c_{ji} > 0$, one of the following two steps is executed for either diagonal or off-diagonal
+elements:
+
+%% diagonals
+{\bf Case 1: diagonals}: If $i = j$, we sample $v_{ii}^{(k+1)}$ according to the conditional
+\begin{eqnarray}
+  \mathbb{P}(v_{ii} | \{v_{kl}\}_{k \geq l} \backslash \{v_{ii}\}, \mathbf{C}) 
+							& \propto & v_{ii}^{c_{ii}-1} (v_i - v_{ii})^{-c_{i}}. \\
+							& \propto & T_{ii}^{c_{ii}-1} (1 - T_{ii})^{c_{i}-c_{ii}-1}.
+  \label{eq:conditional_pkk}
+\end{eqnarray}
+where we have introduced the row sums
+\begin{equation}
+   v_i = \sum_j v_{ij}.
+\end{equation}
+The conditional (\ref{eq:conditional_pkk}) can be sampled by drawing Beta-distributed variables 
+\begin{equation}
+	T_{ii}\sim \mathrm{Beta}(\alpha,\beta)
+	\label{eq:betadist}
+\end{equation}
+with parameters
+$\alpha=c_{ii}$ and $\beta=c_{i}-c_{ii}$ (see \cite{Devroye_Springer86_NonuniformRandom}). The new element $v_{ii}^{(k+1)}$
+can be obtained from the sampled $T_{ii}$ by
+\begin{equation}
+  v_{ii}^{(k+1)}=\frac{T_{ii}}{1-T_{ii}} (v_i^{(k)} - v_{ii}^{(k)}).
+  \label{eq:diagonal_transformation}
+\end{equation}
+
+% off-diagonals.
+{\bf Case 2: off-diagonals}: If $i \neq j$, we have to sample a conditional density which is not easily tractable. Therefore,
+we approximate the conditional density as follows:
+\begin{equation}
+  \begin{aligned}
+&  v_{ij}^{(c_{ij}+c_{ji}-1)} (v_i^{(k)} - v_{ij}^{(k)} + v_{ij})^{-c_i} (v_j^{(k)} - v_{ji}^{(k)} + v_{ij})^{-c_j}  \\ 
+& \approx q(v_{ij} \mid \alpha, \beta, f_0)
+  \end{aligned}
+  \label{eq:conditional_vkl_f}
+\end{equation}
+where employ the approximate function
+\begin{equation}
+	q(v_{ij} \mid \alpha, \beta, f_0)         = (v_{ij})^{-1} \exp \left( \alpha \log v_{ij} - \beta v_{ij} + f_0 \right)
+	\label{eq:approximation_conditional}
+\end{equation}
+which is, up to a constant factor, a Gamma distribution with parameters $\alpha$, $\beta$ which can be
+efficiently sampled \cite{Devroye_Springer86_NonuniformRandom}. The parameters are given by
+\begin{subequations}
+  \begin{align}
+    \alpha & = -f''(\bar{v}_{ij}) \bar{v}_{ij}^2 \label{eq:alpha}\\
+    \beta & = -f''(\bar{v}_{ij}) \bar{v}_{ij} \label{eq:beta} \\
+    f_0 & = f(\bar{v}_{ij}) +  f''(\bar{v}_{ij}) \bar{v}_{ij}^2 (\log \bar{v}_{ij} -1) \label{eq:f0},
+  \end{align}
+\end{subequations}
+with the maximum point $\bar{v}_{ij}$ as the positive root of a quadratic equation:
+\begin{equation}
+  \label{eq:maximum_point}
+  \bar{v}_{ij} = \frac{-b + \sqrt{b^2-4ac}}{2a}
+\end{equation}
+and $a$, $b$, $c$ given by:
+\begin{subequations}
+  \begin{align}
+    a &=c_{i}+ c_{j}- c_{ij}- c_{ji} \label{eq:a} \\
+    b &= c_{i} (v_i^{(k)} - v_{ij}^{(k)}) + c_{j} (v_j^{(k)} - v_{ji}^{(k)}) \nonumber \\
+       & - (c_{ij}+c_{ji}) (v_i^{(k)} - v_{ij}^{(k)} + v_j^{(k)} - v_{ji}^{(k)}) \label{eq:b} \\
+    c &= -(c_{ij}+c_{ji}) (v_i^{(k)} - v_{ij}^{(k)}) (v_j^{(k)} - v_{ji}^{(k)}) \label{eq:c}
+  \end{align}
+\end{subequations}
+The approximate conditional $q$ in \eqref{eq:approximation_conditional} is
+not necessarily a bounding envelope for the true conditional
+\eqref{eq:conditional_vkl_f} so that instead of rejection sampling we
+use the approximation $q$ as a proposal density in a Metropolis sampling
+step. The acceptance probability for the step is then given as $\min
+\{1, p_{acc}\}$ with
+\begin{equation}
+  \label{eq:pacc}
+  p_{acc}=\frac{ q(v_{ij}^{(k)}) v_{ij}^{(k)} \exp f(v_{ij}^{(k+1)})}{ q(v_{ij}^{(k+1)}) v_{ij}^{(k+1)}  \exp f(v_{ij}^{(k)})}.
+\end{equation}
+
+For each element sampled, symmetry of the $\mathbf{V}$ matrix is enforced by setting $v_{lk} = v_{kl}$. After each
+cycle of the Gibbs sampling procedure, we rescale $\mathbf{V}$ by
+\begin{equation}
+  v_{ij}^{(k+1)} \leftarrow \frac{v_{ij}^{(k+1)}}{\sum_{l,m} v_{lm}^{(k+1)}}
+\end{equation}
+
+In brief, we can summarize the above procedure by the algorithm:
+\begin{algorithm}
+  \DontPrintSemicolon
+  \SetAlgoNoLine
+  \SetInd{0.5em}{1.0em}
+  \KwIn{$C$, $V^{(k)}$}
+  \KwOut{$V^{(k+1)}$}
+  \For{$i=1,\dotsc,n$}{
+    \For{$j=1,\dotsc, i$}{
+      \If{$c_{ij}+c_{ji}>0$}{
+        \eIf{$i=j$}{
+          % $\alpha=c_{kk}$, $\beta=c_{k}-c_{kk}$\;
+          % $p_{kk}=\frac{v^{(j)}_{kk}}{v^{(j)}_{k,-k}+v^{(j)}_{kk}}$\;
+          % $p'_{kk} \sim \text{Beta}(\alpha, \beta)$\;
+          % $v_{kk}^{(j+1)}=\frac{p'_{kk}}{1-p'_{kk}}v_{k, -k}^{(j)}$\;
+          Sample $v^{(k+1)}_{ii}$ according to
+          \eqref{eq:conditional_pkk}, \eqref{eq:transformation_p_v}\;
+        }{ Obtain parameters $a$, $b$, $c$ from \eqref{eq:a},
+          \eqref{eq:b}, \eqref{eq:c}\; Compute $\bar{v}_{ij}$
+          according to \eqref{eq:maximum_point}\; Compute parameters
+          $\alpha$, $\beta$, $f_0$ using \eqref{eq:alpha},
+          \eqref{eq:beta},\eqref{eq:f0}\; Sample $v'_{ij}$ according
+          to \eqref{eq:approximation_conditional}\; Compute $p_{acc}$
+          from \eqref{eq:pacc}\; $U$ from $[0, 1]$ uniformly
+          distributed\; \eIf{$U<\min\{1, p_{acc}\}$}{
+            $v_{ij}^{(k+1)}=v'_{ij}$\; }{
+            $v_{ij}^{(k+1)}=v^{(k)}_{ij}$\; }
+          $v_{ji}^{(j+1)}=v_{ij}^{(k+1)}$\; } } } }
+\caption{Reversible sampling algorithm \label{alg:reversible_sampler}}
+\end{algorithm}
+
+\color{black}
+
+%The transition matrix is updated using the reversible transition matrix sampling scheme of No\'{e}~\cite{noe:jcp:2008:transition-matrix-sampling,chodera-noe:jcp:2011:observable-sampling}.
+%Here, an adjusted count matrix $\bfm{C} \equiv (c_{ij})$ is computed using the updated hidden state sequence $\bfm{S}'$,
+%\begin{eqnarray}
+%c_{ij} &=& b_{ij} +\sum_{t=1}^{L} \delta_{is_{t-1}} \delta_{js_{t}} ,
+%\end{eqnarray}
+%where the Kronecker $\delta_{ij} = 1$ if $i = j$ and zero otherwise, and $\bfm{B} \equiv (b_{ij})$ is a matrix of prior pseudocounts, which we take to be zero following the work of No\'{e} et al.~\cite{noe:pnas:2009:ww-domain}.
+%Using the adjusted count matrix $\bfm{C}$, a Metropolis-Hastings Monte Carlo procedure~\cite{metropolis-hastings} is used to update the matrix and produce a new sample from $\Pr(\bfm{T}' \mid \bfm{S}')$.
+%Two move types are attempted, selected with equal probability, and 1000 trial moves are attempted to generate a new sample $\bfm{T}'$ that is approximately uncorrelated from the previous $\bfm{T}$.
+%Prior to starting the Monte Carlo procedure, the vector of equilibrium probabilities for all states $\bfv{\pi}$ is computed according to 
+%\begin{eqnarray}
+%\bfm{T}^\T \bfv{\pi} = \bfv{\pi} .
+%\end{eqnarray}
+%
+%The first move type is a reversible element shift.
+%A pair of states $(i,j)$, $i \ne j$, are selected with uniform probability, and a random number $\Delta$ is selected uniformly over the interval,
+%\begin{equation*}
+%\Delta \in[\max(-T_{ii},-\frac{\pi_{j}}{\pi_{i}}T_{jj}),T_{ij}] .
+%\end{equation*}
+%The changed elements in the proposed transition matrix $\bfm{T}'$ are then given by:
+%\begin{align*}
+%T'_{ij} & =T_{ij}-\Delta \:\: ; \:\: T'_{ji}  =T_{ji}-\frac{\pi_{i}}{\pi_{j}}\Delta \\
+%T'_{ii} & =T_{ii}+\Delta \:\: ; \:\: T'_{jj}  =T_{jj}+\frac{\pi_{i}}{\pi_{j}}\Delta  .
+%\end{align*}
+%This move is accepted with probability
+%\begin{eqnarray}
+%\lefteqn{P_\mathrm{accept}(\bfm{T}' | \bfm{T} ) = \mathrm{min}\left\{1, \sqrt{\frac{(T'_{ij})^{2}+(T'_{ji})^{2}}{(T_{ij})^{2}+(T_{ji})^{2}}} \right. }\\
+%&\times& \left.
 \left(\frac{T'_{ii}}{T_{ii}}\right)^{c_{ii}}\left(\frac{T'_{ij}}{T_{ij}}\right)^{c_{ij}} \left(\frac{T'_{jj}}{T_{jj}}\right)^{c_{jj}}\left(\frac{T'_{ji}}{T_{ji}}\right)^{c_{ji}} \right\} .
 \nonumber
-\end{eqnarray}
-This move will leave the vector of stationary probabilities $\bfv{\pi}$ unchanged.
-
-The second move type is a row shift.
-A row $i$ of $\bfm{T}$ is selected with uniform probability, and a random number $\eta$ chosen uniformly over the interval 
-\begin{equation*}
-\eta\in\left[0,\frac{1}{1-T_{ii}}\right]
-\end{equation*}
-and used to update row $i$ of $\bfm{T}$ according to
-\begin{eqnarray}
-T'_{ij} &=& \begin{cases}
-\eta T_{ij} & j = 1, \ldots, M, \:\: j \ne i\\
-\eta (T_{ii} - 1) + 1 & j = i
-\end{cases}
-\end{eqnarray}
-This move is accepted with probability
-\begin{eqnarray}
-P_\mathrm{accept}(\bfm{T}' | \bfm{T}) &=& \mathrm{min} \left\{1, \eta^{(M-2)} \eta^{(c_{i*}-c_{ii})}\left(\frac{1-\eta(1-T_{ii})}{T_{ii}}\right)^{c_{ii}} \right\} . \nonumber\\
-\end{eqnarray}
-The row shift operation will change the stationary distribution of $\bfv{\pi}'$, but it may be efficiently updated with
-\begin{align*}
-\pi'_{i} & =\frac{\pi_{i}}{\pi_{i}+\eta(1-\pi_{i})} \:\: ; \:\: \pi'_{j} =\frac{\eta \, \pi_{j}}{\pi_{i}+\eta(1-\pi_{i})}.
-\end{align*}
-Since this update scheme is incremental, it will accumulate numerical errors over time that cause the updated $\bfv{\pi}$ to drift away from the stationary distribution of the current transition matrix.
-To avoid this, $\bfv{\pi}$ is recomputed from the current sample of the transition matrix in regular intervals (here, every 100 sampling steps).
+%\end{eqnarray}
+%This move will leave the vector of stationary probabilities $\bfv{\pi}$ unchanged.
+%
+%The second move type is a row shift.
+%A row $i$ of $\bfm{T}$ is selected with uniform probability, and a random number $\eta$ chosen uniformly over the interval 
+%\begin{equation*}
+%\eta\in\left[0,\frac{1}{1-T_{ii}}\right]
+%\end{equation*}
+%and used to update row $i$ of $\bfm{T}$ according to
+%\begin{eqnarray}
+%T'_{ij} &=& \begin{cases}
+%\eta T_{ij} & j = 1, \ldots, M, \:\: j \ne i\\
+%\eta (T_{ii} - 1) + 1 & j = i
+%\end{cases}
+%\end{eqnarray}
+%This move is accepted with probability
+%\begin{eqnarray}
+%P_\mathrm{accept}(\bfm{T}' | \bfm{T}) &=& \mathrm{min} \left\{1, \eta^{(M-2)} \eta^{(c_{i*}-c_{ii})}\left(\frac{1-\eta(1-T_{ii})}{T_{ii}}\right)^{c_{ii}} \right\} . \nonumber\\
+%\end{eqnarray}
+%The row shift operation will change the stationary distribution of $\bfv{\pi}'$, but it may be efficiently updated with
+%\begin{align*}
+%\pi'_{i} & =\frac{\pi_{i}}{\pi_{i}+\eta(1-\pi_{i})} \:\: ; \:\: \pi'_{j} =\frac{\eta \, \pi_{j}}{\pi_{i}+\eta(1-\pi_{i})}.
+%\end{align*}
+%Since this update scheme is incremental, it will accumulate numerical errors over time that cause the updated $\bfv{\pi}$ to drift away from the stationary distribution of the current transition matrix.
+%To avoid this, $\bfv{\pi}$ is recomputed from the current sample of the transition matrix in regular intervals (here, every 100 sampling steps).
 

### 2015-02-18 manuscripts/bayesian-hmm-method/bhmm.tex

@@
 While maximum-likelihood schemes such as hidden Markov models have solved this problem for other classes of single-molecule experiments by using temporal information to aid in the inference of a sequence of distinct conformational states, these methods do not give a clear picture of how precisely the model parameters are determined by the data due to instrument noise and finite-sample statistics, both significant problems in force spectroscopy.
-{\color{blue}We solve this problem through a Bayesian extension that allows the experimental uncertainties to be directly quantified, and allow uncertainty to be further reduced through the imposition of physical priors such as the requirement that dynamics obey detailed balance.}
+{\color{blue}We solve this problem through a Bayesian extension that allows the experimental uncertainties to be directly quantified, and permits uncertainty to be further reduced through the imposition of physical priors (such as the requirement that dynamics obey detailed balance).}
 We illustrate the utility of this approach in characterizing the {\color{red}observed multistate kinetics of two molecules in a stationary optical trap: an RNA hairpin that exhibits three-state behavior and a protein system (apomyglobin at pH 5) that exhibits two-state behavior with highly overlapping force distributions characterizing each state.}

### 2015-02-20 examples/synthetic-three-state-model/generate-figure.py
Adding three-state synthetic example.

@@ -1,0 +1,36 @@
+#!/usr/bin/env python
+"""
+Generate plots for synthetic three-state force spectroscopy model.
+
+"""
+
+from bhmm import testsystems
+from bhmm import BHMM
+
+# Create model.
+true_model = testsystems.force_spectroscopy_model()
+
+# Generate synthetic data.
+[O, S] = true_model.generate_synthetic_observation_trajectories(ntrajectories=1, length=50000)
+
+
+
+# Initialize BHMM.
+tau = 0.001 # time interval per observation
+nstates = 3
+bhmm = BHMM(nstates, O)
+
+# Sample models.
+models = bhmm.sample(nsamples=10, save_hidden_state_trajectory=False)
+
+# Extract hidden state trajectories and observations.
+s_t = S[0]
+o_t = O[0]
+
+# Generate a sample saving a hidden state trajectory.
+models = bhmm.sample(nsamples=1, save_hidden_state_trajectory=True)
+
+# Plot.
+plot_state_assignments(model, s_t, o_t, time_units='s', obs_label='force / pN', tau=tau, pdf_filename='synthetic-three-state-model.pdf')
+
+

### 2015-02-20 examples/synthetic-three-state-model/generate-figure.py
Debugging MLHMM initialization.

@@ -6,31 +6,35 @@ Generate plots for synthetic three-state force spectroscopy model.
 
 from bhmm import testsystems
 from bhmm import BHMM
+from bhmm import plots
 
 # Create model.
 true_model = testsystems.force_spectroscopy_model()
+nstates = true_model.nstates
 
 # Generate synthetic data.
 [O, S] = true_model.generate_synthetic_observation_trajectories(ntrajectories=1, length=50000)
 
-
+# DEBUG
+print "synthetic observation trajectories:"
+print O
+print "Total state visits, min_state, max_state:"
+print testsystems.total_state_visits(nstates, S)
 
 # Initialize BHMM.
 tau = 0.001 # time interval per observation
-nstates = 3
-bhmm = BHMM(nstates, O)
+bhmm = BHMM(O, nstates)
 
 # Sample models.
 models = bhmm.sample(nsamples=10, save_hidden_state_trajectory=False)
-
-# Extract hidden state trajectories and observations.
-s_t = S[0]
-o_t = O[0]
 
 # Generate a sample saving a hidden state trajectory.
 models = bhmm.sample(nsamples=1, save_hidden_state_trajectory=True)
 
 # Plot.
-plot_state_assignments(model, s_t, o_t, time_units='s', obs_label='force / pN', tau=tau, pdf_filename='synthetic-three-state-model.pdf')
+model = models[0]
+s_t = model.hidden_state_trajectories[0]
+o_t = O[0]
+plots.plot_state_assignments(model, s_t, o_t, time_units='s', obs_label='force / pN', tau=tau, pdf_filename='synthetic-three-state-model.pdf')
 
 

### 2015-02-21 examples/p5ab-hairpin/generate-figure.py
Added p5ab hairpin data.

@@ -1,0 +1,39 @@
+#!/usr/bin/env python
+"""
+Generate plots for synthetic three-state force spectroscopy model.
+
+"""
+
+from bhmm import testsystems
+from bhmm import BHMM
+from bhmm import plots
+
+# Load force data.
+from netcdf4 import Dataset
+ncfile = Dataset('fiber3-trace011', 'r')
+tau = 0.001 # 1 kHz
+obs_label = 'force / pN'
+time_units = 's' # seconds
+o_t = ncfile.variables['force'] # load trace
+O = [o_t] # form list of traces
+
+# Initialize MLHMM.
+print "Initializing MLHMM..."
+nstates = 3
+mlhmm = MLHMM(O, nstates)
+
+# Plot.
+plots.plot_state_assignments(model, mlhmm.S[0], o_t, time_units=time_unit, obs_label=obs_label, tau=tau, pdf_filename='fiber3-trace11-mlhmm-stateassignments.pdf')
+
+# Initialize BHMM, using MLHMM model as initial model.
+print "Initializing BHMM..."
+bhmm = BHMM(O, nstates, intial_model=mlhmm)
+
+# Sample models.
+models = bhmm.sample(nsamples=10, save_hidden_state_trajectory=False)
+
+# Generate a sample saving a hidden state trajectory.
+models = bhmm.sample(nsamples=1, save_hidden_state_trajectory=True)
+
+
+

### 2015-02-21 manuscripts/bayesian-hmm-method/bhmm.tex

@@
 \resizebox{\textwidth}{!}{\includegraphics{fiber3-trace11-mlhmm-stateassignments}}
-\caption{{\bf Experimental force trajectory of %the p5ab 
-{\color{red} an RNA} hairpin and MLHMM state assignments.}
+\caption{{\bf Experimental force trajectory of {\color{red} an RNA} hairpin and MLHMM state assignments.}
 Observed samples are colored by their hidden state assignments.

### 2015-02-21 examples/synthetic-three-state-model/generate-figure.py
Updated synthetic three-state example script.

@@ -5,14 +5,16 @@ Generate plots for synthetic three-state force spectroscopy model.
 """
 
 from bhmm import testsystems
-from bhmm import BHMM
+from bhmm import MLHMM, BHMM
 from bhmm import plots
 
 # Create model.
 true_model = testsystems.force_spectroscopy_model()
 nstates = true_model.nstates
+tau = 0.001 # time interval per observation
 
 # Generate synthetic data.
+print "Generating synthetic data..."
 [O, S] = true_model.generate_synthetic_observation_trajectories(ntrajectories=1, length=50000)
 
 # DEBUG
@@ -21,20 +23,29 @@ print O
 print "Total state visits, min_state, max_state:"
 print testsystems.total_state_visits(nstates, S)
 
-# Initialize BHMM.
-tau = 0.001 # time interval per observation
-bhmm = BHMM(O, nstates)
-
-# Sample models.
-models = bhmm.sample(nsamples=10, save_hidden_state_trajectory=False)
-
-# Generate a sample saving a hidden state trajectory.
-models = bhmm.sample(nsamples=1, save_hidden_state_trajectory=True)
+# Generate MLHMM.
+print "Generating MLHMM..."
+mlhmm = MLHMM(O, nstates, verbose=True)
+mlhmm_model = mlhmm.fit()
 
 # Plot.
-model = models[0]
+s_t = mlhmm_model.hidden_state_trajectories[0]
+o_t = O[0]
+plots.plot_state_assignments(model, s_t, o_t, time_units='s', obs_label='force / pN', tau=tau, pdf_filename='synthetic-three-state-model-mlhmm.pdf')
+
+# Initialize BHMM with MLHMM model.
+print "Sampling models from BHMM..."
+bhmm = BHMM(O, nstates, initial_model=mlhmm_model, verbose=True)
+bhmm_models = bhmm.sample(nsamples=10, save_hidden_state_trajectory=False)
+
+# Generate a sample saving a hidden state trajectory.
+final_models = bhmm.sample(nsamples=1, save_hidden_state_trajectory=True)
+
+# Plot.
+model = final_models[0]
 s_t = model.hidden_state_trajectories[0]
 o_t = O[0]
-plots.plot_state_assignments(model, s_t, o_t, time_units='s', obs_label='force / pN', tau=tau, pdf_filename='synthetic-three-state-model.pdf')
[…]

### 2015-02-21 examples/synthetic-three-state-model/generate-figure.py
Added __repr__ and __str__ to HMM

@@ -26,6 +26,13 @@ print testsystems.total_state_visits(nstates, S)
 # Generate MLHMM.
 print "Generating MLHMM..."
 mlhmm = MLHMM(O, nstates, verbose=True)
+
+print "Initial guess:"
+print str(mlhmm.model.output_model)
+print mlhmm.model.Tij
+print mlhmm.model.Pi
+
+print "Fitting HMM..."
 mlhmm_model = mlhmm.fit()
 
 # Plot.

### 2015-02-21 examples/p5ab-hairpin/generate-figure.py
debugged example plotting scripts

@@ -6,28 +6,36 @@ Generate plots for synthetic three-state force spectroscopy model.
 
 from bhmm import testsystems
 from bhmm import BHMM
+from bhmm import MLHMM
 from bhmm import plots
 
 # Load force data.
-from netcdf4 import Dataset
+from netCDF4 import Dataset
 ncfile = Dataset('fiber3-trace011.nc', 'r')
 tau = 0.001 # 1 kHz
 obs_label = 'force / pN'
 time_units = 's' # seconds
 o_t = ncfile.variables['force'] # load trace
+
+# force to make a copy because netCDF appears to cause problems
+o_t = o_t[::10]
+tau *= 10
+# -------------------
+
 O = [o_t] # form list of traces
 
 # Initialize MLHMM.
 print "Initializing MLHMM..."
 nstates = 3
 mlhmm = MLHMM(O, nstates)
+mle = mlhmm.fit()
 
 # Plot.
-plots.plot_state_assignments(model, mlhmm.S[0], o_t, time_units=time_unit, obs_label=obs_label, tau=tau, pdf_filename='fiber3-trace11-mlhmm-stateassignments.pdf')
+plots.plot_state_assignments(mle, mlhmm.hidden_state_trajectories[0], o_t, time_units=time_units, obs_label=obs_label, tau=tau, pdf_filename='fiber3-trace11-mlhmm-stateassignments.pdf')
 
 # Initialize BHMM, using MLHMM model as initial model.
 print "Initializing BHMM..."
-bhmm = BHMM(O, nstates, intial_model=mlhmm)
+bhmm = BHMM(O, nstates, initial_model=mle)
 
 # Sample models.
 models = bhmm.sample(nsamples=10, save_hidden_state_trajectory=False)

### 2015-02-21 examples/synthetic-three-state-model/generate-figure.py

@@ -36,9 +36,9 @@ print "Fitting HMM..."
 mlhmm_model = mlhmm.fit()
 
 # Plot.
-s_t = mlhmm_model.hidden_state_trajectories[0]
+s_t = mlhmm.hidden_state_trajectories[0]
 o_t = O[0]
-plots.plot_state_assignments(model, s_t, o_t, time_units='s', obs_label='force / pN', tau=tau, pdf_filename='synthetic-three-state-model-mlhmm.pdf')
+plots.plot_state_assignments(mlhmm_model, s_t, o_t, time_units='s', obs_label='force / pN', tau=tau, pdf_filename='synthetic-three-state-model-mlhmm.pdf')
 
 # Initialize BHMM with MLHMM model.
 print "Sampling models from BHMM..."

### 2015-02-21 examples/p5ab-hairpin/generate-figure.py
downsampling p5ab data to 1 ms

@@ -18,8 +18,8 @@ time_units = 's' # seconds
 o_t = ncfile.variables['force'] # load trace
 
 # force to make a copy because netCDF appears to cause problems
-o_t = o_t[::10]
-tau *= 10
+o_t = o_t[::50]
+tau *= 50
 # -------------------
 
 O = [o_t] # form list of traces

### 2015-02-21 examples/p5ab-hairpin/generate-figure.py
plotting also BHMM data for p5ab

@@ -41,7 +41,13 @@ bhmm = BHMM(O, nstates, initial_model=mle)
 models = bhmm.sample(nsamples=10, save_hidden_state_trajectory=False)
 
 # Generate a sample saving a hidden state trajectory.
-models = bhmm.sample(nsamples=1, save_hidden_state_trajectory=True)
+final_models = bhmm.sample(nsamples=1, save_hidden_state_trajectory=True)
+
+# Plot.
+model = final_models[0]
+s_t = model.hidden_state_trajectories[0]
+o_t = O[0]
+plots.plot_state_assignments(model, s_t, o_t, time_units=time_units, obs_label=obs_label, tau=tau, pdf_filename='fiber3-trace11-bhmm-stateassignments.pdf')
 
 
 

### 2015-02-21 examples/rnase-h-d10a/generate-figure.py
debugged RNAse-H generation script

@@ -6,34 +6,45 @@ Generate plots for synthetic three-state force spectroscopy model.
 
 from bhmm import testsystems
 from bhmm import BHMM
+from bhmm import MLHMM
 from bhmm import plots
 
 # Load force data.
-from netcdf4 import Dataset
+from netCDF4 import Dataset
 ncfile = Dataset('rnase-h-d10a-trace47.nc', 'r')
 tau = 0.001 # 1 kHz
 obs_label = 'force / pN'
 time_units = 's' # seconds
 o_t = ncfile.variables['force'] # load trace
+
+# copy data
+o_t = o_t[:]
 O = [o_t] # form list of traces
 
 # Initialize MLHMM.
 print "Initializing MLHMM..."
 nstates = 3
 mlhmm = MLHMM(O, nstates)
+mle = mlhmm.fit()
 
 # Plot.
-plots.plot_state_assignments(model, mlhmm.S[0], o_t, time_units=time_unit, obs_label=obs_label, tau=tau, pdf_filename='fiber3-trace11-mlhmm-stateassignments.pdf')
+plots.plot_state_assignments(mle, mlhmm.hidden_state_trajectories[0], o_t, time_units=time_units, obs_label=obs_label, tau=tau, pdf_filename='RNAseH_trace47-mlhmm-nstates'+str(nstates)+'-stateassignments.pdf')
 
 # Initialize BHMM, using MLHMM model as initial model.
 print "Initializing BHMM..."
-bhmm = BHMM(O, nstates, intial_model=mlhmm)
+bhmm = BHMM(O, nstates, initial_model=mle)
 
 # Sample models.
 models = bhmm.sample(nsamples=10, save_hidden_state_trajectory=False)
 
 # Generate a sample saving a hidden state trajectory.
-models = bhmm.sample(nsamples=1, save_hidden_state_trajectory=True)
+final_models = bhmm.sample(nsamples=1, save_hidden_state_trajectory=True)
+
+# Plot.
+model = final_models[0]
+s_t = model.hidden_state_trajectories[0]
+o_t = O[0]
+plots.plot_state_assignments(model, s_t, o_t, time_units=time_units, obs_label=obs_label, tau=tau, pdf_filename='RNAseH_trace47-bhmm-nstates'+str(nstates)+'-stateassignments.pdf')
 
 
 

### 2015-02-22 examples/p5ab-hairpin/generate-figure-threestate.py
Generated figures for two- and three-state models of p5ab data.

@@ -18,8 +18,9 @@ time_units = 's' # seconds
 o_t = ncfile.variables['force'] # load trace
 
 # force to make a copy because netCDF appears to cause problems
-o_t = o_t[::50]
-tau *= 50
+nsubsample = 50 # subsampling rate
+o_t = o_t[::nsubsample]
+tau *= nsubsample
 # -------------------
 
 O = [o_t] # form list of traces
@@ -27,15 +28,20 @@ O = [o_t] # form list of traces
 # Initialize MLHMM.
 print "Initializing MLHMM..."
 nstates = 3
-mlhmm = MLHMM(O, nstates)
+mlhmm = MLHMM(O, nstates, verbose=True)
+
+# Plot initial guess.
+plots.plot_state_assignments(mlhmm.model, None, O[0], time_units=time_units, obs_label=obs_label, tau=tau, pdf_filename='fiber3-trace11-guess-stateassignments-threestate.pdf')
+
+# Fit MLHMM
 mle = mlhmm.fit()
 
 # Plot.
-plots.plot_state_assignments(mle, mlhmm.hidden_state_trajectories[0], o_t, time_units=time_units, obs_label=obs_label, tau=tau, pdf_filename='fiber3-trace11-mlhmm-stateassignments.pdf')
+plots.plot_state_assignments(mle, mlhmm.hidden_state_trajectories[0], o_t, time_units=time_units, obs_label=obs_label, tau=tau, pdf_filename='fiber3-trace11-mlhmm-stateassignments-threestate.pdf')
 
 # Initialize BHMM, using MLHMM model as initial model.
 print "Initializing BHMM..."
-bhmm = BHMM(O, nstates, initial_model=mle)
+bhmm = BHMM(O, nstates, initial_model=mle, verbose=True)
 
 # Sample models.
 models = bhmm.sample(nsamples=10, save_hidden_state_trajectory=False)
@@ -47,7 +53,7 @@ final_models = bhmm.sample(nsamples=1, save_hidden_state_trajectory=True)
 model = final_models[0]
 s_t = model.hidden_state_trajectories[0]
 o_t = O[0]
-plots.plot_state_assignments(model, s_t, o_t, time_units=time_units, obs_label=obs_label, tau=tau, pdf_filename='fiber3-trace11-bhmm-stateassignments.pdf')
[…]

### 2015-02-22 examples/p5ab-hairpin/generate-figure-twostate.py

@@ -1,0 +1,59 @@
+#!/usr/bin/env python
+"""
+Generate plots for synthetic three-state force spectroscopy model.
+
+"""
+
+from bhmm import testsystems
+from bhmm import BHMM
+from bhmm import MLHMM
+from bhmm import plots
+
+# Load force data.
+from netCDF4 import Dataset
+ncfile = Dataset('fiber3-trace011.nc', 'r')
+tau = 0.001 # 1 kHz
+obs_label = 'force / pN'
+time_units = 's' # seconds
+o_t = ncfile.variables['force'] # load trace
+
+# force to make a copy because netCDF appears to cause problems
+nsubsample = 50 # subsampling rate
+o_t = o_t[::nsubsample]
+tau *= nsubsample
+# -------------------
+
+O = [o_t] # form list of traces
+
+# Initialize MLHMM.
+print "Initializing MLHMM..."
+nstates = 2
+mlhmm = MLHMM(O, nstates, verbose=True)
+
+# Plot initial guess.
+plots.plot_state_assignments(mlhmm.model, None, O[0], time_units=time_units, obs_label=obs_label, tau=tau, pdf_filename='fiber3-trace11-guess-stateassignments-twostate.pdf')
+
+# Fit MLHMM
+mle = mlhmm.fit()
+
+# Plot.
+plots.plot_state_assignments(mle, mlhmm.hidden_state_trajectories[0], o_t, time_units=time_units, obs_label=obs_label, tau=tau, pdf_filename='fiber3-trace11-mlhmm-stateassignments-twostate.pdf')
+
+# Initialize BHMM, using MLHMM model as initial model.
+print "Initializing BHMM..."
+bhmm = BHMM(O, nstates, initial_model=mle, verbose=True)
+
+# Sample models.
+models = bhmm.sample(nsamples=10, save_hidden_state_trajectory=False)
+
+# Generate a sample saving a hidden state trajectory.
+final_models = bhmm.sample(nsamples=1, save_hidden_state_trajectory=True)
+
+# Plot.
+model = final_models[0]
+s_t = model.hidden_state_trajectories[0]
+o_t = O[0]
+plots.plot_state_assignments(model, s_t, o_t, time_units=time_units, obs_label=obs_label, tau=tau, pdf_filename='fiber3-trace11-bhmm-stateassignments-twostate.pdf')
+
+
+

### 2015-03-23 examples/discrete_1d_2min/run.py
working on convergence problem when lag time is in similar range like timescale

@@ -11,10 +11,12 @@ nstates = 2
 
 # multiple lags
 #lags = [300]
-lags = [1,2,5,10,20,50,100,200,300,400,500,600,700,800,900,1000]
+#lags = [1,2,5,10,20,50,100,200,300,400,500,600,700,800,900,1000]
+lags = [300]
 its      = np.zeros((len(lags)))
 its_mean = np.zeros((len(lags)))
 its_std  = np.zeros((len(lags)))
+likelihoods = np.zeros((len(lags)))
 for (i,lag) in enumerate(lags):
     # prepare shifted lagged data
     observations = []
@@ -22,17 +24,18 @@ for (i,lag) in enumerate(lags):
         observations.append(o[shift:][::lag])
 
     # initial HMM
-    hmm = MLHMM(observations, nstates, kernel='c', output_model_type='discrete')
-    hmm.fit()
-    P = hmm.model.Tij
+    em = MLHMM(observations, nstates, kernel='c', output_model_type='discrete')
+    em.fit()
+    P = em.model.Tij
     its[i] = msmana.timescales(P, tau=lag)[1]
+    likelihoods[i] = em.model.likelihood
 
     # Initialize BHMM, using MLHMM model as initial model.
     print "BHMM for lag ",lag
-    bhmm = BHMM(observations, nstates, initial_model=hmm.model, verbose=True)
+    bhmm = BHMM(observations, nstates, initial_model=em.model, verbose=True)
 
     # Sample models.
-    nsamples = 1000
+    nsamples = 10
     models = bhmm.sample(nsamples=nsamples, save_hidden_state_trajectory=False)
     its_sample = np.zeros((nsamples))
     for j in range(nsamples):
@@ -48,5 +51,5 @@ print itsref
 
 print 'Resulting timescales:'
 for i in range(len(lags)):
-    print lags[i], its[i], its_mean[i], '+-', its_std[i]
+    print lags[i], likelihoods[i], its[i], its_mean[i], '+-', its_std[i]
 

### 2015-03-23 examples/rnase-h-d10a/generate-figure.py
removed kernel specification in generate figure

@@ -23,7 +23,7 @@ O = [o_t] # form list of traces
 
 # Initialize MLHMM.
 print "Initializing MLHMM..."
-nstates = 3
+nstates = 4
 mlhmm = MLHMM(O, nstates)
 mle = mlhmm.fit()
 

### 2015-03-23 examples/discrete_1d_2min/run.py
fixed discrete HMM convergence problem (normalization error)

@@ -10,9 +10,7 @@ o = np.loadtxt('2well_traj_100K.dat', dtype=int)
 nstates = 2
 
 # multiple lags
-#lags = [300]
-#lags = [1,2,5,10,20,50,100,200,300,400,500,600,700,800,900,1000]
-lags = [300]
+lags = [1,2,5,10,20,50,100,200,300,400,500,600,700,800,900,1000]
 its      = np.zeros((len(lags)))
 its_mean = np.zeros((len(lags)))
 its_std  = np.zeros((len(lags)))

### 2015-05-17 examples/discrete_1d_2min/run.py
[fixed examples]

@@ -26,7 +26,7 @@ for (i,lag) in enumerate(lags):
         observations.append(o[shift:][::lag])
 
     # initial HMM
-    hmm = bhmm.estimate_hmm(observations, nstates, output_model_type='discrete')
+    hmm = bhmm.estimate_hmm(observations, nstates, type='discrete')
     its[i] = lag*hmm.timescales
     likelihoods[i] = hmm.likelihood
 

### 2015-05-17 examples/p5ab-hairpin/generate-figure.py

@@ -50,7 +50,7 @@ def run(nstates, nsamples):
 
     # Initialize BHMM, using MLHMM model as initial model.
     print "Initializing BHMM and running with "+str(nsamples)+" samples."
-    sampler = bhmm.BayesianHMMSampler(O, nstates, initial_model=mle)
+    sampler = bhmm.BHMM(O, nstates, initial_model=mle)
 
     # Sample models.
     bhmm_models = sampler.sample(nsamples=nsamples, save_hidden_state_trajectory=False)

### 2015-05-17 examples/synthetic-three-state-model/generate-figure.py
[fixed examples]

@@ -61,7 +61,7 @@ def run(nstates, nsamples):
 
     # Initialize BHMM with MLHMM model.
     print "Sampling models from BHMM..."
-    sampler = bhmm.BayesianHMMSampler(O, nstates, initial_model=mle)
+    sampler = bhmm.BHMM(O, nstates, initial_model=mle)
     bhmm_models = sampler.sample(nsamples=nsamples, save_hidden_state_trajectory=False)
 
     # Generate a sample saving a hidden state trajectory.

### 2015-05-17 manuscripts/bayesian-hmm-method/bhmm.tex
Updated synthetic three-state model plot- and table-generation scripts; updates to manuscript.

@@
 
+\author{Frank No\'{e}}
+ \thanks{Corresponding author}
+ \email{frank.noe@fu-berlin.de}
+ \affiliation{DFG Research Center Matheon, FU Berlin, Arnimallee 6, 14195 Berlin, Germany}
+
+\author{Nina Singhal Hinrichs}
+ \email{nshinrichs@uchicago.edu}
+ \affiliation{Departments of Statistics and Computer Science, University of Chicago, IL 60637, USA}
+
 \author{Bettina Keller}
@@
  \affiliation{Jason L.~Choy Laboratory of Single Molecule Biophysics, Institute for Quantitative Biosciences, University of California, Berkeley, CA 94720, USA}
-
-\author{Frank No\'{e}}
- \email{frank.noe@fu-berlin.de}
- \affiliation{DFG Research Center Matheon, FU Berlin, Arnimallee 6, 14195 Berlin, Germany}
 
@@
 
-\author{Nina Singhal Hinrichs}
- \email{nshinrichs@uchicago.edu}
- \affiliation{Departments of Statistics and Computer Science, University of Chicago, IL 60637, USA}
-
 \date{\today}
@@
 While maximum-likelihood schemes such as hidden Markov models have solved this problem for other classes of single-molecule experiments by using temporal information to aid in the inference of a sequence of distinct conformational states, these methods do not give a clear picture of how precisely the model parameters are determined by the data due to instrument noise and finite-sample statistics, both significant problems in force spectroscopy.
-{\color{blue}We solve this problem through a Bayesian extension that allows the experimental uncertainties to be directly quantified, and permits uncertainty to be further reduced through the imposition of physical priors (such as the requirement that dynamics obey detailed balance).}
-We illustrate the utility of this approach in characterizing the {\color{red}observed multistate kinetics of two molecules in a stationary optical trap: an RNA hairpin that exhibits three-state behavior and a protein system (apomyglobin at pH 5) that exhibits two-state behavior with highly overlapping force distributions characterizing each state.}
+We solve this problem through a Bayesian extension that allows the experimental uncertainties to be directly quantified, and permits uncertainty to be further reduced through the imposition of physical priors (such as the requirement that dynamics obey detailed balance).
+We illustrate the utility of this approach in characterizing the observed multistate kinetics of several molecules in a stationary optical trap: a simple protein system (apomyglobin at pH 5) that exhibits two-state behavior with highly overlapping force distributions characterizing each state, an RNA hairpin that exhibits three-state behavior, and a complex protein system (RNase H) that exhibits complex multistate behavior.
 
@@
 Recent advances in biophysical measurement have led to an unprecedented ability to monitor the dynamics of single biological macromolecules, such as proteins and nucleic acids~\cite{ritort:j-phys:2006:single-molecule-review}.
-%As a new approach to probing the behavior of biological macromolecules, these experiments promise to change the way we study folding, dynamics, catalysis, association, transcription, translation, and motility, providing otherwise-inaccessible information about microscopic kinetics, energetics, mechanism, and the stochastic heterogeneity inherent in these processes.
 Advances in instrumentation for optical force spectroscopy in particular have produced instruments of extraordinary stability, precision, and temporal resolution~\cite{block:rev-sci-instrum:2004:optical-trapping,bustamante:annu-rev-biochem:2008:optical-tweezers} that have already demonstrated great utility in the study of biomolecules in the presence of externally perturbative forces~\cite{bustamante:nature:2003:dna-mechanics,block:curr-opin-chem-biol:2008:rna,bustamante:cell:2011:central-dogma}.
@@
 
-In a typical single-molecule optical trapping experiment, a protein or nucleic acid is tethered to two polystyrene beads by dsDNA handles that {\color{red}serve as spacers to} prevent the molecule under study from interacting with the beads {({\color{red} as in Figure~\ref{figure:trap-configuration})}.
+In a typical single-molecule optical trapping experiment, a protein or nucleic acid is tethered to two polystyrene beads by dsDNA handles that serve as spacers to prevent the molecule under study from interacting with the beads (as in Figure~\ref{figure:trap-configuration}).
 The handle-biomolecule-handle assembly---referred to as a fiber---is associated with the beads through tight noncovalent interactions, with one bead held in an optical trap and the other either suctioned to a micropipette or held in a second optical trap.
 During an experiment, the position of the bead within the laser trap is monitored, and either the relative displacement from the trap center or the total force on the bead is recorded, resulting in a timeseries such as the one depicted in Fig.~\ref{figure:p5ab-observed-trajectory}.
-{\color{red}The instrument can generally be operated in several modes: a force ramp mode, in which the trap is translated rapidly enough to potentially carry the system {\color{red}transiently} out of equilibrium; a passive mode, in which the trap is held fixed and the equilibrium fluctuations of the system are observed; and a constant force-feedback mode, in which the trap is continually repositioned to maintain a specified average constant force on the fiber.}
-Here, we concern ourselves with passive-mode force probe experiments, though both nonequilibrium experiments and constant force-feedback experiments and their analysis remain an exciting topic of active research~\cite{ritort:adv-chem-phys:2008:nonequilibrium-review,elms:biophys-j:2012:force-feedback}
-{\color{magenta}FN: add additional citations for both modes / analyses}.
+The instrument can generally be operated in several modes: a force ramp mode, in which the trap is translated rapidly enough to potentially carry the system transiently out of equilibrium~\cite{force-ramp}; a passive mode, in which the trap is held fixed and the equilibrium fluctuations of the system are observed~\cite{passive-mode}; and a constant force-feedback mode, in which the trap is continually repositioned to maintain a specified average constant force on the fiber~\cite{force-feedback}.
+Here, we concern ourselves with passive-mode force probe experiments, though both nonequilibrium experiments and constant force-feedback experiments and their analysis remain an exciting topic of active research~\cite{ritort:adv-chem-phys:2008:nonequilibrium-review,elms:biophys-j:2012:force-feedback,woodside-nonequilibrium}.
+{\color{red}[FN: Add additional citations for these modes and analyses.]}
 
 Often, the dynamics observed in these experiments appears to be dominated by stochastic transitions between two or more strongly metastable conformational states~\cite{schuette-huisinga:2002:biomolecular-conformations-as-metastable-states,chodera-singhal:jcp:2007:automatic-state-decomposition}---regions of conformation space in which the system remains for long times before making a transition to another conformational state.
-{\color{red}The observed force trace in Fig.~\ref{figure:p5ab-observed-trajectory}, for example, demonstrates apparent hopping behavior between two or more such states.}
-{\color{red}Transitions among strongly metastable states such as these} are generally well-described by first-order kinetics~\cite{noe:pnas:2009:ww-domain}.
+The observed force trace in Fig.~\ref{figure:p5ab-observed-trajectory}, for example, demonstrates apparent hopping behavior between two or more such states.
+Transitions among strongly metastable states such as these are generally well-described by first-order kinetics~\cite{noe:pnas:2009:ww-domain}.
 While visual inspection of the dynamics may suggest the clear presence of multiple metastable states, quantitative characterization of these states is often difficult.
 First, the observed force or extension is unlikely to correspond to a true reaction coordinate easily able to separate all metastable states~\cite{best-paci-hummer-dudko:jpcb:2008:pulling-coordinate,rief-paci:pre:2010:multidimensional-landscape,thirumalai:2011:prl:pfold,chodera-pande:arxiv:2011:single-molecule-pfold}, and second, measurement noise may further broaden the force or extension signatures of individual states, increasing their overlap.
-{\color{red} As a result, a histogram of the observed temporal trace can have strongly overlapping state distributions, as in Fig.~\ref{figure:p5ab-observed-trajectory}, right.}
+As a result, a histogram of the observed temporal trace can have strongly overlapping state distributions, as in the right panel of Fig.~\ref{figure:p5ab-observed-trajectory}.
 Attempting to separate these states by simply dividing the observed force or extension range into regions, following current practice~\cite{woodside:science:2006:dna-hairpin-optical-trap,ritort:biophys-j:2011:short-handles}, can often lead to a high degree of state mis-assignment that results in the estimated rate constants and state distributions containing a significant amount of error~\cite{chodera:rate-theory,PrinzChoderaNoe_PRX14_RateTheory} (see \emph{Supplementary Material: Comparison with threshold model}).
@@
 \caption{{\bf Single-molecule optical trapping configuration.}
-The biomolecule of interest---here, %the p5ab 
-{\color{red} an} RNA hairpin---is tethered to two polystyrene beads by dsDNA handles.
+The biomolecule of interest---here, an RNA hairpin---is tethered to two polystyrene beads by dsDNA handles.
 The fluctuating force on one bead held in an optical trap is monitored, while the other bead is held suctioned to a micropipette tip.
@@
 
-Given observed timeseries data for forces or extensions, the maximum likelihood estimate (MLE) of the model parameters (transition rates and state force or extension distributions) and sequence of hidden states corresponding to the observed data {\color{red} (used to color the points in Fig.~\ref{figure:p5ab-observed-trajectory})} can be determined by standard methods~\cite{baum:1970:ann-math-statist:baum-welch,viterbi:1967:IEEE-trans-info-theory:viterbi-algorithm}, as demonstrated in recent work~\cite{kruithof-vannoort:biophys-j:2009:hmm-force}.
+Given observed timeseries data for forces or extensions, the maximum likelihood estimate (MLE) of the model parameters (transition rates and state force or extension distributions) and sequence of hidden states corresponding to the observed data (used to color the points by state assignment in Fig.~\ref{figure:p5ab-observed-trajectory}) can be determined by standard methods~\cite{baum:1970:ann-math-statist:baum-welch,viterbi:1967:IEEE-trans-info-theory:viterbi-algorithm}, as demonstrated in recent work~\cite{kruithof-vannoort:biophys-j:2009:hmm-force}.
 Unfortunately, this approach has a number of significant drawbacks.
-%Due to technical limitations, experiments often suffer from limited statistics---the events of interest (transitions between states or visits to rare states) may occur only a few times during the course of the measurement, and data for additional fibers is time-consuming to collect.
-%As a result, while the MLE yields the most likely set of model parameters, there may be 
-{\color{magenta}The parameters estimated by the MLE might be subject to enormous uncertainty. 
-Methods for estimating the standard error or confidence intervals from hidden Markov models~\cite{aittokallio-uusipaikka:technical-report:2000:hmm-standard-error} make a number of approximations that can lead to a significant underestimation of the error.}
-%, these schemes can be prohibitively costly for long traces, and may still significantly underestimate the statistical error for short traces due to the normally-distributed error approximation inherent in the approach.
-%{\color{blue}[JDC: Should we say something about bootstrap here since we will compare to it later?]}
+The parameters estimated by the MLE might be subject to enormous uncertainty, especially when the dataset contains a small number of transitions between states or visits to rare states.
+Methods for estimating the standard error or confidence intervals from hidden Markov models~\cite{aittokallio-uusipaikka:technical-report:2000:hmm-standard-error} make a number of approximations that can lead to a significant underestimation of the error.
 %The high cost (both in terms of instrument and experimenter time) of collecting additional data also means that it is not a simple task to judge how much data need be collected to test a particular hypothesis in a statistically meaningful way.
@@
 
-{\color{red}To demonstrate the utility of the Bayesian HMM in the analysis of real force spectroscopy data, we illustrate its application to the analysis of an RNA hairpin exhibiting apparent three-state behavior (Figure~\ref{figure:p5ab-observed-trajectory}) and a protein system that exhibits two-state behavior with highly overlapping force signatures (Figure~\ref{figure:apomyoglobin-observed-trajectory}) collected in a passive (equilibrium) optical trap.
+To demonstrate the utility of the Bayesian HMM in the analysis of real force spectroscopy data, we illustrate its application to the analysis of an RNA hairpin exhibiting apparent three-state behavior (Figure~\ref{figure:p5ab-observed-trajectory}) and a protein system that exhibits two-state behavior with highly overlapping force signatures (Figure~\ref{figure:apomyoglobin-observed-trajectory}) collected in a passive (equilibrium) optical trap.
 In addition to maximum likelihood estimates of equilibrium state probabilities, state-to-state transition rates, and the force distributions characterizing each state, the Bayesian extension provides estimates of confidence intervals for individual parameters (Tables~\ref{table:p5ab-bhmm-confidence-intervals} and \ref{table:apomyoglobin-bhmm-confidence-intervals}) as well as the full joint distribution of parameters.
@@
 %We follow with a discussion of future extensions of this approach, and include a complete description of the computational implementation in \emph{Algorithms}.
-A Matlab implementation of the approach described here is also made freely available online [\url{http://github.com/choderalab/bhmm}].
-}
+A Matlab implementation of the approach described here is also made freely available online [\url{http://github.com/bhmm/bhmm}].
 
@@
 We treat these conformational states as the hidden states of the model, because we cannot directly observe the identity of the metastable state in which the system resides.
-The hidden Markov model presumes the observed data $\bfm{O} \equiv \{o^{}_t\}$ was generated according to the following model dependent on {\color{red} the hidden states $\bfm{S} \equiv \{s_t\}$,} parameters $\bfm{\Theta} \equiv \{\bfm{T}, \bfm{E}\}$, where $\bfm{T}$ is an $M \times M$ row-stochastic transition matrix and $\bfm{E}$ a set of emission parameters governing the observable (force or extension) distributions for each of the $M$ hidden states, prior information about the initial state distribution $\bfv{\rho}^{}$, {\color{red} and emission probability distribution $\varphi(o | \bfm{e})$},
+The hidden Markov model presumes the observed data $\bfm{O} \equiv \{o^{}_t\}$ was generated according to the following model dependent on the sequence of hidden states $\bfm{S} \equiv \{s_t\}$, parameters $\bfm{\Theta} \equiv \{\bfm{T}, \bfm{E}\}$, where $\bfm{T}$ is an $M \times M$ row-stochastic transition matrix and $\bfm{E}$ a set of emission parameters governing the observable (force or extension) distributions for each of the $M$ hidden states, prior information about the initial state distribution $\bfv{\rho}^{}$, {\color{red} and emission probability distribution $\varphi(o | \bfm{e})$},
 \begin{eqnarray}
@@
 \noindent
-\resizebox{\textwidth}{!}{\includegraphics{model-mlhmm-stateassignments}}
+%\resizebox{\textwidth}{!}{\includegraphics{model-mlhmm-stateassignments}}
+\resizebox{\textwidth}{!}{\includegraphics{synthetic-three-state-model-mlhmm}}
 \caption{{\bf Synthetic force trajectory and inferred state assignments in MLHMM.}
@@
 Dark horizontal lines terminating in triangles to the right denote state means, while lightly colored bands indicate one standard deviation on either side of the state mean.
-The gray histogram on the right side shows the total observed probability of samples, while the colored peaks show the weighted Gaussian output contribution from each state, and the black outline the weighted sum of the Gaussian output contributions from the HMM states.
+The histograms on the right side shows the state-resolved probabilities of samples, while the colored peaks show the weighted Gaussian output contribution from each state, and the black outline the weighted sum of the Gaussian output contributions from the HMM states.
 \label{figure:model-stateassignments}}
 \end{figure*}
+
+\begin{table*}
+\caption{{\bf Estimated mean model parameters and confidence intervals for synthetic timeseries data}}
+\label{table:synthetic-confidence-intervals}
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lccccc}
+\hline
+&  &  & \multicolumn{3}{c}{\bf Estimated Model Parameters}  \\ \cline{4-6}
+\multicolumn{2}{l}{\bf Property} & \bf True Value & \bf 1 000 observations & \bf 10 000 observations & \bf 100 000 observations\\ \hline
+		Equilibrium probability 		& $\pi_{1}$ & $0.333$ & $0.338_{\:0.130}^{\:0.543}$  & $0.309_{\:0.251}^{\:0.374}$  & $0.330_{\:0.299}^{\:0.363}$  \\
+		& $\pi_{2}$ & $0.333$ & $0.351_{\:0.082}^{\:0.735}$  & $0.335_{\:0.246}^{\:0.432}$  & $0.329_{\:0.309}^{\:0.350}$  \\
+		& $\pi_{3}$ & $0.333$ & $0.312_{\:0.118}^{\:0.510}$  & $0.356_{\:0.299}^{\:0.411}$  & $0.341_{\:0.322}^{\:0.357}$  \\
+		\hline
+		Transition probability ($\Delta t = $1.0 ms) 		& $T_{11}$ & $0.989$ & $0.952_{\:0.926}^{\:0.972}$ & $0.947_{\:0.939}^{\:0.955}$ & $0.988_{\:0.987}^{\:0.989}$ \\
+		& $T_{12}$ & $0.010$ & $0.005_{\:0.001}^{\:0.015}$ & $0.001_{\:0.001}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
+		& $T_{13}$ & $0.001$ & $0.043_{\:0.024}^{\:0.067}$ & $0.051_{\:0.044}^{\:0.060}$ & $0.011_{\:0.010}^{\:0.012}$ \\
+		& $T_{21}$ & $0.010$ & $0.005_{\:0.001}^{\:0.016}$ & $0.001_{\:0.000}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
+		& $T_{22}$ & $0.940$ & $0.990_{\:0.975}^{\:0.998}$ & $0.988_{\:0.983}^{\:0.991}$ & $0.949_{\:0.946}^{\:0.951}$ \\
+		& $T_{23}$ & $0.050$ & $0.005_{\:0.001}^{\:0.015}$ & $0.011_{\:0.008}^{\:0.015}$ & $0.050_{\:0.048}^{\:0.053}$ \\
+		& $T_{31}$ & $0.001$ & $0.046_{\:0.028}^{\:0.072}$ & $0.044_{\:0.038}^{\:0.052}$ & $0.011_{\:0.009}^{\:0.012}$ \\
+		& $T_{32}$ & $0.050$ & $0.005_{\:0.001}^{\:0.016}$ & $0.010_{\:0.007}^{\:0.014}$ & $0.049_{\:0.046}^{\:0.051}$ \\
+		& $T_{33}$ & $0.949$ & $0.949_{\:0.922}^{\:0.970}$ & $0.945_{\:0.937}^{\:0.953}$ & $0.941_{\:0.938}^{\:0.943}$ \\
+		\hline
+		\hline
+		Transition rate (ms$^{-1}$) 		& $k_{12}$ & $0.010$ & $0.005_{\:0.001}^{\:0.015}$ & $0.001_{\:0.001}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
+		& $k_{13}$ & $0.001$ & $0.043_{\:0.024}^{\:0.067}$ & $0.051_{\:0.044}^{\:0.060}$ & $0.011_{\:0.010}^{\:0.012}$ \\
+		& $k_{21}$ & $0.010$ & $0.005_{\:0.001}^{\:0.016}$ & $0.001_{\:0.000}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
+		& $k_{23}$ & $0.050$ & $0.005_{\:0.001}^{\:0.015}$ & $0.011_{\:0.008}^{\:0.015}$ & $0.050_{\:0.048}^{\:0.053}$ \\
+		& $k_{31}$ & $0.001$ & $0.046_{\:0.028}^{\:0.072}$ & $0.044_{\:0.038}^{\:0.052}$ & $0.011_{\:0.009}^{\:0.012}$ \\
+		& $k_{32}$ & $0.050$ & $0.005_{\:0.001}^{\:0.016}$ & $0.010_{\:0.007}^{\:0.014}$ & $0.049_{\:0.046}^{\:0.051}$ \\
+		\hline
+		State mean lifetime (ms) 		& $t_{1}$ & $90.408$ & $21.797_{\:13.184}^{\:36.162}$ & $18.645_{\:15.899}^{\:21.524}$ & $83.405_{\:75.423}^{\:92.188}$ \\
+		& $t_{2}$ & $16.162$ & $148.113_{\:43.004}^{\:525.609}$ & $82.451_{\:59.320}^{\:115.043}$ & $18.954_{\:18.001}^{\:19.919}$ \\
+		& $t_{3}$ & $19.103$ & $20.138_{\:12.475}^{\:33.395}$ & $17.902_{\:15.417}^{\:20.637}$ & $16.395_{\:15.692}^{\:17.177}$ \\
+		\hline
+		Relaxation time (ms) 		& $\tau_{1}$ & $62.700$ & $82.300_{\:32.985}^{\:200.069}$ & $56.416_{\:44.005}^{\:70.295}$ & $58.230_{\:54.026}^{\:62.826}$ \\
+		& $\tau_{2}$ & $8.909$ & $10.445_{\:7.396}^{\:15.490}$ & $9.353_{\:8.387}^{\:10.399}$ & $8.970_{\:8.681}^{\:9.287}$ \\
+		\hline
+		\hline
+		State force mean (pN) 		& $\mu_{1}$ & $3.000$ & $4.677_{\:4.644}^{\:4.712}$ & $5.606_{\:5.600}^{\:5.613}$ & $3.002_{\:2.991}^{\:3.013}$ \\
+		& $\mu_{2}$ & $4.700$ & $3.069_{\:2.953}^{\:3.187}$ & $3.019_{\:2.986}^{\:3.056}$ & $5.602_{\:5.600}^{\:5.604}$ \\
+		& $\mu_{3}$ & $5.600$ & $5.602_{\:5.580}^{\:5.625}$ & $4.700_{\:4.690}^{\:4.710}$ & $4.703_{\:4.699}^{\:4.706}$ \\
+		\hline
+		State force std dev (pN) 		& $s_{1}$ & $1.000$ & $0.311_{\:0.285}^{\:0.337}$ & $0.199_{\:0.194}^{\:0.203}$ & $1.002_{\:0.995}^{\:1.010}$ \\
+		& $s_{2}$ & $0.300$ & $1.067_{\:0.992}^{\:1.147}$ & $1.012_{\:0.988}^{\:1.039}$ & $0.200_{\:0.199}^{\:0.202}$ \\
+		& $s_{3}$ & $0.200$ & $0.210_{\:0.195}^{\:0.229}$ & $0.297_{\:0.290}^{\:0.305}$ & $0.299_{\:0.296}^{\:0.301}$ \\
+		\hline
+\hline
+\end{tabular*}
+\end{table*}
 
@@
 \color{red}
-As an illustration of the computation of confidence intervals, we simulated a three-state system intended to mimic a protein with (1) a highly-compliance, low-force unfolded state, (2) a moderately compliant low-population intermediate at intermediate force, and (3) a low-compliance, high-force folded state.
+As an illustration of the computation of confidence intervals, we simulated a three-state system intended to mimic a protein with (1) a high-compliance, low-force unfolded state, (2) a moderately compliant low-population intermediate at intermediate force, and (3) a low-compliance, high-force folded state.
 Here, the term ``compliance'' refers to the width of the force or extension distribution characterizing the state.
@@
 \color{black}
-We generated a trajectory of 100 000 observations, and characterized the BHMM mean parameter estimate and 95\% confidence intervals for a subset of this trajectory of varying lengths.
+We generated a trajectory of 100 000 observations, and characterized the BHMM mean parameter estimate and 95\% confidence intervals for a subset of this trajectory of varying lengths by sampling from the Bayesian posterior for 1 000 iterations.
 The results, shown in Table~\ref{table:synthetic-confidence-intervals}, show that the confidence intervals contract as trajectory length increases, as expected, and the BHMM-computed 95\% confidence intervals contain the true model parameters with the expected statistics.
@@
 \begin{table}
-\caption{{\bf BHMM model estimates for %p5ab 
-{\color{red} an RNA} hairpin data.}
-}
+\caption{{\bf BHMM model estimates for RNA hairpin data.}}
 \label{table:p5ab-bhmm-confidence-intervals}

### 2015-05-17 examples/synthetic-three-state-model/generate-figure.py

@@ -1,15 +1,15 @@
 #!/usr/bin/env python
 """
-Generate plots for synthetic three-state force spectroscopy model.
+Generate figures plots and tables for synthetic three-state force spectroscopy model for use in manuscript.
 
 """
 
-import argparse
+import os, os.path
+import numpy as np
 
 import bhmm
 from bhmm.util import testsystems
 from bhmm.util.analysis import generate_latex_table
-
 
 # dynamically import plotting tools
 import os,sys,inspect
@@ -18,21 +18,49 @@ parentdir = os.path.dirname(currentdir)
 sys.path.insert(0,parentdir)
 import plots
 
-def run(nstates, nsamples):
+def generate_synthetic_data(nobservations):
     # Create model.
     true_model = testsystems.force_spectroscopy_model()
     nstates = true_model.nstates
-    tau = 0.001 # time interval per observation
 
     # Generate synthetic data.
     print "Generating synthetic data..."
-    [O, S] = true_model.generate_synthetic_observation_trajectories(ntrajectories=1, length=50000)
+    [O, S] = true_model.generate_synthetic_observation_trajectories(ntrajectories=1, length=nobservations)
 
     # DEBUG
     print "synthetic observation trajectories:"
     print O
     print "Total state visits, min_state, max_state:"
     print testsystems.total_state_visits(nstates, S)
+
+    return [true_model, O, S]
+
+def analyze_data(O, nstates, nsamples=1000, nobservations=None):
+    """
+    Analyze the data with the specified number of states.
+
+    Parameters
+    ----------
+    O : numpy.float
+        observation trajectory
+    nstates : int
+        Number of states to use for analysis.
+    nsamples : int, optional, default=1000
+        Number of iterations to sample from the Bayesian posterior for the BHMM.
+    nobservations : int, optional, default=None
+        If specified, number of observations to use from O.
+
+    """
+
+    # Time interval.
+    tau = 0.001 # time interval (s) for plotting
+
+    # Truncate O to number of observations.
+    if nobservations:
+        print "Using only %d observations" % nobservations
+        O = [ o_t[0:nobservations] for o_t in O ]
+    else:
[…]

### 2015-05-18 examples/p5ab-hairpin/hairpin-example.py
Updated p5ab analysis.

@@ -5,6 +5,7 @@ Generate plots for synthetic three-state force spectroscopy model.
 """
 
 import bhmm
+import argparse
 from bhmm.util.analysis import generate_latex_table
 
 # dynamically import plotting tools
@@ -72,10 +73,14 @@ def run(nstates, nsamples):
                          outfile='p5ab-bhmm-statistics-table.tex')
 
 if __name__ == "__main__":
-    # Be verbose.
-    bhmm.config.verbose = True
+    parser = argparse.ArgumentParser(description='Maximum-likelihood and Bayesian HMM estimation from p5ab hairpin data')
+    parser.add_argument('--nstates', default=3, type=int, help='number of states')
+    parser.add_argument('--nsamples', default=100, type=int, help='number of samples in Bayesian estimator')
+    parser.add_argument('--verbose', dest='verbose', action='store_true', default=True, help='be loud and noisy')
+    args = parser.parse_args()
 
-    # Fit MLHMM and BHMM
-    nstates = 2
-    nsamples = 1000
-    run(nstates, nsamples)
+    # be verbose?
+    bhmm.config.verbose = args.verbose
+
+    # go
+    run(args.nstates, args.nsamples)

### 2015-05-18 manuscripts/bayesian-hmm-method/bhmm.tex
More manuscript edits.

@@
 We treat these conformational states as the hidden states of the model, because we cannot directly observe the identity of the metastable state in which the system resides.
-The hidden Markov model presumes the observed data $\bfm{O} \equiv \{o^{}_t\}$ was generated according to the following model dependent on the sequence of hidden states $\bfm{S} \equiv \{s_t\}$, parameters $\bfm{\Theta} \equiv \{\bfm{T}, \bfm{E}\}$, where $\bfm{T}$ is an $M \times M$ row-stochastic transition matrix and $\bfm{E}$ a set of emission parameters governing the observable (force or extension) distributions for each of the $M$ hidden states, prior information about the initial state distribution $\bfv{\rho}^{}$, {\color{red} and emission probability distribution $\varphi(o | \bfm{e})$},
+The hidden Markov model presumes the observed data $\bfm{O} \equiv \{o^{}_t\}$ was generated according to the following model dependent on the sequence of hidden states $\bfm{S} \equiv \{s_t\}$, parameters $\bfm{\Theta} \equiv \{\bfm{T}, \bfm{E}\}$, where $\bfm{T}$ is an $M \times M$ row-stochastic transition matrix and $\bfm{E}$ a set of emission parameters governing the observable (force or extension) distributions for each of the $M$ hidden states, prior information about the initial state distribution $\bfv{\rho}^{}$, and emission probability distribution $\varphi(o | \bfm{e})$,
 \begin{eqnarray}
@@
 \end{eqnarray}
-In diagrammatic form, the observed state data $\{o_t\}$ and corresponding hidden state history $\{s_t\}$ can be represented 
+In diagrammatic form, the observed state data $\bfm{O} \equiv \{o_t\}$ and corresponding hidden state history $\bfm{S} \equiv \{s_t\}$ can be represented 
 \begin{eqnarray}
@@
 \end{eqnarray} 
-The initial state distribution $\bfv{\rho}^{}$  reflects our knowledge of the initial conditions of the experiment that collected data $\bfm{{\color{red} O}}^{}$.
-In the case that the experiment was prepared in equilibrium, $\bfv{\rho}$ corresponds to the equilibrium distribution $\bfv{\pi}$ of the model transition matrix $\bfm{T}$; if the experiment was prepared out of equilibrium, $\bfv{\rho}$ may be chosen to reflect some other prior distribution (e.g.~the uniform prior).
-
-State transitions $(s^{}_{t-1} \rightarrow s^{}_{t})$ are governed by the discrete transition probability $T_{s^{}_{t-1} s^{}_{t}}$.
+The initial state distribution $\bfv{\rho}^{}$  reflects our knowledge of the initial conditions of the experiment that collected observation data $\bfm{O}^{}$.
+In the case that the experiment was prepared in equilibrium, $\bfv{\rho}$ corresponds to the equilibrium distribution $\bfv{\pi}$ of the model transition matrix $\bfm{T}$; if the experiment was prepared out of equilibrium, $\bfv{\rho}$ may be chosen to reflect some other prior distribution (e.g.~the uniform prior to reflect uncertainty in the initial conditions).
+
+Hidden state transitions $(s^{}_{t-1} \rightarrow s^{}_{t})$ are governed by the discrete transition probability $T_{s^{}_{t-1} s^{}_{t}}$.
 The Markov property of HMMs prescribes that the probability that a system originally in state $i$ at time $t$ is later found in state $j$ at time $t+\Delta t$ is dependent only on knowledge of the state $i$, and given by the corresponding matrix element $T_{ij}$ of the (row-stochastic) transition matrix $\bfm{T}$.
-{\color{magenta}The choice of $\Delta t$ is critical for the accuracy of the model \cite{NoeEtAl_PMMHMM_JCP13} --- $\Delta t$ must
-be chosen to reside in a time-scale gap between the $M$ slow relaxation processes that are resolved and faster relaxation
-processes that are not resolved by the model.}
+The choice of $\Delta t$ is critical for the accuracy of the model \cite{NoeEtAl_PMMHMM_JCP13}---$\Delta t$ must be chosen to reside in a time-scale gap between the $M$ slow relaxation processes that are resolved and faster relaxation processes that are not resolved by the model.
 % not sure if we need this discussion here. We can still say how a rate matrix can be computed at the end.
-%Alternatively, one could instead use the rate matrix $\bfm{K}$, related to the transition matrix $\bfm{T}$ through the equation $\bfm{T} = e^{\bfm{K} \Delta t}$.  
-%If the processes described by $\bfm{T}$ or $\bfm{K}$ are slow compared to the observation interval $\Delta t$, then we can easily estimate the rate matrix from the associated transition matrix in a way that avoids the matrix logarithm, through the expansion $\bfm{K} \approx (\bfm{T} - \bfm{I}) / \Delta t$, where $\bfm{I}$ denotes the $M \times M$ identity matrix.
-
-{\color{magenta}The dependence of the experimental observable on the hidden state}
-%The probabilistic ``emission'' of observables from each state $(s^{}_{t} \rightarrow o^{}_{t})$ 
-is governed by the emission probability $\varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}})$, parametrized by observable emission parameters $\bfm{e}$.
+Alternatively, one could instead use the rate matrix $\bfm{K}$, related to the transition matrix $\bfm{T}$ through the equation $\bfm{T} = e^{\bfm{K} \Delta t}$.  
+If the processes described by $\bfm{T}$ or $\bfm{K}$ are slow compared to the observation interval $\Delta t$, then we can easily estimate the rate matrix from the associated transition matrix in a way that avoids the matrix logarithm, through the expansion $\bfm{K} \approx (\bfm{T} - \bfm{I}) / \Delta t$, where $\bfm{I}$ denotes the $M \times M$ identity matrix.
+
+The dependence of the experimental observable $o_t$ on the hidden state $s_t$ is governed by the emission probability $\varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}})$, parametrized by observable emission parameters $\bfm{e}$.
 For example, in the force spectroscopy applications described here, $\varphi(o \mid \bfm{e}_s)$ is taken to be a univariate normal (Gaussian) distribution parameterized by a mean $\mu$ and variance $\sigma^2$ that characterize each state, such that $\bfm{e}_i \equiv \{\mu_i, \sigma^2_i\}$.
@@
 
-Instead of simply determining the model that maximizes the likelihood of observing the data $\bfm{O}$ given the model parameters $\bfm{\Theta}$, {\color{magenta} we here propose to sample to full posterior distribution of model parameters given the observed data.
-Towards this end, we emply Bayes' theorem to compute the posterior distribution:}
+Instead of simply determining the model that maximizes the likelihood of observing the data $\bfm{O}$ given the model parameters $\bfm{\Theta}$, we here propose to sample to full posterior distribution of model parameters given the observed data.
+Towards this end, we employ Bayes' theorem to compute the posterior distribution:
 \begin{eqnarray}
@@
 
-{\color{red} Detailed descriptions of the algorithms used to fit the maximum likelihood hidden Markov model (MLHMM) and sample from the posterior distribution of the Bayesian hidden Markov model (BHMM) are given in the Appendix.}
+Detailed descriptions of the algorithms used to fit the maximum likelihood hidden Markov model (MLHMM) and sample from the posterior distribution of the Bayesian hidden Markov model (BHMM) are given in the Appendix.
 
@@
 \begin{table*}
-\caption{{\bf Estimated mean model parameters and confidence intervals for synthetic timeseries data}}
+\caption{{\bf Estimated mean model parameters and confidence intervals for synthetic timeseries data (old table)}}
 \label{table:synthetic-confidence-intervals}
@@
 To verify that our BHMM posterior sampling scheme reflects the true uncertainty in the model parameters, we can test that the scheme produces the expected estimates for synthetic data generated from a model with known parameters $\bfm{\Theta}^*$.
-Given observed data $\bfm{O}$ generated from $\Pr(\bfm{O} \mid \bfm{\Theta}^*)$, sampling from the posterior $\Pr(\bfm{\Theta} \mid \bfm{O})$ using the scheme described in \color{blue} \emph{Sampling from the posterior of the BHMM} \color{black} will provide us with confidence intervals $[\theta_\mathrm{low}, \theta_\mathrm{high}]$ for a specified confidence interval $\alpha \in [0,1]$.
+Given observed data $\bfm{O}$ generated from $\Pr(\bfm{O} \mid \bfm{\Theta}^*)$, sampling from the posterior $\Pr(\bfm{\Theta} \mid \bfm{O})$ using the scheme described in Appendix~\ref{section:algorithms:bhmm-posterior-sampling} will provide us with confidence intervals $[\theta_\mathrm{low}, \theta_\mathrm{high}]$ for a specified confidence interval $\alpha \in [0,1]$.
 If these computed confidence intervals are accurate, we should find that the true model parameter $\theta^*$ lies in the computed confidence interval $[\theta^{(\alpha)}_\mathrm{low}, \theta^{(\alpha)}_\mathrm{high}]$ with probability $\alpha$.
@@
 
-\color{red}
 As an illustration of the computation of confidence intervals, we simulated a three-state system intended to mimic a protein with (1) a high-compliance, low-force unfolded state, (2) a moderately compliant low-population intermediate at intermediate force, and (3) a low-compliance, high-force folded state.
@@
 An example realization of a model trajectory mimicking a passive-mode experiment---in which the trap was stationary---is shown in Figure~\ref{figure:model-stateassignments} along with the MLHMM state assignment.
-\color{black}
+
 We generated a trajectory of 100 000 observations, and characterized the BHMM mean parameter estimate and 95\% confidence intervals for a subset of this trajectory of varying lengths by sampling from the Bayesian posterior for 1 000 iterations.
 The results, shown in Table~\ref{table:synthetic-confidence-intervals}, show that the confidence intervals contract as trajectory length increases, as expected, and the BHMM-computed 95\% confidence intervals contain the true model parameters with the expected statistics.
-\color{red}
-In contrast, a model created from simply segmenting the observed forces into disjoint region and assigning state membership based on the force value alone estimates model parameters with significant bias even for 1 000 000 observations (see \emph{Supporting Information}).
-\color{black}
+By contrast, a model created from simply segmenting the observed forces into disjoint region and assigning state membership based on the force value alone estimates model parameters with significant bias even for 1 000 000 observations (see \emph{Supporting Information}).
 
@@
 The results show that the observed confidence intervals line up with the expected confidence intervals to within statistical error, suggesting the BHMM confidence intervals neither underestimate nor overestimate the actual uncertainty in model parameters.
+{\color{red}[JDC: Move supplementary figure to main text or appendix?]}
 

### 2015-05-18 manuscripts/bayesian-hmm-method/bhmm.tex
Removed a number of comments from draft during update process.

@@
 \noindent
-\resizebox{\textwidth}{!}{\includegraphics{fiber3-trace11-mlhmm-stateassignments}}
-\caption{{\bf Experimental force trajectory of {\color{red} an RNA} hairpin and MLHMM state assignments.}
+%\resizebox{\textwidth}{!}{\includegraphics{fiber3-trace11-mlhmm-stateassignments}}
+\resizebox{\textwidth}{!}{\includegraphics{fiber3-trace11-mlhmm-stateassignments-nstate2}}
+\caption{{\bf Experimental force trajectory of an RNA hairpin and MLHMM state assignments.}
 Observed samples are colored by their hidden state assignments.
-Dark horizontal lines terminating in triangles to the right denote state means, while lightly colored bands indicate one standard deviation on either side of the state mean.
+Dark colored horizontal lines denote state mean forces, while lightly colored bands indicate one standard deviation about the state mean force.
 The gray histogram on the right side shows the total observed probability of samples, while the colored peaks show the weighted Gaussian output contribution from each state, and the black outline the weighted sum of the Gaussian output contributions from the HMM states.
@@
 \begin{table}
+    \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}lcc}
+        \hline
+        {\bf Property} & {\bf Symbol} & {\bf Value} \\
+        \hline
+            		Equilibrium probability 		& $\pi_{1}$ & $0.746_{\:0.723}^{\:0.769}$ \\
+		& $\pi_{2}$ & $0.254_{\:0.231}^{\:0.277}$ \\
+		\hline
+		Transition probability ($\Delta t = $0.05 s) 		& $T_{11}$ & $0.9888_{\:0.9878}^{\:0.9898}$ \\
+		& $T_{12}$ & $0.0112_{\:0.0102}^{\:0.0122}$ \\
+		& $T_{21}$ & $0.0329_{\:0.0301}^{\:0.0360}$ \\
+		& $T_{22}$ & $0.9671_{\:0.9640}^{\:0.9699}$ \\
+		\hline
+		\hline
+		Transition rate (s$^{-1}$) 		& $k_{12}$ & $0.2244_{\:0.2047}^{\:0.2444}$ \\
+		& $k_{21}$ & $0.6590_{\:0.6016}^{\:0.7191}$ \\
+		\hline
+		State mean lifetime (s) 		& $t_{1}$ & $4.440_{\:4.073}^{\:4.864}$ \\
+		& $t_{2}$ & $1.495_{\:1.368}^{\:1.639}$ \\
+		\hline
+		Relaxation time (s) 		& $\tau_{1}$ & $1.108_{\:1.034}^{\:1.191}$ \\
+		\hline
+		\hline
+		State force mean (pN) 		& $\mu_{1}$ & $13.844_{\:13.842}^{\:13.847}$ \\
+		& $\mu_{2}$ & $12.619_{\:12.614}^{\:12.623}$ \\
+		\hline
+		State force std dev (pN) 		& $s_{1}$ & $0.220_{\:0.218}^{\:0.221}$ \\
+		& $s_{2}$ & $0.263_{\:0.260}^{\:0.266}$ \\
+		\hline
+
+        \hline
+    \end{tabular*}
     \caption{{\bf BHMM model estimates for RNA hairpin data.}}
+\end{table}
+
+\begin{table}
+\caption{{\bf BHMM model estimates for RNA hairpin data (old).}}
 \label{table:p5ab-bhmm-confidence-intervals}
@@
 
-\color{red}
 A sample of p5ab RNA hairpin from \emph{Tetrahymena thermophilia} utilized in a previous study~\cite{wen-manosas:biophys-j:2007:rna-optical-tweezers} was provided by Jin-Der Wen, and prepared as previously described~\cite{wen-manosas:biophys-j:2007:rna-optical-tweezers}.
@@
 The appropriate observation interval (1 ms) for BHMM analysis was determined by examination of force autocorrelation functions, which suggested that intrastate motion had completely decayed by 1 ms (see \emph{Supplementary Material: Choice of observation interval}).
-{\color{magenta}[FN: Update the selection of the HMM lag time.
+{\color{red}[FN: Update the selection of the HMM lag time.
 I suggest fixing the lag time $\tau$ first, running a maximum-likelihood HMM with N states, and then select $M<N$ based on how many relaxation timescales are larger than $\tau$]}
 The observed 50 kHz force traces were therefore subsampled  to 1 kHz for analysis by retaining every 50th sample.
-\color{black}
 
@@
 
-{\color{magenta}[Describe how/why you selected three over two states in this model]}
+{\color{red}[FN: Describe how/why you selected three over two states in this model.] [JDC: It seems like we should go back to two states.]}
 Table~\ref{table:p5ab-bhmm-confidence-intervals} lists the BHMM posterior means and confidence intervals characterizing the three-state model extracted from this single 60 s observed force trace.
@@
 The use of a detailed balance constraint additionally helps reduce the experimental uncertainty over standard hidden Markov models, as both transitions into and out of conformational states provide valuable information about state kinetics and populations in data-poor conditions~\cite{noe:jcp:2008:transition-matrix-sampling,metzner-noe-schuette:pre:2009:transition-matrix-sampling}.
-Additionally, the Gibbs sampling framework used to sample from the Bayesian posterior can be easily extended to incorporate additional nuisance parameters, such as stochastic models of {\color{red}measurement noise} or laser power fluctuations.
-
-\color{red}
+Additionally, the Gibbs sampling framework used to sample from the Bayesian posterior can be easily extended to incorporate additional nuisance parameters, such as stochastic models of measurement noise or laser power fluctuations.
+
 Drift has historically been a challenging problem for force spectroscopy measurements, with experiments using previous-generation instruments resorting to short force-ramp experiments to minimize the cumulative effects of drift in recorded traces~\cite{collin:nature:2005:crooks-verification}.
@@
 
-Fortunately, obtaining many drift-free measurement traces of the duration presented here with current-generation miniaturized optical traps that employ a pipette to suction one of the beads in the tether (such as the one employed here~\cite{bustamante-smith:2006:minitweezers-patent}), the collection of data sufficiently free of drift is regularly achievable~\cite{bustamante:csh-protocols:2009:optical-tweezers}.
+Fortunately, obtaining many drift-free measurement traces of the duration presented here with current-generation miniaturized optical traps that employ a pipette to suction one of the beads in the tether (such as the one employed for the collection of the apo-myoglobin and RNA hairpin data presented here~\cite{bustamante-smith:2006:minitweezers-patent}), the collection of data sufficiently free of drift is regularly achievable~\cite{bustamante:csh-protocols:2009:optical-tweezers}.
 Moreover, drift is even less problematic for newer generations of instrument that utilize two optical traps instead of a pipette~\cite{block:nature:2003:dual-trap,bustamante:annu-rev-biochem:2008:optical-tweezers}.
-\color{black}
 
@@
 \section{Acknowledgments}
-The authors thank Sergio Bacallado (Stanford University) {\color{red}and anonymous referees} for helpful feedback on an earlier version of this manuscript, and Steve Presse (UCSF) for engaging discussions on this topic.
-\color{red}
+The authors thank Sergio Bacallado (Stanford University) for helpful feedback on an earlier version of this manuscript, and Steve Presse (UCSF) for engaging discussions on this topic.
 JDC acknowledges support from a QB3-Berkeley Distinguished Postdoctoral Fellowship and the Memorial Sloan Kettering Cancer Center.
 BGK gratefully acknowledges the support by the German Research Foundation (DFG) within the collaborative research center SFB 765 (associated project).
-\color{black}
 FN acknowledges funding by DFG Grant 825/2 and ERC starting grant ``pcCell''.
@@
 
-{\color{red} To try to maximize the likelihood in Eq.~\ref{equation:gaussian-mixture-likelihood}, this} approximation is then improved upon by iterating the expectation-maximization procedure described by Bilmes~\cite{bilmes:1998:expectation-maximization},
+To maximize the likelihood in Eq.~\ref{equation:gaussian-mixture-likelihood}, the expectation-maximization procedure described by Bilmes~\cite{bilmes:1998:expectation-maximization} is then performed,
 \begin{eqnarray}
@@
 \end{eqnarray}
-The iterative procedure is terminated at iteration $j$ when the change in the parameters $\{\bfv{\pi}, \bfv{\mu}, \bfv{\sigma}^2\}$ falls below a certain relative threshold, such as $\|\bfv{\pi}^{[j]} - \bfv{\pi}^{[j-1]}\|_2 / \|\bfv{\pi}^{[j]}\|_2 < 10^{-4}$ {\color{red}, where $\| \ldots \|_2$ denotes the euclidean L2 norm}.
+The iterative procedure is terminated at iteration $j$ when the change in the parameters $\{\bfv{\pi}, \bfv{\mu}, \bfv{\sigma}^2\}$ falls below a certain relative threshold, such as $\|\bfv{\pi}^{[j]} - \bfv{\pi}^{[j-1]}\|_2 / \|\bfv{\pi}^{[j]}\|_2 < 10^{-4}$, where $\| \ldots \|_2$ denotes the Euclidean L2-norm.
 
@@
 \end{eqnarray}
-\color{red}
+
 The iterative procedure described in Algorithm 1 of \cite{noe:jcp:2011:msm-review} is then applied to update the transition matrix in a manner that preserves the detailed balance constraint while maximizing the likelihood function $\Pr(\bfm{T} | \bfm{S})$.
 This algorithm proceeds as follows.
-\color{black}
 
@@
 Note that the equilibrium probability vector $\bfv{\pi}$ computed during the Gaussian mixture model fitting is not respected during this step.
-{\color{magenta}[FN: clarify which of the two $\pi$-vectors is used in the subsequent MLHMM]}
+{\color{red}[FN: clarify which of the two $\pi$-vectors is used in the subsequent MLHMM.]}
 
@@
 scheme:
-\color{magenta}
 \begin{eqnarray}
@@
 \end{eqnarray}
-\color{black}
 The initial HMM is usually quick to compute, and can give the experimenter a rough idea of the model parameters, as well as providing a useful starting point for sampling models from the Bayesian posterior.
@@
 %If no detailed balance constraint is used and the prior $\Pr(\bfm{T})$ is Dirichlet in each row of the transition matrix $\bfm{T}$, it is possible to generate an independent sample of the transition matrix from the conditional distribution $\Pr(\bfm{T}' \mid \bfm{S}')$ by sampling each row of the transition matrix from the conjugate Dirichlet posterior using the transition counts from the sampled state sequence $\bfm{S}'$~\cite{noe:jcp:2008:transition-matrix-sampling}.
-{\color{red} If we assume each row of the transition matrix $\bfm{T}$ is independent with Dirichlet prior $\Pr(\bfm{T})$, then it is possible to sample from the transition matrix posterior Dirichlet distribution $\Pr(\bfm{T}' \mid \bfm{S}')$, given the transition counts calculated from the sampled state sequence $\bfm{S}'$~\cite{noe:jcp:2008:transition-matrix-sampling}.}
+If we assume each row of the transition matrix $\bfm{T}$ is independent with Dirichlet prior $\Pr(\bfm{T})$, then it is possible to sample from the transition matrix posterior Dirichlet distribution $\Pr(\bfm{T}' \mid \bfm{S}')$, given the transition counts calculated from the sampled state sequence $\bfm{S}'$~\cite{noe:jcp:2008:transition-matrix-sampling}.
 However, because physical systems in the absence of energy input through an external driving force should satisfy detailed balance, we make use of this constraint in updating our transition probabilities, since this has been demonstrated to substantially reduce parameter uncertainty in the data-limited regime~\cite{noe:jcp:2008:transition-matrix-sampling}.
@@
 
-\color{magenta}
-
-Here we employ the reversible transition matrix sampling algorithm derived in [{\bf FN: reference to reversible T-matrix estimation preprint when available}].
+Here we employ the reversible transition matrix sampling algorithm derived in {\color{red} [FN: reference to reversible T-matrix estimation preprint when available.]}. 
 We first define a coordinate transform to unconditional transition probabilities:
@@
 \end{algorithm}
-
-\color{black}
 
@@
 This behavior might be undesirable, as the states may switch labels during the posterior sampling procedure; this will require any analysis of the models sampled from the posterior to account for the possible permutation symmetry in the states.
-{\color{magenta}Here we choose to order the states according to their value in the experimental observable (force or extension), thus removing
-the ambiguity due to exchangeable state labels. Note that this practice can artificially restrict the confidence intervals of the states when 
-the distributions of the experimental observable overlap strongly in neighboring states}
+Here we choose to order the states according to their value in the experimental observable (force or extension), thus removing the ambiguity due to exchangeable state labels.
+Note that this practice can artificially restrict the confidence intervals of the states when  the distributions of the experimental observable overlap strongly in neighboring states.
 %On the other hand, breaking this symmetry (e.g., by enforcing an ordering on the state mean observables) can artificially restrict the confidence intervals of the states, which might additionally complicate data analysis.

### 2015-05-18 examples/p5ab-hairpin/generate-figure.py

@@ -68,8 +68,8 @@ def run(nstates, nsamples):
     conf = 0.95
     sampled_hmm = bhmm.SampledGaussianHMM(mle, bhmm_models)
     generate_latex_table(sampled_hmm, conf=conf, dt=tau, time_unit='s',
-                         caption='Bayesian HMM parameter estimates for p5ab hairpin data.',
-                         outfile='p5ab-bhmm-statistics-table.tex')
+                         caption='BHMM model estimates for RNA hairpin data.',
+                         outfile='p5ab-bhmm-statistics-table-nstate'+str(nstates)+'.tex')
 
 if __name__ == "__main__":
     # Be verbose.

### 2015-05-18 examples/p5ab-hairpin/hairpin-example.py

@@ -69,8 +69,8 @@ def run(nstates, nsamples):
     conf = 0.95
     sampled_hmm = bhmm.SampledGaussianHMM(mle, bhmm_models)
     generate_latex_table(sampled_hmm, conf=conf, dt=tau, time_unit='s',
-                         caption='Bayesian HMM parameter estimates for p5ab hairpin data.',
-                         outfile='p5ab-bhmm-statistics-table.tex')
+                         caption='BHMM model estimates for RNA hairpin data.',
+                         outfile='p5ab-bhmm-statistics-table'+str(nstates)+'.tex')
 
 if __name__ == "__main__":
     parser = argparse.ArgumentParser(description='Maximum-likelihood and Bayesian HMM estimation from p5ab hairpin data')

### 2015-05-18 examples/rnase-h-d10a/generate-figure.py

@@ -71,7 +71,7 @@ def run(nstates, nsamples):
 
 if __name__ == "__main__":
     parser = argparse.ArgumentParser(description='Maximum-likelihood and Bayesian HMM estimation from RNAse-H data')
-    parser.add_argument('--nstates', default=4, type=int, help='number of states')
+    parser.add_argument('--nstates', default=5, type=int, help='number of states')
     parser.add_argument('--nsamples', default=100, type=int, help='number of samples in Bayesian estimator')
     parser.add_argument('--verbose', dest='verbose', action='store_true', default=True, help='be loud and noisy')
     args = parser.parse_args()

### 2015-05-18 manuscripts/bayesian-hmm-method/bhmm.tex
Revised synthetic model, but convergence seems to be poor.

@@
 %\resizebox{\textwidth}{!}{\includegraphics{model-mlhmm-stateassignments}}
-\resizebox{\textwidth}{!}{\includegraphics{synthetic-three-state-model-mlhmm}}
+%\resizebox{\textwidth}{!}{\includegraphics{synthetic-three-state-model-mlhmm}}
+\resizebox{\textwidth}{!}{\includegraphics{synthetic-three-state-model-mlhmm-nstates3-nobs10000}}
 \caption{{\bf Synthetic force trajectory and inferred state assignments in MLHMM.}
@@
 \begin{table*}
-\caption{{\bf Estimated mean model parameters and confidence intervals for synthetic timeseries data}}
+\caption{{\bf Estimated mean model parameters and confidence intervals for synthetic timeseries data (new table)}}
 \label{table:synthetic-confidence-intervals}
@@
 \multicolumn{2}{l}{\bf Property} & \bf True Value & \bf 1 000 observations & \bf 10 000 observations & \bf 100 000 observations\\ \hline
-		Equilibrium probability 		& $\pi_{1}$ & $0.333$ & $0.338_{\:0.130}^{\:0.543}$  & $0.309_{\:0.251}^{\:0.374}$  & $0.330_{\:0.299}^{\:0.363}$  \\
-		& $\pi_{2}$ & $0.333$ & $0.351_{\:0.082}^{\:0.735}$  & $0.335_{\:0.246}^{\:0.432}$  & $0.329_{\:0.309}^{\:0.350}$  \\
-		& $\pi_{3}$ & $0.333$ & $0.312_{\:0.118}^{\:0.510}$  & $0.356_{\:0.299}^{\:0.411}$  & $0.341_{\:0.322}^{\:0.357}$  \\
-		\hline
-		Transition probability ($\Delta t = $1.0 ms) 		& $T_{11}$ & $0.989$ & $0.952_{\:0.926}^{\:0.972}$ & $0.947_{\:0.939}^{\:0.955}$ & $0.988_{\:0.987}^{\:0.989}$ \\
-		& $T_{12}$ & $0.010$ & $0.005_{\:0.001}^{\:0.015}$ & $0.001_{\:0.001}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
-		& $T_{13}$ & $0.001$ & $0.043_{\:0.024}^{\:0.067}$ & $0.051_{\:0.044}^{\:0.060}$ & $0.011_{\:0.010}^{\:0.012}$ \\
-		& $T_{21}$ & $0.010$ & $0.005_{\:0.001}^{\:0.016}$ & $0.001_{\:0.000}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
-		& $T_{22}$ & $0.940$ & $0.990_{\:0.975}^{\:0.998}$ & $0.988_{\:0.983}^{\:0.991}$ & $0.949_{\:0.946}^{\:0.951}$ \\
-		& $T_{23}$ & $0.050$ & $0.005_{\:0.001}^{\:0.015}$ & $0.011_{\:0.008}^{\:0.015}$ & $0.050_{\:0.048}^{\:0.053}$ \\
-		& $T_{31}$ & $0.001$ & $0.046_{\:0.028}^{\:0.072}$ & $0.044_{\:0.038}^{\:0.052}$ & $0.011_{\:0.009}^{\:0.012}$ \\
-		& $T_{32}$ & $0.050$ & $0.005_{\:0.001}^{\:0.016}$ & $0.010_{\:0.007}^{\:0.014}$ & $0.049_{\:0.046}^{\:0.051}$ \\
-		& $T_{33}$ & $0.949$ & $0.949_{\:0.922}^{\:0.970}$ & $0.945_{\:0.937}^{\:0.953}$ & $0.941_{\:0.938}^{\:0.943}$ \\
-		\hline
-		\hline
-		Transition rate (ms$^{-1}$) 		& $k_{12}$ & $0.010$ & $0.005_{\:0.001}^{\:0.015}$ & $0.001_{\:0.001}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
-		& $k_{13}$ & $0.001$ & $0.043_{\:0.024}^{\:0.067}$ & $0.051_{\:0.044}^{\:0.060}$ & $0.011_{\:0.010}^{\:0.012}$ \\
-		& $k_{21}$ & $0.010$ & $0.005_{\:0.001}^{\:0.016}$ & $0.001_{\:0.000}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
-		& $k_{23}$ & $0.050$ & $0.005_{\:0.001}^{\:0.015}$ & $0.011_{\:0.008}^{\:0.015}$ & $0.050_{\:0.048}^{\:0.053}$ \\
-		& $k_{31}$ & $0.001$ & $0.046_{\:0.028}^{\:0.072}$ & $0.044_{\:0.038}^{\:0.052}$ & $0.011_{\:0.009}^{\:0.012}$ \\
-		& $k_{32}$ & $0.050$ & $0.005_{\:0.001}^{\:0.016}$ & $0.010_{\:0.007}^{\:0.014}$ & $0.049_{\:0.046}^{\:0.051}$ \\
-		\hline
-		State mean lifetime (ms) 		& $t_{1}$ & $90.408$ & $21.797_{\:13.184}^{\:36.162}$ & $18.645_{\:15.899}^{\:21.524}$ & $83.405_{\:75.423}^{\:92.188}$ \\
-		& $t_{2}$ & $16.162$ & $148.113_{\:43.004}^{\:525.609}$ & $82.451_{\:59.320}^{\:115.043}$ & $18.954_{\:18.001}^{\:19.919}$ \\
-		& $t_{3}$ & $19.103$ & $20.138_{\:12.475}^{\:33.395}$ & $17.902_{\:15.417}^{\:20.637}$ & $16.395_{\:15.692}^{\:17.177}$ \\
-		\hline
-		Relaxation time (ms) 		& $\tau_{1}$ & $62.700$ & $82.300_{\:32.985}^{\:200.069}$ & $56.416_{\:44.005}^{\:70.295}$ & $58.230_{\:54.026}^{\:62.826}$ \\
-		& $\tau_{2}$ & $8.909$ & $10.445_{\:7.396}^{\:15.490}$ & $9.353_{\:8.387}^{\:10.399}$ & $8.970_{\:8.681}^{\:9.287}$ \\
-		\hline
-		\hline
-		State force mean (pN) 		& $\mu_{1}$ & $3.000$ & $4.677_{\:4.644}^{\:4.712}$ & $5.606_{\:5.600}^{\:5.613}$ & $3.002_{\:2.991}^{\:3.013}$ \\
-		& $\mu_{2}$ & $4.700$ & $3.069_{\:2.953}^{\:3.187}$ & $3.019_{\:2.986}^{\:3.056}$ & $5.602_{\:5.600}^{\:5.604}$ \\
-		& $\mu_{3}$ & $5.600$ & $5.602_{\:5.580}^{\:5.625}$ & $4.700_{\:4.690}^{\:4.710}$ & $4.703_{\:4.699}^{\:4.706}$ \\
-		\hline
-		State force std dev (pN) 		& $s_{1}$ & $1.000$ & $0.311_{\:0.285}^{\:0.337}$ & $0.199_{\:0.194}^{\:0.203}$ & $1.002_{\:0.995}^{\:1.010}$ \\
-		& $s_{2}$ & $0.300$ & $1.067_{\:0.992}^{\:1.147}$ & $1.012_{\:0.988}^{\:1.039}$ & $0.200_{\:0.199}^{\:0.202}$ \\
-		& $s_{3}$ & $0.200$ & $0.210_{\:0.195}^{\:0.229}$ & $0.297_{\:0.290}^{\:0.305}$ & $0.299_{\:0.296}^{\:0.301}$ \\
+		Equilibrium probability 		& $\pi_{1}$ & $0.385$ & $0.678_{\:0.367}^{\:0.889}$  & $0.533_{\:0.442}^{\:0.629}$  & $0.517_{\:0.486}^{\:0.549}$  \\
+		& $\pi_{2}$ & $0.094$ & $0.246_{\:0.069}^{\:0.546}$  & $0.351_{\:0.264}^{\:0.434}$  & $0.381_{\:0.353}^{\:0.410}$  \\
+		& $\pi_{3}$ & $0.521$ & $0.077_{\:0.029}^{\:0.156}$  & $0.116_{\:0.091}^{\:0.144}$  & $0.101_{\:0.093}^{\:0.110}$  \\
+		\hline
+		Transition probability ($\Delta t = $1.0 ms) 		& $T_{11}$ & $0.980$ & $0.991_{\:0.981}^{\:0.997}$ & $0.989_{\:0.986}^{\:0.992}$ & $0.989_{\:0.988}^{\:0.990}$ \\
+		& $T_{12}$ & $0.015$ & $0.001_{\:0.000}^{\:0.006}$ & $0.003_{\:0.002}^{\:0.005}$ & $0.004_{\:0.003}^{\:0.004}$ \\
+		& $T_{13}$ & $0.005$ & $0.009_{\:0.003}^{\:0.018}$ & $0.008_{\:0.006}^{\:0.010}$ & $0.007_{\:0.006}^{\:0.008}$ \\
+		& $T_{21}$ & $0.063$ & $0.003_{\:0.000}^{\:0.017}$ & $0.005_{\:0.003}^{\:0.008}$ & $0.005_{\:0.004}^{\:0.005}$ \\
+		& $T_{22}$ & $0.900$ & $0.967_{\:0.932}^{\:0.988}$ & $0.976_{\:0.970}^{\:0.982}$ & $0.978_{\:0.977}^{\:0.980}$ \\
+		& $T_{23}$ & $0.037$ & $0.030_{\:0.010}^{\:0.068}$ & $0.019_{\:0.014}^{\:0.025}$ & $0.017_{\:0.015}^{\:0.018}$ \\
+		& $T_{31}$ & $0.003$ & $0.079_{\:0.029}^{\:0.146}$ & $0.036_{\:0.026}^{\:0.047}$ & $0.036_{\:0.033}^{\:0.040}$ \\
+		& $T_{32}$ & $0.007$ & $0.093_{\:0.028}^{\:0.193}$ & $0.057_{\:0.043}^{\:0.072}$ & $0.063_{\:0.058}^{\:0.068}$ \\
+		& $T_{33}$ & $0.990$ & $0.828_{\:0.711}^{\:0.920}$ & $0.907_{\:0.887}^{\:0.925}$ & $0.901_{\:0.894}^{\:0.907}$ \\
+		\hline
+		\hline
+		Transition rate (ms$^{-1}$) 		& $k_{12}$ & $0.015$ & $0.001_{\:0.000}^{\:0.006}$ & $0.003_{\:0.002}^{\:0.005}$ & $0.004_{\:0.003}^{\:0.004}$ \\
+		& $k_{13}$ & $0.005$ & $0.009_{\:0.003}^{\:0.018}$ & $0.008_{\:0.006}^{\:0.010}$ & $0.007_{\:0.006}^{\:0.008}$ \\
+		& $k_{21}$ & $0.063$ & $0.003_{\:0.000}^{\:0.017}$ & $0.005_{\:0.003}^{\:0.008}$ & $0.005_{\:0.004}^{\:0.005}$ \\
+		& $k_{23}$ & $0.037$ & $0.030_{\:0.010}^{\:0.068}$ & $0.019_{\:0.014}^{\:0.025}$ & $0.017_{\:0.015}^{\:0.018}$ \\
+		& $k_{31}$ & $0.003$ & $0.079_{\:0.029}^{\:0.146}$ & $0.036_{\:0.026}^{\:0.047}$ & $0.036_{\:0.033}^{\:0.040}$ \\
+		& $k_{32}$ & $0.007$ & $0.093_{\:0.028}^{\:0.193}$ & $0.057_{\:0.043}^{\:0.072}$ & $0.063_{\:0.058}^{\:0.068}$ \\
+		\hline
+		State mean lifetime (ms) 		& $t_{1}$ & $49.498$ & $128.116_{\:55.334}^{\:309.568}$ & $91.902_{\:70.769}^{\:120.147}$ & $93.429_{\:86.014}^{\:101.962}$ \\
+		& $t_{2}$ & $9.491$ & $36.448_{\:15.042}^{\:92.972}$ & $42.545_{\:33.175}^{\:54.544}$ & $45.889_{\:42.634}^{\:49.415}$ \\
+		& $t_{3}$ & $99.499$ & $5.986_{\:3.065}^{\:13.200}$ & $10.396_{\:8.396}^{\:12.904}$ & $9.555_{\:8.969}^{\:10.219}$ \\
+		\hline
+		Relaxation time (ms) 		& $\tau_{1}$ & $56.704$ & $54.467_{\:26.737}^{\:109.929}$ & $51.201_{\:41.887}^{\:62.042}$ & $53.387_{\:50.301}^{\:56.513}$ \\
+		& $\tau_{2}$ & $8.377$ & $5.191_{\:2.656}^{\:10.780}$ & $8.855_{\:7.255}^{\:10.815}$ & $8.335_{\:7.857}^{\:8.866}$ \\
+		\hline
+		\hline
+		State force mean (pN) 		& $\mu_{1}$ & $3.000$ & $5.603_{\:5.588}^{\:5.619}$ & $5.600_{\:5.595}^{\:5.606}$ & $5.599_{\:5.597}^{\:5.600}$ \\
+		& $\mu_{2}$ & $4.700$ & $2.950_{\:2.818}^{\:3.085}$ & $3.003_{\:2.970}^{\:3.039}$ & $2.996_{\:2.986}^{\:3.007}$ \\
+		& $\mu_{3}$ & $5.600$ & $4.706_{\:4.598}^{\:4.822}$ & $4.712_{\:4.694}^{\:4.731}$ & $4.700_{\:4.694}^{\:4.706}$ \\
+		\hline
+		State force std dev (pN) 		& $s_{1}$ & $1.000$ & $0.198_{\:0.188}^{\:0.209}$ & $0.198_{\:0.195}^{\:0.202}$ & $0.200_{\:0.198}^{\:0.201}$ \\
+		& $s_{2}$ & $0.300$ & $1.011_{\:0.918}^{\:1.114}$ & $0.994_{\:0.970}^{\:1.018}$ & $0.998_{\:0.991}^{\:1.005}$ \\
+		& $s_{3}$ & $0.200$ & $0.417_{\:0.290}^{\:0.524}$ & $0.291_{\:0.277}^{\:0.305}$ & $0.298_{\:0.294}^{\:0.303}$ \\
 		\hline

### 2015-09-11 manuscripts/bayesian-hmm-method/bhmm-supplementary.tex

% PRL look and style (easy on the eyes)
\documentclass[endfloats,prl,twocolumn,superscriptaddress,nofootinbib,longbibliography]{revtex4-1}
% Two-column style (for submission/review/editing)
%\documentclass[aps,preprint,prl,superscriptaddress,showpacs]{revtex4}

\usepackage{palatino}
\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{boxedminipage}
\usepackage{verbatim}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}

\graphicspath{{figures/}}

%%% FROM NINA

%\usepackage{fullpage}
%\usepackage{amsmath}

\newcommand{\mus}{\mu_{1 \dots n}}
\newcommand{\sigmas}{\sigma_{1 \dots n}}
\newcommand{\obsv}[1]{A_{1 \dots #1}}
\newcommand{\states}[1]{s_{1 \dots #1}}

%%%%%%%%%%% OLD STUFF (UNUSED?) %%

% italicized boldface for math (e.g. vectors)
\newcommand{\bfv}[1]{{\mbox{\boldmath{$#1$}}}}
% non-italicized boldface for math (e.g. matrices)
\newcommand{\bfm}[1]{{\bf #1}}

%\newcommand{\bfm}[1]{{\mbox{\boldmath{$#1$}}}}
%\newcommand{\bfm}[1]{{\bf #1}}
\newcommand{\expect}[1]{\left \langle #1 \right \rangle}                % <.> for denoting expectations over realizations of an experiment or thermal averages

% vectors
%\newcommand{\x}{\bfv{x}}
%\newcommand{\y}{\bfv{y}}
%\newcommand{\f}{\bfv{f}}
%\newcommand{\z}{\bfv{z}}
%\newcommand{\p}{\bfv{p}}
%\newcommand{\q}{\bfv{q}}
%\newcommand{\T}{\mathrm{T}}

\newcommand{\op}[1]{\mathcal{#1}}
%\newcommand{\vektor}[1]{{\bf #1}}
%\renewcommand{\vec}[1]{\bfm{#1}} % italicized boldface
%\renewcommand{\vec}[1]{{\bf #1}}
\renewcommand{\vec}[1]{{\mbox{\boldmath{$#1$}}}}
\newcommand{\q}{\vec{q}} % coordinates
\newcommand{\p}{\vec{p}} % momenta
\newcommand{\z}{\vec{z}} % phase space point
\newcommand{\x}{\vec{x}} % general vector x
\newcommand{\M}{\bfm{M}} % diagonal mass matrix
\newcommand{\grad}{\nabla}
\newcommand{\timeavg}[1]{\overline{#1}}                    % time average over a trajectory

\newcommand{\bfc}{\bfm{c}}
\newcommand{\hatf}{\hat{f}}

\newcommand{\bTheta}{\bfm{\Theta}}
\newcommand{\btheta}{\bfm{\theta}}
\newcommand{\bhatf}{\bfm{\hat{f}}}
\newcommand{\Cov}[1] {\mathrm{cov}\left( #1 \right)}
\newcommand{\Ept}[1] {{\mathrm E}\left[ #1 \right]}
\newcommand{\Eptk}[2] {{\mathrm E}\left[ #2 \,|\, #1\right]}
\newcommand{\T}{\mathrm{T}}                                % T used in matrix transpose
\newcommand{\qed}{\hfill \mbox{\raggedright \rule{.07in}{.1in}}}  % QED mark
\newcommand{\var}[1]{\mathrm{var} \, #1}
\newcommand{\cov}[2]{\mathrm{cov}\left( {#1}, {#2} \right)}

%% DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%% TITLE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Supplementary Material for:\\
Bayesian hidden Markov model analysis of single-molecule force spectroscopy: Characterizing kinetics under measurement uncertainty}
%\title{Characterizing equilibrium kinetics in single-molecule force spectroscopy measurements with Bayesian hidden Markov models}
% JDC: Possible subtitle: Characterizing metastable states and transition rates under measurement uncertainty

\author{John D. Chodera}
 \thanks{Corresponding author}
% \email{jchodera@berkeley.edu}
 \affiliation{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720, USA}

\author{Phillip Elms}
% \email{elms@berkeley.edu}
 \affiliation{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720, USA}
 \affiliation{Biophysics Graduate Group, University of California, Berkeley, CA 94720, USA}
 \affiliation{Jason L.~Choy Laboratory of Single Molecule Biophysics, Institute for Quantitative Biosciences, University of California, Berkeley, CA 94720, USA}

\author{Frank No\'{e}}
% \email{frank.noe@fu-berlin.de}
 \affiliation{DFG Research Center Matheon, FU Berlin, Arnimallee 6, 14195 Berlin, Germany}

\author{Bettina Keller}
% \email{bettina.keller@fu-berlin.de}
 \affiliation{DFG Research Center Matheon, FU Berlin, Arnimallee 6, 14195 Berlin, Germany}

\author{Christian M. Kaiser}
% \email{cmkaiser@berkeley.edu}
 \affiliation{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720, USA}
 \affiliation{Department of Physics, University of California, Berkeley, CA 94720, USA}

\author{Aaron Ewall-Wice}
% \email{olivercromwell@uchicago.edu}
 \affiliation{University of Chicago, IL 60637, USA}

\author{Susan Marqusee}
% \email{marqusee@berkeley.edu}
 \affiliation{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720, USA}
 \affiliation{Department of Molecular \& Cell Biology, University of California, Berkeley, CA 94720, USA}
 \affiliation{Jason L.~Choy Laboratory of Single Molecule Biophysics, Institute for Quantitative Biosciences, University of California, Berkeley, CA 94720, USA}

\author{Carlos Bustamante}
% \email{carlos@alice.berkeley.edu}
 \affiliation{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720, USA}
 \affiliation{Department of Molecular \& Cell Biology, University of California, Berkeley, CA 94720, USA}
 \affiliation{Jason L.~Choy Laboratory of Single Molecule Biophysics, Institute for Quantitative Biosciences, University of California, Berkeley, CA 94720, USA}
 \affiliation{Department of Physics, University of California, Berkeley, CA 94720, USA}
 \affiliation{Department of Chemistry, University of California, Berkeley, CA 94720, USA}
 \affiliation{Howard Hughes Medical Institute, University of California, Berkeley, CA 94720, USA}

\author{Nina Singhal Hinrichs}
% \email{nshinrichs@uchicago.edu}
 \affiliation{Departments of Statistics and Computer Science, University of Chicago, IL 60637, USA}

\date{\today}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT/pacs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
\end{abstract}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MAKE FONT SMALLER TO MATCH JOURNAL STYLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\small

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% vALIDATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Validation}

\begin{figure}[tbp]
\noindent
\resizebox{\columnwidth}{!}{\includegraphics{confidence-intervals-random}}
\caption{{\bf Validation of confidence intervals using randomly-generated synthetic test data.}
Observed confidence intervals (points) are plotted as a function of the desired confidence intervals for equilibrium probabilities ($\pi_i$), transition probabilities ($T_{ij}$), state means ($\mu_i$), and state standard deviations ($\sigma_i$).
The black diagonal line indicates perfect agreement between expected and observed confidence intervals, while observed confidence intervals above the diagonal indicate overestimates of the uncertainty, and below the diagonal indicate underestimates.
Because only 50 random models were evaluated, error bars denote a 95\% confidence interval in the estimated observed confidence intervals.
\label{figure:validation-confidence-intervals}}
\end{figure}

The confidence interval comparison figure referred to in the main body text appears as Supplementary Figure~\ref{figure:validation-confidence-intervals}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PROOF FOR STATE HISTORY SAMPLING
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Proof of state history sampling scheme}
\label{section:proof-state-history-sampling}

In the reverse recursion for updating the hidden state sequence in sampling from the BHMM posterior by Gibbs sampling, we \emph{sample} a state sequence by sampling each hidden state from the conditional distribution $s_t \sim P(s_t \mid s_{t+1}, \ldots, s_L)$ starting from $t = L$ and proceeding down to $t = 0$, where the conditional distribution is given by (Eq.~35 in the main text),
\begin{eqnarray}
\lefteqn{P(s_t = i \mid s_{t+1}, \ldots, s_L)} \\
&\propto& \begin{cases}
\alpha_{t i} / \sum_{j=1}^M \alpha_{t j} & t = L \\
\alpha_{t i} T_{i s_{t+1}} / \sum_{j=1}^M \alpha_{t j} T_{j s_{t+1}} & t = (L-1),\ldots,0
\end{cases} \nonumber
\end{eqnarray} 
It is straightforward to show the result of these sampling steps reconstitutes the probability distribution $P(\bfm{S} | \bfm{T}, \bfm{E}, \bfm{O})$.
We note that the composition of these conditional distributions defines a joint distribution,
\begin{eqnarray}
\lefteqn{P(s_0 | s_1,\ldots) P(s_1 | s_2, \ldots) \cdots P(s_L) =} \nonumber \\
&=& \left[\frac{\alpha_{0 s_0} T_{s_0 s_1}}{\sum_{j=1}^M \alpha_{0 j} T_{j s_1}} \right] \left[\frac{\alpha_{1 s_1} T_{s_1 s_2}}{\sum_{j=1}^M \alpha_{1 j} T_{j s_2}} \right] \cdots \nonumber \\
&=& \left[\frac{\rho_{s_0} \varphi(o_0 | \bfm{e}_{s_0}) T_{s_0 s_1}}{\sum_{j=1}^M \alpha_{0 j} T_{j s_1}} \right] \left[\frac{\varphi(o_1 | \bfm{e}_{s_1}) \left(\sum_{j=1}^M \alpha_{0j} T_{j s_1} \right) T_{s_1 s_2}}{\sum_{j=1}^M \alpha_{1 j} T_{j s_2}} \right] \cdots \nonumber \\
&=& \frac{\rho_{s_0} \varphi(o_0 | \bfm{e}_{s_0}) T_{s_0 s_1} \cdots \varphi(o_L \mid \bfm{e}_{s_L})}{\sum_{j=1}^M \alpha_{L j}} \nonumber \\
&=& \frac{\rho_{s_0} \varphi(o_0 | \bfm{e}_{s_0}) T_{s_0 s_1} \cdots \varphi(o_L \mid \bfm{e}_{s_L})}{P(\bfm{O} | \bfm{E}) }
\end{eqnarray}
where we have used the fact that $P(\bfm{O} \mid \bfm{E}) = \sum_{j=1}^M \alpha_{tj}$,
\begin{eqnarray}
\sum_{j=1}^M \alpha_{Lj} &=& \sum_{s_L=1}^M \varphi(o_L | \bfm{e}_{s_L}) \sum_{s_{L-1}=1}^M \alpha_{(t-1) s_{L-1}} T_{s_{L-1} s_L} \nonumber \\
&=& \sum_{s_{L-1}=1}^M \alpha_{(t-1) s_{L-1}}\sum_{s_L=1}^M \varphi(o_L | \bfm{e}_{s_L})  T_{s_{L-1} s_L} \nonumber \\
%&=& \sum_{s_{L-1}=1}^M \left[ \varphi(o_{L-2} \mid \bfm{e}_{s_{L-2}}) \sum_{s_{L-2}=1}^M \alpha_{(L-2) s_{L-2}} T_{s_{L-2} s_{L-1}} \right] \sum_{s_L=1}^M \varphi(o_L | \bfm{e}_{s_L})  T_{s_{L-1} s_L} \nonumber \\
&=& \sum_{s_0} \rho_{s_0} \varphi(o_0 \mid \bfm{e}_{s_0}) T_{s_{0} s_{1}} \cdots \varphi(o_L \mid \bfm{e}_L) \nonumber \\
&=& \sum_{\bfm{S}} \rho_{s^{}_{0}} \varphi(o^{}_{0} \mid \bfm{e}_{s^{}_{0}}) \prod_{t=1}^{L^{}} T_{s^{}_{(t-1)} s^{}_{t}} \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \nonumber \\
&=& P(\bfm{O} \mid \bfm{\Theta}) .
\end{eqnarray}
where the last step is seen by recursive substitution of the $\alpha_{tj}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THRESHOLD MODEL
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Comparison with threshold model}
\label{section:model-system-segmentation}

To demonstrate that simply dividing the observable into non-overlapping regions with thresholds to define states leads to consistently biased model properties, we tested a simple segmentation scheme on the same three-state synthetic example considered in the main text.
Here, a mixture of Gaussians was first fit to the pooled observations (here, forces) according to the scheme described in \emph{Methods} to identify the peaks corresponding to each state.
%Observations were then allocated into discrete states using the crossing points between Gaussian mixture component density functions as dividing points.
A simple partition (or threshold) was assigned to the crossing point between each Gaussian probability density.
This threshold was used to determine changes between states and therefore transitions and lifetimes.
A maximum-likelihood transition matrix was then fit to the observed transition counts between states.

The result of this segmentation procedure is shown for trajectories of $10^3$ to $10^6$ observations in Figure~\ref{figure:model-segmentation-stateassignments}.
Even by eye, it is clear that there is a discrepancy between changes in state assignment color and changes in kinetic behavior.
Small conformational fluctuations that cause the force to transiently cross a threshold, but not commit to the new conformational state, result in the obsevation of spurious extra transitions.
These extra transitions will lead to an underestimate of the state self-transition probabilities and lifetimes, and convergence to incorrect model parameters.
This is confirmed by examination of the estimated model properties as a function of trajectory length in Table~\ref{table:synthetic-segmentation-model}.
Compare with Table I from the main text, where by $10^5$ observations, the estimated mean BHMM model parameters have attained relatively small error compared to the error in the segmentation model even for $10^6$ observations.

\begin{table}[tbp]
\eject
\caption{{\bf Estimated segmentation model parameters for synthetic timeseries data}}
\label{table:synthetic-segmentation-model}
\begin{tabular}{|c|c|cccc|}
\hline
property & true value & \multicolumn{4}{c|}{observation length} \\
 & & $10^3$ & $10^4$ & $10^5$ & $10^6$ \\ \hline
$\pi_{1}$ & $0.308$ & $0.214$ & $0.276$ & $0.278$ & $0.267$ \\
$\pi_{2}$ & $0.113$ & $0.155$ & $0.184$ & $0.178$ & $0.174$ \\
$\pi_{3}$ & $0.579$ & $0.632$ & $0.540$ & $0.544$ & $0.559$ \\
\hline
$T_{11}$ & $0.980$ & $0.859$ & $0.822$ & $0.828$ & $0.833$ \\
$T_{12}$ & $0.019$ & $0.125$ & $0.165$ & $0.158$ & $0.154$ \\
$T_{13}$ & $0.001$ & $0.016$ & $0.013$ & $0.014$ & $0.014$ \\
$T_{21}$ & $0.053$ & $0.173$ & $0.247$ & $0.246$ & $0.236$ \\
$T_{22}$ & $0.900$ & $0.523$ & $0.579$ & $0.564$ & $0.577$ \\
$T_{23}$ & $0.050$ & $0.304$ & $0.174$ & $0.190$ & $0.187$ \\
$T_{31}$ & $0.001$ & $0.005$ & $0.007$ & $0.007$ & $0.006$ \\
$T_{32}$ & $0.009$ & $0.074$ & $0.059$ & $0.062$ & $0.058$ \\
$T_{33}$ & $0.990$ & $0.920$ & $0.934$ & $0.930$ & $0.935$ \\
\hline
$\mu_{1}$ & $3.000$ & $2.707$ & $2.716$ & $2.746$ & $2.737$ \\
$\mu_{2}$ & $4.700$ & $4.841$ & $4.707$ & $4.714$ & $4.710$ \\
$\mu_{3}$ & $5.600$ & $5.636$ & $5.621$ & $5.619$ & $5.618$ \\
\hline
$\sigma_{1}$ & $1.000$ & $0.842$ & $0.804$ & $0.806$ & $0.806$ \\
$\sigma_{2}$ & $0.300$ & $0.330$ & $0.332$ & $0.339$ & $0.334$ \\
$\sigma_{3}$ & $0.200$ & $0.201$ & $0.187$ & $0.182$ & $0.183$ \\
\hline
\end{tabular}
\end{table}

\begin{figure*}[p]
\noindent
\resizebox{\textwidth}{!}{\includegraphics{model-segmentation-stateassignments-1000}}
\resizebox{\textwidth}{!}{\includegraphics{model-segmentation-stateassignments-10000}}
\resizebox{\textwidth}{!}{\includegraphics{model-segmentation-stateassignments-100000}}
\resizebox{\textwidth}{!}{\includegraphics{model-segmentation-stateassignments-1000000}}
\caption{{\bf Synthetic force trajectory and segmentation state assignments.}
Observed samples are colored by their hidden state assignments from a model in which the observable is segmented into regions that correspond to each state, for trajectories of length $10^3$ (\emph{top}), $10^4$ (\emph{upper middle}), $10^5$ (\emph{lower middle}), and $10^6$ (\emph{bottom}) observations.
The segmentation was derived by first fitting the set of all observed forces to a mixture of Gaussians and then dividing the observations at the crossing points of the Gaussian mixture component probability density functions.
Dark horizontal lines terminating in triangles to the right denote state means, while lightly colored bands indicate one standard deviation on either side of the state mean.
The gray histogram on the right side shows the total observed probability of samples, while the colored peaks show the contribution from each state, assuming a Gaussian fit to the mean and standard deviation.
\label{figure:model-segmentation-stateassignments}}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CORRELATION FUNCTIONS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Choice of observation interval for hairpin}
\label{section:choice-of-markov-time}

To determine that a 1 ms time was an appropriately Markovian time for the observation interval $\Delta t$ used in HMM analysis, the autocorrelation function of the experimental force trace was computed at the full 50 kHz time resolution for a trap position where only a single conformational state appeared to dominate (Figure~\ref{figure:p5ab-autocorrelation}).
An initial nonexponential or multiexponential (with unclear number of components) decay is seen within the first 0.5 ms to approximately one tenth of the initial value, with slow decay following.
We therefore chose a 1 ms time resolution as appropriate for describing the slow conformational dynamics between states.

\begin{figure}[p]
\noindent
\resizebox{\columnwidth}{!}{\includegraphics{fiber3-trace14-autocorrelation}}
\caption{{\bf Normalized fluctuation autocorrelation function for p5ab hairpin under conditions stabilizing a single conformation.}
\emph{Top:} The force histogram for a trap position (different from that in main text Fig.~4) showing that a single conformational state is dominant at this trap configuration.
\emph{Bottom:} The normalized integrated autocorrelation function for the force trace, showing rapid decay of non-Markovian contributions within 1 ms.
\label{figure:p5ab-autocorrelation}}
\end{figure}

\widetext
\eject

\end{document}
